### 2022-11-09 manuscript/manuscript.rmd
Adding new analyses

---
title: Analyzing COVID-19 Mortality at the County Level in the U.S. Using Generalized Additive
  Models
author: "Christian Testa"
date: '`r Sys.Date()`'
output: pdf_document
abstract: this is an abstract
---

```{r setup, echo = FALSE}
library(here)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = '100%'
)
```

# Introduction

As we enter the third winter with novel coronavirus disease (COVID-19) in the
United States (U.S.), evidence reflecting intense disparities in COVID-19 mortality rates 
among socially advantaged and disadvantaged populations persists.  Eliminating
inequities in health outcomes is a major policy goal of the Biden
administration, representing an revitalized focus on health justice
and underlining the importance of adequate data reporting systems that measure
the prevalent health inequities.  To this end, generalized additive models
(GAMs) provide a flexible regression framework which can illustrate the evolving
roles and relationships social conditions have with trends in COVID-19
mortality.

# Background

Having passed over 1 million COVID-19 deaths in the United States earlier this
year, and facing uncertain prospects for the third COVID-19 winter looming even
as new iterations on the COVID-19 vaccines become available, it remains critical
that inequities in COVID-19 outcomes are documented and analyzed to reckon with
the unjust and unfair burden of preventable illness.  Even though the first
vaccines were granted emergency use authorization by the U.S. Food and Drug
Administration in 2020, with the first shots going in arms in December 2020,
COVID-19 is still continuing to cause hundreds of deaths a day in the U.S. in
the late summer and early fall of 2022.  The new bivalent vaccines released at
the end of August 2022 contain messenger RNA components of the original strain
as well as the BA.4 and BA.5 lineages in an effort to make the nation's immunity
more up-to-date and robust against the myriad of phylogenetic directions the
COVID-19 virus is exploring.  Despite the updated bivalent boosters representing
a significant step forward in prevention strategy, less than 4% of eligible
Americans had taken the booster in the first month after it became available. As
such, and with an enduring history of inequities in health care availability, it
is not clear that all communities will be equally able to take advantage of the
new vaccines and inequities in COVID-19 illness and mortality may persist
despite the technological innovations in vaccine therapy.

```{r figure 1, fig.cap = "Monthly estimates of COVID-19 mortality per 100,000 residents by state are shown.  States are in distinct colors to make each state's trajectory visible. Reference the online version TODO: XXX"}
knitr::include_graphics(here("analysis/01_state_trajectories/state_covid19_mortality_trajectories.png"))
```

# Methods

## Data Sources

The following variables were retrieved at the county level:

  * Counts of deaths by month with underlying cause of death as
  COVID-19 (ICD-10 code U07.1) from CDC WONDER, Provisional Multiple Cause of Death
  Data
  * Population Size Estimates for 2020 from the U.S. Census (Redistricting 
  File Public-Law 94-171 Dataset). *Note:*  While the complete data for the 2020
  U.S. Census have not yet been released, the summary tables including
  population counts have been released as mandated by Congress for the purposes
  of redistricting and updating the boundaries of the 435 voting Congressional
  Districts, which are what is made us of here as population size estimates.
  * Percent Black, percent Hispanic, percent Asian, percent American Indian or
  Alaska Native, Percent below the federal poverty line, median household
  income, median age, percent 65 years or older, and percent 75 years or older
  from the 2014-2019 5-year American Community Survey. 
  *Note:* while more recent American Community Survey 5-year estimates are
  available, survey data collection in 2020 was severely affected by the
  pandemic with the 2020 data release denoted as "experimental".

## Generalized Additive Models

# Results

We used GAMs to describe the following:

  * the changing associations between individual area-based measures
  and COVID-19 mortality over time;
  * the changing associations between bivariate area-based measures
  and COVID-19 mortality over time; 
  * associations with area-based measures and COVID-19 mortality after
  controlling for spatio-temporal autocorrelation.

Additionally, as sensitivity analyses, random forest models were fit to compare
with the coefficient estimates from the GAMs.

# Discussion

## Limitations

  * Missing data 
  * Ecological fallacy 
  * Age effects
  * Fixed population denominators -> real-time denominator estimates
  * Vaccination and mobility effects

# Conclusions

# References

<!-- 1 million covid deaths: 
https://coronavirus.jhu.edu/from-our-experts/u-s-officially-surpasses-1-million-covid-19-deaths -->

<!-- nearly 225,000 Americans have died since the start of 2022
https://abcnews.go.com/Health/225000-americans-died-covid-19-start-2022-viral/story?id=90339579 -->

<!-- 
vaccine timeline
https://www.mayoclinic.org/coronavirus-covid-19/history-disease-outbreaks-vaccine-timeline/covid-19
-->

<!--
less than 4% of americans had gotten the booster 1 month into availability
https://www.nbcnews.com/health/health-news/updated-covid-booster-shots-doses-administered-cdc-rcna48960
-->

<!-- 2020 ACS data are experimental 
https://www.census.gov/newsroom/press-releases/2021/experimental-2020-acs-1-year-data.html -->

<!--
Spatial-temporal generalized additive model for modeling COVID-19 mortality risk in Toronto, Canada
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8257405/ -->

<!-- 
Using GAM functions and Markov-Switching models in an evaluation framework to assess countries’ performance in controlling the COVID-19 pandemic
https://bmcpublichealth.biomedcentral.com/articles/10.1186/s12889-021-11891-6 -->

<!--
Statistical modelling of COVID-19 data: Putting generalized additive models to work
https://journals.sagepub.com/doi/full/10.1177/1471082X221124628
-->

<!--
Inference following multiple imputation for generalized additive models: an investigation of the median p-value rule with applications to the Pulmonary Hypertension Association Registry and Colorado COVID-19 hospitalization data
https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-022-01613-w
-->

<!--
Generalized Additive Models to Capture the Death Rates in Canada COVID-19
https://link.springer.com/chapter/10.1007/978-3-030-85053-1_7 -->

<!--
Contribution of Two-Dose Vaccination Toward the Reduction of COVID-19 Cases, ICU Hospitalizations and Deaths in Chile Assessed Through Explanatory Generalized Additive Models for Location, Scale, and Shape
https://www.frontiersin.org/articles/10.3389/fpubh.2022.815036/full -->

<!-- 
Combining graph neural networks and spatio-temporal disease models to improve the prediction of weekly COVID-19 cases in Germany
https://www.nature.com/articles/s41598-022-07757-5 -->

<!-- 
Association between mobility patterns and COVID-19 transmission in the USA: a mathematical modelling study
https://www.sciencedirect.com/science/article/pii/S1473309920305533 -->

<!-- 
Examining spatiotemporal evolution of racial/ethnic disparities in human mobility and COVID-19 health outcomes: Evidence from the contiguous United States
https://www.sciencedirect.com/science/article/pii/S2210670721007721 -->

<!--
Social, economic, and environmental factors influencing the basic reproduction number of COVID-19 across countries
https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0252373 -->

<!-- 
https://asbates.rbind.io/2019/05/03/gams-for-time-series/
-->

<!--
https://environmentalcomputing.net/statistics/gams/
-->

<!--
https://petolau.github.io/Analyzing-double-seasonal-time-series-with-GAM-in-R/
-->

# Supplementary Materials

### 2022-11-09 R/create_analytic_dataset.R

@@ -1,0 +1,77 @@
+
+#' Create Analytic Dataset
+#'
+#' @examples
+#' library(ggdist)
+#' df <- create_analytic_dataset()
+#'
+#'
+create_analytic_dataset <- function() {
+
+  # load deaths and 2020 population size data
+  deaths <- load_county_covid19_deaths()
+  popsizes <- get_county_popsizes()
+
+  # create month column
+  deaths %<>% mutate(
+    month = paste0(lubridate::year(date), "-", lubridate::month(date))
+  )
+  # aggregate by month
+  deaths %<>% group_by(geoid, county, state, month) %>%
+    summarize(
+      deaths = sum(deaths_avg)
+    )
+
+  # create date version of month column
+  deaths %<>% mutate(date = lubridate::ym(month)) %>% select(-month)
+
+  # drop October 2022 (incomplete data)
+  deaths %<>% filter(date < lubridate::ymd("2022-10-01"))
+
+  # join in population size estimates
+  deaths %<>% left_join(
+    popsizes %>% select(GEOID, value) %>% rename(popsize = value),
+    by = c('geoid' = 'GEOID')
+  )
+
+  # create deaths per 100,000 person-years
+  deaths %<>% mutate(crude_rate_per_100k_py = deaths / popsize * 1e5 * 12)
+
+  # add divisions and regions to the dataset
+  deaths %<>% mutate(
+    division = c(setNames(as.character(state.division), state.name), 'District of Columbia' = 'South Atlantic')[state],
+    region = c(setNames(as.character(state.region), state.name), 'District of Columbia' = 'South')[state]
+  )
+
+  # incorporate state divisions
+  deaths %<>% mutate(division = factor(
+    division,
+    levels = c(
+      'Pacific',
+      'West North Central',
+      'New England',
+      'Mountain',
+      'East North Central',
+      'Middle Atlantic',
+      'West South Central',
+      'East South Central',
+      'South Atlantic'
+    )),
+    region = factor(region,
+                    levels = c('West', 'Northeast', 'South', 'North Central')))
+
+  # deaths have to be integer for use in the poisson/negative binomial type models
+  deaths$deaths %<>% round() %>% as.integer()
+
+  # load cached absms, generated by get_absms_from_acs
+  absms <- load_cached_absms()
+
+  # join in area based social metrics
+  deaths %<>% left_join(absms, by = c('geoid' = 'GEOID'))
+
+  # get political lean data
[…]

### 2022-11-09 R/get_absms_from_acs.R

@@ -1,0 +1,154 @@
+
+#' Get Area Based Social Metrics from ACS
+#'
+#' @examples
+#' absms <- get_absms_from_acs()
+#' saveRDS(absms, file.path(system.file('', package = 'covid.gradient.estimation'), 'cached_absms.rds'))
+get_absms_from_acs <- function() {
+
+  absms_dictionary <- tibble::tribble(
+    ~var, ~varname, ~description,
+    # total population
+    "B01001_001",  "total_popsize", "total population estimate",
+
+    # age breakdown
+    "B01002_001",   "median_age",   "median age for total population",
+    "B06001_001",   "total_for_age", "total population for age distribution",
+    "B06001_011",   "65_to_74",     "population aged 65 to 74",
+    "B06001_012",   "75_and_up",    "population aged 75 and up",
+
+    # college education
+    "B06009_001",   "education_popsize", "total population estimate for education tables",
+    "B06009_002",   "less_than_hs",   "less than a high school or GED equivalent",
+    "B06009_005",   "college_degree", "bachelors degree",
+    "B06009_006",   "graduate_degree", "graduate degree",
+
+    # racial composition
+    'B01003_001',  "race_ethnicity_total", "race_ethnicity_total",
+
+    # ICEraceinc
+    "B19001_001",  'hhinc_total',   "total population for household income estimates",
+    "B19001A_002", 'hhinc_w_1',     "white n.h. pop with household income <$10k",
+    "B19001A_003", 'hhinc_w_2',     "white n.h. pop with household income $10k-14 999k",
+    "B19001A_004", 'hhinc_w_3',     "white n.h. pop with household income $15k-19 999k",
+    "B19001A_005", 'hhinc_w_4',     "white n.h. pop with household income $20k-24 999k",
+    "B19001A_014", 'hhinc_w_5',     "white n.h. pop with household income $100 000 to $124 999",
+    "B19001A_015", 'hhinc_w_6',     "white n.h. pop with household income $125k-149 999k",
+    "B19001A_016", 'hhinc_w_7',     "white n.h. pop with household income $150k-199 999k",
+    "B19001A_017", 'hhinc_w_8',     "white n.h. pop with household income $196k+",
+    "B19001_002",  'hhinc_total_1', "total pop with household income <$10k",
+    "B19001_003",  'hhinc_total_2', "total pop with household income $10k-14 999k",
+    "B19001_004",  'hhinc_total_3', "total pop with household income $15k-19 999k",
[…]

### 2022-11-09 R/get_popsizes.R

@@ -7,3 +7,14 @@ get_state_popsizes <- function() {
     year = 2020
   )
 }
+
+
+#' Get County Population Data
+get_county_popsizes <- function() {
+  tidycensus::get_decennial(
+    geography = 'county',
+    variables = "P1_001N",
+    year = 2020
+  )
+
+}

### 2022-11-09 R/load_covid19_deaths.R

@@ -18,3 +18,27 @@ load_state_covid19_deaths <- function(clean = TRUE) {
 
   return(df)
 }
+
+
+#' Load County COVID-19 Deaths
+load_county_covid19_deaths <- function(clean = TRUE) {
+
+  # load data
+  dfs <- purrr::map(
+    2020:2022, ~ readr::read_csv(system.file(
+      paste0(
+        "nytimes/",
+        "us-counties-", ., ".csv"
+      ),
+      package = 'covid.gradient.estimation'
+    ))
+  )
+
+  # clean data
+  if (clean) {
+    dfs %<>% bind_rows()
+    dfs$geoid %<>% stringr::str_extract("[0-9]{5}")
+  }
+
+  return(dfs)
+}

### 2022-11-09 R/load_political_lean.R

@@ -1,0 +1,32 @@
+
+#' Load MIT Election Lab County Level Presidential Returns
+load_mit_election_data <- function() {
+  df <- readr::read_csv(
+    system.file("mit_election_lab/countypres_2000-2020.csv", package = "covid.gradient.estimation")
+  )
+
+  df %<>% filter(year == 2020) # filter for just 2020 results
+
+  # combine the election day, early mail-in, and provisional votes
+  df %<>% select(county_fips, party, candidatevotes, totalvotes) %>%
+    group_by(county_fips, party) %>% summarize(across(everything(), ~sum(.)))
+
+  df %<>% ungroup()
+
+  # pivot wider so we can calculate political lean
+  df %<>% tidyr::pivot_wider(
+      id_cols = c(county_fips, totalvotes),
+      names_from = party,
+      values_from = candidatevotes
+    )
+
+  # calculate political lean
+  df %<>% mutate(
+    political_lean = (REPUBLICAN - DEMOCRAT) / totalvotes
+  )
+
+  # save only necessary info
+  df %<>% select(county_fips, political_lean)
+
+  return(df)
+}

### 2022-11-09 analysis/02_plot_county_trends/plot_county_trends.R

@@ -1,0 +1,86 @@
+
+# investigating suppression
+
+# setup
+devtools::load_all(".")
+library(here)
+
+# get data
+df <- create_analytic_dataset()
+
+# create plot
+df %>%
+  filter(!is.na(region)) %>%
+  ggplot(aes(x = date, y = crude_rate_per_100k_py, group = geoid)) +
+  geom_point(alpha = .05) +
+  stat_lineribbon(
+    mapping = aes(
+      group = region,
+      fill = region,
+      color = region,
+      fill_ramp = stat(level)
+    ),
+    alpha = .75,
+    color = 'grey20'
+  ) +
+  facet_wrap( ~ region) +
+  scale_y_continuous(
+    trans = scales::pseudo_log_trans(sigma = 100),
+    breaks = c(0, 100, 1000, 10000, 30000),
+    limits = c(0, NA),
+    labels = scales::comma_format()
+  ) +
+  theme_bw() +
+  guides(
+    fill = guide_none(),
+    color = guide_none(),
+    fill_ramp = guide_legend(title =
+                               'Quantile Range')
+  ) +
+  ylab("COVID-19 Mortality Rate per 100,000 Person-Years") +
+  ggtitle(
+    "County Observations of COVID-19 Mortality Rates by Region",
+    "Median and Quantile Ranges Across Counties"
+  ) +
+  theme(legend.position = 'bottom')
+
+ggsave(here("analysis/02_plot_county_trends/county_trends.png"), height = 6, width = 8)
+
+
+df %>%
+  filter(!is.na(division)) %>%
+  ggplot(aes(x = date, y = crude_rate_per_100k_py, group = geoid)) +
+  geom_point(alpha = .05) +
+  stat_lineribbon(
+    mapping = aes(
+      group = division,
+      fill = division,
+      color = division,
+      fill_ramp = stat(level)
+    ),
+    alpha = .75,
+    color = 'grey20'
+  ) +
+  facet_wrap( ~ division) +
+  scale_y_continuous(
+    trans = scales::pseudo_log_trans(sigma = 100),
+    breaks = c(0, 250, 1000, 3500, 10000, 30000),
+    limits = c(0, NA),
+    labels = scales::comma_format()
+  ) +
+  theme_bw() +
+  guides(
+    fill = guide_none(),
+    color = guide_none(),
+    fill_ramp = guide_legend(title =
+                               'Quantile Range')
+  ) +
[…]

### 2022-11-09 analysis/03_county_distributions/plot_county_distributions.R

@@ -1,0 +1,141 @@
+library(here)
+devtools::load_all(".")
+library(ggcorrplot)
+
+# load our area based measures we'll be considering
+absms <- load_cached_absms()
+
+# load our analytic dataset too
+df <- create_analytic_dataset()
+
+# pull out the region information
+absms %<>% left_join(df %>% select(geoid, region) %>% unique(), by = c('GEOID' = 'geoid'))
+
+# first I think we want a panel of univariate distributions (histograms)
+# showing how the counties fall across the measures considered
+
+# broadly we can categorize these into:
+#
+# purely demographic
+#
+# purely economic
+#
+# and socioeconomic
+#
+
+library(ggridges)
+
+absms %>% filter(!is.na(region)) %>% mutate(region = as.character(region)) %>%
+  ggplot(aes(x = prop_black, y = region)) +
+  geom_density_ridges(stat = 'binline', scale = .80) +
+  # scale_x_continuous(labels = scales::percent_format()) +
+  scale_x_continuous(trans = scales::pseudo_log_trans(sigma = .05)) + # breaks = c(0, 10, 100, 500, 10000), labels = scales::comma_format()) +
+  scale_y_discrete(expand = expansion(add = c(.3, 1.1))) +
+  theme_bw() +
+  labs(x = "Percent Black")
+
+absms %>% filter(!is.na(region)) %>% mutate(region = as.character(region)) %>%
+  ggplot(aes(x = prop_hispanic, y = region)) +
+  geom_density_ridges(stat = 'binline', scale = .90) +
+  scale_x_continuous(labels = scales::percent_format()) +
+  theme_bw() +
+  labs(x = "Percent Hispanic")
+
+absms %>% filter(!is.na(region)) %>% mutate(region = as.character(region)) %>%
+  ggplot(aes(x = prop_white_nh, y = region)) +
+  geom_density_ridges(stat = 'binline', scale = .90) +
+  scale_x_continuous(labels = scales::percent_format()) +
+  theme_bw() +
+  labs(x = "Percent Non-Hispanic White")
+
+absms %>% filter(!is.na(region)) %>% mutate(region = as.character(region)) %>%
+  ggplot(aes(x = prop_aian, y = region)) +
+  geom_density_ridges(stat = 'binline', scale = .90) +
+  scale_x_continuous(labels = scales::percent_format()) +
+  theme_bw() +
[…]

### 2022-11-09 analysis/04_one_variable_over_time/one_variable_over_time.R

@@ -1,0 +1,175 @@
+library(here)
+library(mgcv)
+library(mgcViz)
+devtools::load_all(".")
+
+# load our analytic dataset
+df <- create_analytic_dataset()
+
+# create a numeric version of date
+df$date_int <- as.integer(lubridate::year(df$date)) + (as.integer(lubridate::month(df$date))-1)/12
+
+# poverty
+model <- gam(
+  formula = deaths ~ te(date_int, prop_in_poverty, d = c(1,1)),
+  offset = log(popsize/1e5/12),
+  data = df,
+  family = nb()
+)
+
+png(
+  filename = here("analysis/04_one_variable_over_time/poverty.png")
+)
+mgcv::vis.gam(model, n.grid = 50, theta = 35, phi = 32, zlab = "", contour.col = 'topo', type = 'response',
+              ticktype = "detailed", color = "topo", main = "Interaction by Poverty and Time",
+              xlab = "\n\nDate", ylab = '\n\nProportion in Poverty', nticks = 4)
+dev.off()
+
+
+# median income
+model <- gam(
+  formula = deaths ~ te(date_int, median_income, d = c(1,1)),
+  offset = log(popsize/1e5/12),
+  data = df,
+  family = nb()
+)
+png(filename = here("analysis/04_one_variable_over_time/income_and_time.png"))
+mgcv::vis.gam(model, n.grid = 50, theta = 35, phi = 32, zlab = "", contour.col = 'topo', type = 'response',
+              ticktype = "detailed", color = "topo", main = "Interaction by Median Income and Time",
+              xlab = "\n\nDate", ylab = '\n\nMedian Income', nticks = 4)
+dev.off()
+
+# percent black
+model <- gam(
+  formula = deaths ~ te(date_int, prop_black, d = c(1,1)),
+  offset = log(popsize/1e5/12),
+  data = df,
+  family = nb()
+)
+
+png(filename = here("analysis/04_one_variable_over_time/proportion_black_and_time.png"))
+mgcv::vis.gam(model, n.grid = 50, theta = 35, phi = 32, zlab = "", contour.col = 'topo', type = 'response',
+              ticktype = "detailed", color = "topo", main = "Interaction by Proportion Black and Time",
+              xlab = "\n\nDate", ylab = '\n\nProportion Black', nticks = 4)
+dev.off()
+
+# percent crowding
+model <- gam(
+  formula = deaths ~ te(date_int, crowding, d = c(1,1)),
[…]

### 2022-11-09 analysis/05_two_variables_at_a_time/two_variables_at_a_time.R

@@ -1,0 +1,109 @@
+library(here)
+devtools::load_all(".")
+library(mgcv)
+
+# load analytic dataset
+df <- create_analytic_dataset()
+
+# create a numeric version of date
+df$date_int <- as.integer(lubridate::year(df$date)) + (as.integer(lubridate::month(df$date))-1)/12
+
+
+# ICEraceinc and Median Age -----------------------------------------------
+
+# construct model
+model <- bam(
+  formula = deaths ~ te(date_int, ICEraceinc, median_age, d = c(1,1,1)),
+  offset = log(popsize/1e5/12),
+  data = df,
+  family = nb()
+)
+
+# function to select frames and animate them
+plot_two_way_interaction_single_frames <- function(
+    date = 2020.25,
+    view = c('ICEraceinc', 'median_age'),
+    xlab = '\n\nIndex of Concentration at the Extremes\nRacialized Economic Segregation',
+    ylab = '\n\nMedian Age',
+    main = "March 2020"
+) {
+  vis.gam(
+    model,
+    view = view,
+    cond = list(date_int = date),
+    n.grid = 50,
+    theta = 35,
+    phi = 32,
+    zlab = "",
+    contour.col = 'topo',
+    type = 'response',
+    ticktype = "detailed",
+    color = "topo",
+    main = main,
+    xlab = xlab,
+    ylab = ylab,
+    nticks = 4
+  )
+}
+
+# create a set of monthly date-times for animation and corresponding "Month YYYY" titles
+animation_frames <- c(2020 + 0:11/12, 2021 + 0:11/12, 2022 + 0:8/12)[3:32]
+titles <- apply(expand.grid(month.name, 2020:2022), 1, function(x) paste0(x[1], " ", x[2]))[3:32]
+
+# render each animation frame
+for (i in seq_along(animation_frames)) {
+  # open png device/file
+  png(filename = paste0(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/"), stringr::str_pad(i, width = 2, side = 'left', pad = '0'), ".png"),
+      width = 6, height = 6, units = "in", res = 300)
+
+  # create visualization
+  plot_two_way_interaction_single_frames(
+    date = animation_frames[i],
+    main = titles[i]
+  )
+  # close file/device
+  dev.off()
+}
+
+# move into the directory where the frames were animated
+setwd(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/"))
+
[…]

### 2022-11-09 analysis/06_random_forest/random_forest.R

@@ -1,0 +1,48 @@
+library(here)
+library(ranger)
+library(vip)
+devtools::load_all(".")
+
+# create analytic dataset
+df <- create_analytic_dataset()
+
+# create a numeric version of date
+df$date_int <- as.integer(lubridate::year(df$date)) + (as.integer(lubridate::month(df$date))-1)/12
+
+# fit model
+model <-
+  df %>%
+    na.omit() %$%
+    ranger(
+    formula = crude_rate_per_100k_py ~ ICEraceinc + prop_in_poverty + median_income + median_age + prop_65_and_up +
+      prop_75_and_up + crowding + prop_black + prop_hispanic + prop_white_nh + prop_asian + prop_aian + prop_nhopi +
+      pop_density + date_int,
+    data = .,
+    case.weights = .$popsize,
+    importance = "impurity_corrected"
+)
+
+# variable importance plot
+vip::vip(model) +
+  ggtitle("Variable Importance from Random Forest Using Impurity-Corrected Method")
+ggsave(here("analysis/06_random_forest/impurity_corrected_importance.png"))
+
+
+# and if we add in divisions
+model <-
+  df %>%
+  na.omit() %$%
+  ranger(
+    formula = crude_rate_per_100k_py ~ division + ICEraceinc + prop_in_poverty + median_income + median_age + prop_65_and_up +
+      prop_75_and_up + crowding + prop_black + prop_hispanic + prop_white_nh + prop_asian + prop_aian + prop_nhopi +
+      pop_density + date_int,
+    data = .,
+    case.weights = .$popsize,
+    importance = "impurity_corrected"
+  )
+
+# variable importance plot
+vip::vip(model) +
+  ggtitle("Variable Importance from Random Forest Using impurity Method")
+
+

### 2022-11-09 analysis/07_xgboost/xgboost.R

@@ -1,0 +1,51 @@
+
+library(here)
+library(xgboost)
+library(vip)
+devtools::load_all(".")
+
+# create analytic dataset
+df <- create_analytic_dataset()
+
+# create a numeric version of date
+df$date_int <- as.integer(lubridate::year(df$date)) + (as.integer(lubridate::month(df$date))-1)/12
+
+# fit model
+df_mat <-
+  df %>%
+  na.omit() %>%
+  sf::st_drop_geometry() %>%
+  ungroup() %>%
+  select(
+    ICEraceinc,
+    prop_in_poverty,
+    median_income,
+    median_age,
+    prop_65_and_up,
+    prop_75_and_up,
+    crowding,
+    prop_black,
+    prop_hispanic,
+    prop_white_nh,
+    prop_asian,
+    prop_aian,
+    prop_nhopi,
+    pop_density,
+    date_int
+  ) %>%
+  as.matrix()
+
+
+model <- xgboost(
+    # formula = crude_rate_per_100k_py ~ ICEraceinc + prop_in_poverty + median_income + median_age + prop_65_and_up +
+    #   prop_75_and_up + crowding + prop_black + prop_hispanic + prop_white_nh + prop_asian + prop_aian + prop_nhopi +
+    #   pop_density + date_int,
+    label = na.omit(df)$crude_rate_per_100k_py,
+    nrounds = 1000,
+    data = df_mat)
+
+# variable importance plot
+vip::vip(model) +
+  ggtitle("Variable Importance from XGBoost")
+ggsave(here("analysis/07_xgboost/xgboost_variable_importance.png"))
+

### 2022-11-09 analysis/08_spatiotemporal_autocorrelation/spatiotemporal_autocorrelation.R

@@ -1,0 +1,416 @@
+library(here)
+devtools::load_all(".")
+library(ggdist)
+library(mgcv)
+library(mgcViz)
+library(urbnmapr)
+library(spdep)
+library(sf)
+library(gtable)
+library(grid)
+
+# load analytic dataset
+df <- create_analytic_dataset()
+
+# create a numeric version of date
+df$date_int <- as.integer(lubridate::year(df$date)) + (as.integer(lubridate::month(df$date))-1)/12
+
+
+# temporal autocorrelation ------------------------------------------------
+
+
+# helper function to calculate missingness proportion
+missing_proportion <- function(x) { sum(is.na(x)) / length(x) }
+
+# format county observations for calculating ACF
+mat_for_acf <-
+  df %>% sf::st_drop_geometry() %>%
+  ungroup() %>%
+  select(geoid, date, crude_rate_per_100k_py) %>%
+  tidyr::pivot_wider(
+    id_cols = date,
+    names_from = geoid,
+    values_from = crude_rate_per_100k_py
+  ) %>%
+  arrange(date) %>%
+  select(-date) %>%
+  select(where(~ missing_proportion(.x) < .1)) %>%
+  as.matrix()
+
+# specify a maximum lag in months
+lag_max <- 24
+
+# calculate ACF for each county
+ACF_mat <-
+  purrr::map(1:ncol(mat_for_acf), ~ as.numeric(acf(
+    na.omit(mat_for_acf[4:nrow(mat_for_acf), .]), lag.max = lag_max, plot =
+      F
+  )$acf))
+
+# convert to data frame
+ACF_mat %<>% bind_cols()
+
+# add lag indicator column
+ACF_mat %<>% mutate(lag = seq(0,lag_max))
+
+# convert to tidy format
+ACF_mat %<>% tidyr::pivot_longer(
+  cols = setdiff(colnames(.), "lag"),
+  names_to = "id",
+  values_to = "cor"
+)
+
+# visualize distribution of ACF over time
+ggplot(ACF_mat, aes(x = lag, y = cor)) +
+  ggdist::stat_lineribbon() +
+  scale_fill_brewer(palette = 'Reds') +
+  geom_hline(yintercept = .35, linetype = 'dashed') +
+  geom_hline(yintercept = 0, linetype = 'dotted') +
+  geom_hline(yintercept = -.35, linetype = 'dashed') +
[…]

### 2022-11-09 analysis/08_spatiotemporal_autocorrelation/spatiotemporal_autocorrelation.R
updates to readme; new animation figure

@@ -394,14 +394,6 @@ ggplot({county_map %>% left_join(
     filter(! state_abbv %in% c('AK', 'HI')) %>%
     sf::st_transform(crs = "EPSG:2163"), aes(fill = predictions)) +
   geom_sf(size = 0) +
-  # scale_fill_distiller(
-  #   palette = 'Oranges',
-  #   breaks = c(0, 100, 500, 1000),
-  #   direction = 1,
-  #   trans = scales::sqrt_trans(),
-  #   limits = c(0, 500),
-  #   oob = scales::squish
-  # ) +
   scale_fill_viridis_c(
     option = 'turbo',
     direction = 1,

### 2022-11-09 analysis/09_spatiotemporal_models/spatiotemporal_models.R

@@ -1,0 +1,432 @@
+# dependencies
+library(here)
+devtools::load_all(".")
+library(ggdist)
+library(mgcv)
+library(mgcViz)
+library(urbnmapr)
+library(spdep)
+library(sf)
+library(gtable)
+library(grid)
+
+
+# load analytic dataset
+df <- create_analytic_dataset()
+
+# create a numeric version of date
+df$date_int <- as.integer(lubridate::year(df$date)) + (as.integer(lubridate::month(df$date))-1)/12
+
+# add year
+df %<>% mutate(year = lubridate::year(date))
+df %<>% mutate(month = lubridate::month(date))
+
+
+# get county and state maps for using in figures and spatial autocorrelation assessment
+county_map <- urbnmapr::get_urbn_map('counties', sf=TRUE)
+state_map <- urbnmapr::get_urbn_map('states', sf=TRUE)
+
+# calculate county centroids
+center_coords <- sf::st_centroid(county_map)
+center_coord_latlons <- center_coords %>% sf::st_coordinates()
+
+center_coords <- center_coord_latlons %>% as.data.frame() %>% rename('latitude' = X, 'longitude' = Y) %>%
+  bind_cols(county_fips = county_map$county_fips)
+
+# construct spatiotemporal data frame
+spatiotemporal_df <-
+  df %>%
+  left_join(center_coords, by = c('geoid' = 'county_fips'))
+
+# restrict to contiguous 48 states
+spatiotemporal_df %<>% filter(state %in% setdiff(c(state.name, 'District of Columbia'), c("Alaska", "Hawaii")))
+
+# fit bam (e.g., GAM optimized for very large datasets)
+model <- bam(
+  formula = deaths ~ te(latitude, longitude, date_int, d = c(2, 1)),
+  offset = log(popsize/1e5/12),
+  family = nb(),
+  data = spatiotemporal_df
+)
+
+# introduce custom color bar
+my_triangle_colourbar <- function(...) {
+  guide <- guide_colourbar(...)
+  class(guide) <- c("my_triangle_colourbar", class(guide))
+  guide
+}
+guide_gengrob.my_triangle_colourbar <- function(...) {
+  # First draw normal colourbar
+  guide <- NextMethod()
+  # Extract bar / colours
+  is_bar <- grep("^bar$", guide$layout$name)
+  bar <- guide$grobs[[is_bar]]
[…]

### 2022-11-09 analysis/02_plot_county_trends/plot_county_trends.R
add panel figure

@@ -4,6 +4,7 @@
 # setup
 devtools::load_all(".")
 library(here)
+library(Hmisc)
 
 # get data
 df <- create_analytic_dataset()
@@ -47,7 +48,7 @@ df %>%
 ggsave(here("analysis/02_plot_county_trends/county_trends.png"), height = 6, width = 8)
 
 
-df %>%
+plt_orig <- df %>%
   filter(!is.na(division)) %>%
   ggplot(aes(x = date, y = crude_rate_per_100k_py, group = geoid)) +
   geom_point(alpha = .05) +
@@ -61,7 +62,11 @@ df %>%
     alpha = .75,
     color = 'grey20'
   ) +
+  geom_line(
+    mapping = aes(size = ' ')
+  ) +
   facet_wrap( ~ division) +
+  scale_size_manual(values = c(' ' = 1)) +
   scale_y_continuous(
     trans = scales::pseudo_log_trans(sigma = 100),
     breaks = c(0, 250, 1000, 3500, 10000, 30000),
@@ -73,14 +78,84 @@ df %>%
     fill = guide_none(),
     color = guide_none(),
     fill_ramp = guide_legend(title =
-                               'Quantile Range')
+                               'Weighted Quantile Range')
   ) +
   ylab("COVID-19 Mortality Rate per 100,000 Person-Years") +
+  labs(size = 'Weighted Median') +
   ggtitle(
     "County Observations of COVID-19 Mortality Rates by Month and by U.S. County Division",
     "Median and Quantile Ranges Across Counties"
   ) +
   theme(legend.position = 'bottom')
 
-ggsave(here("analysis/02_plot_county_trends/county_trends_by_division.png"), height = 6, width = 10)
+
+legend <- cowplot::get_legend(plt_orig)
+
+# create manually using weighted quantiles --------------------------------
+
+wtd_quantile_df <-
+  df %>%
+  filter(! is.na(division)) %>%
+  group_by(division, date) %>%
+  dplyr::summarize(
+    median = Hmisc::wtd.quantile(x = crude_rate_per_100k_py, weights = popsize, probs = .5, na.rm=T),
+    ci50_high = Hmisc::wtd.quantile(x = crude_rate_per_100k_py, weights = popsize, probs = .75, na.rm=T),
[…]

### 2022-11-09 analysis/09_spatiotemporal_models/spatiotemporal_models.R

@@ -9,6 +9,7 @@ library(spdep)
 library(sf)
 library(gtable)
 library(grid)
+library(cowplot)
 
 
 # load analytic dataset
@@ -359,7 +360,7 @@ vis.gam(
   type = 'response',
   ticktype = "detailed",
   color = "topo",
-  main = "Interaction of Population Density and Time Without\nAdjusting for Spatiotemporal Autocorrelation",
+  main = "Interaction of Political Lean and Time Without\nAdjusting for Spatiotemporal Autocorrelation",
   ylab = '\n\nPolitical Lean',
   xlab = '\n\nDate in Years',
   nticks = 4
@@ -388,7 +389,7 @@ vis.gam(
   type = 'response',
   ticktype = "detailed",
   color = "topo",
-  main = "Interaction of Population Density and Time After\nAdjusting for Spatiotemporal Autocorrelation",
+  main = "Interaction of Political Lean and Time After\nAdjusting for Spatiotemporal Autocorrelation",
   ylab = '\n\nPolitical Lean',
   xlab = '\n\nDate in Years',
   nticks = 4
@@ -427,6 +428,25 @@ dev.off()
 
 
 
+# make composite images ---------------------------------------------------
+
+plot_grid(
+  ggdraw() + draw_image(here("analysis/09_spatiotemporal_models/log_pop_density_and_time_no_spatiotemporal.png")),
+  ggdraw() + draw_image(here("analysis/09_spatiotemporal_models/median_age_and_time_no_spatiotemporal.png")),
+  ggdraw() + draw_image(here("analysis/09_spatiotemporal_models/median_income_and_time_no_spatiotemporal.png")),
+  ggdraw() + draw_image(here("analysis/09_spatiotemporal_models/ICEraceinc_and_time_no_spatiotemporal.png")),
+  ggdraw() + draw_image(here("analysis/09_spatiotemporal_models/political_lean_and_time_no_spatiotemporal.png")),
+
+
+  ggdraw() + draw_image(here("analysis/09_spatiotemporal_models/log_pop_density_and_time.png")),
+  ggdraw() + draw_image(here("analysis/09_spatiotemporal_models/median_age_and_time.png")),
+  ggdraw() + draw_image(here("analysis/09_spatiotemporal_models/median_income_and_time.png")),
+  ggdraw() + draw_image(here("analysis/09_spatiotemporal_models/ICEraceinc_and_time.png")),
[…]

### 2022-11-14 manuscript/manuscript.rmd
Checkin; drafting more intro

@@
 ---
-title: Analyzing COVID-19 Mortality at the County Level in the U.S. Using Generalized Additive
-  Models
-author: "Christian Testa"
-date: '`r Sys.Date()`'
-output: pdf_document
-abstract: this is an abstract
+title: Documenting Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models
+author: 
+  - name: "Christian Testa"
+    institute: "1"
+    email: "ctesta@hsph.harvard.edu"
+    correspondence: true
+institute: 
+  - "1": "Department of Social and Behavioral Sciences, Harvard T.H. Chan School of Public Health"
+date: '`r format(Sys.Date(), "%B %d, %Y")`'
+output: 
+  bookdown::pdf_document2:
+    pandoc_args:
+      - '--lua-filter=scholarly-metadata.lua'
+      - '--lua-filter=author-info-blocks.lua'
+    toc: false
+    number_sections: false
+abstract: |
+  We present three types of applications of generalized additive models (GAMs) to
+  COVID-19 mortality rates in the US for the purpose of advancing methods to
+  document inequities with respect to which communities suffered disproportionate
+  COVID-19 mortality rates. First, GAMs can be used to describe the changing
+  relationship between COVID-19 mortality and county-level covariates
+  (sociodemographic, economic, and political metrics) over time. Second, GAMs can
+  be used to perform spatiotemporal smoothing that pools information over time and
+  space so that county level mortality rates that tend to be noisy due to small
+  population counts or stochasticity can be summarized by a smooth, dynamic latent
+  surface describing the mortality risk associated with geographic locations over
+  time. Third, estimation of the associations between county-level covariates and
+  COVID-19 mortality after controlling for spatiotemporal effects allow for the
+  distinguishing of what patterns in COVID-19 mortality were more plausibly due to
+  geography than variation in county-level covariates. Each of these approaches
+  provides a valuable approach and perspective to documenting inequities in
+  COVID-19 mortality by addressing the question of which populations have suffered
+  the worst burden of COVID-19 mortality taking into account the nonlinear
+  spatial, temporal, and social patterning of disease.
+  \newline
+
+  \noindent Abbreviations used: United States (US), Coronavirus Disease 2019 (COVID-19), Generalized
+  Additive Model (GAM), Centers for Disease Control and Prevention (CDC)
+bibliography: bibliography.bib
+csl: apa.csl
+link-citations: true
+urlcolor: blue
+linkcolor: black
 ---
 
+<!--
+TODO:
+
+Throughout/Overarching: 
+  - Do I need to update all the models to separately control for age? Probably.
+  - Do I want to have a supplementary materials section?
+  - Talk about ecosocial theory (?) (Who benefits from harm done to whom?)
+
+Abstract:
+  - Add link to GitHub repository
+
+Intro: 
+  - add geographic background, like where the first cases were reported, 
+  how COVID exploded in NYC, then the surges in the Midwest and South
+  - add a primer on the relevance of spatiotemporal modeling
+
+Methods: 
+  - add intro to ICEraceinc (methods)
+  - describe negative binomial models
+
+Results:
+  - add figure for temporal two-way interaction effects
+  - present rate ratios (results)
+  - present model comparisons (results, F-tests, AIC)
+
+Discussion:
+  - talk about how COVID has not been a win for health equity (discussion)
+  - talk about limitations to being causal
+  - talk about future directions
+
+Conclusion:
+  - summarize
+-->
+
 ```{r setup, echo = FALSE}
-library(here)
+suppressMessages(library(here))
 
@@
 
-As we enter the third winter with novel coronavirus disease (COVID-19) in the
-United States (U.S.), evidence reflecting intense disparities in COVID-19 mortality rates 
-among socially advantaged and disadvantaged populations persists.  Eliminating
-inequities in health outcomes is a major policy goal of the Biden
-administration, representing an revitalized focus on health justice
-and underlining the importance of adequate data reporting systems that measure
-the prevalent health inequities.  To this end, generalized additive models
-(GAMs) provide a flexible regression framework which can illustrate the evolving
-roles and relationships social conditions have with trends in COVID-19
-mortality.
+As we enter the third winter with the novel coronavirus disease 
+COVID-19 in the United States, evidence documenting the intense disparities in
+COVID-19 mortality rates comparing socially advantaged and disadvantaged
+populations continues to mount. Eliminating inequities in health outcomes has
+been stated as a major policy goal of the Biden administration
+[@the_white_house_executive_2021; @the_white_house_advancing_2022], representing
+a revitalized focus on health justice and underlining the importance of adequate
+data reporting systems that measure the prevalent health inequities. To this
+end, we use generalized additive models (GAMs) as a flexible regression
+framework to illustrate the evolving roles and relationships sociodemographic,
+geographic, and economic conditions have with respect to trends in COVID-19
+mortality. The code, data, and documentation necessary to reproduce the analyses
+contained in this paper are online and free to access at <https://github.com/ctesta01/covid.gradient.estimation>.
 
-# Background
+### Background
 
 Having passed over 1 million COVID-19 deaths in the United States earlier this
-year, and facing uncertain prospects for the third COVID-19 winter looming even
-as new iterations on the COVID-19 vaccines become available, it remains critical
-that inequities in COVID-19 outcomes are documented and analyzed to reckon with
-the unjust and unfair burden of preventable illness.  Even though the first
-vaccines were granted emergency use authorization by the U.S. Food and Drug
-Administration in 2020, with the first shots going in arms in December 2020,
-COVID-19 is still continuing to cause hundreds of deaths a day in the U.S. in
-the late summer and early fall of 2022.  The new bivalent vaccines released at
-the end of August 2022 contain messenger RNA components of the original strain
-as well as the BA.4 and BA.5 lineages in an effort to make the nation's immunity
-more up-to-date and robust against the myriad of phylogenetic directions the
-COVID-19 virus is exploring.  Despite the updated bivalent boosters representing
-a significant step forward in prevention strategy, less than 4% of eligible
-Americans had taken the booster in the first month after it became available. As
-such, and with an enduring history of inequities in health care availability, it
-is not clear that all communities will be equally able to take advantage of the
-new vaccines and inequities in COVID-19 illness and mortality may persist
-despite the technological innovations in vaccine therapy.
+year [@donovan_us_2022], and facing uncertain prospects for the third COVID-19
+winter looming even as new iterations on the COVID-19 vaccines become available,
+it remains critical that inequities in COVID-19 outcomes are documented and
+analyzed to reckon with the unjust and unfair burden of preventable illness.
+Even though the first vaccines were granted emergency use authorization by the
+U.S. Food and Drug Administration in 2020 [@mayo_clinic_history_2022], with the
+first shots going in arms in December 2020, COVID-19 is still continuing to
+cause hundreds of deaths a day in the US in the fall of 2022
+[@noauthor_united_2022].  The new bivalent vaccines released at the end of
+August 2022 contain mRNA sequences from both the original strain as well as
+the recently emergent BA.4 and BA.5 lineages in an effort to make the nation's immunity more
+up-to-date and robust against the myriad of phylogenetic directions the COVID-19
+virus is evolving to explore [@office_of_the_commissioner_coronavirus_2022].
+Despite the updated bivalent boosters representing a significant step forward in
+prevention strategy, less than 4% of eligible Americans had taken the booster in
+the first month after it became available [@bendix_less_2022; @lambert_most_2022]. 
+As such, and with an enduring history of inequities in health care access 
+in the US [@ortega_ending_2021; @blendon_inequities_2002; @bailey_how_2021;
+@okonkwo_covid-19_2021; @rapp_statelevel_2022;
+@chrisler_ageism_2016; @feldman_health_2021; @carpenter_health_2021;
+@whitehead_outness_2016], it is clear that without further intervention not all
+communities will be equally able to benefit from the updated vaccines and
+inequities in COVID-19 illness and mortality may persist despite the
+technological innovations in vaccine technology.
 
-```{r figure 1, fig.cap = "Monthly estimates of COVID-19 mortality per 100,000 residents by state are shown.
-States are in distinct colors to make each state's trajectory visible.
-Reference the online version TODO: XXX"}
-knitr::include_graphics(here("analysis/01_state_trajectories/state_covid19_mortality_trajectories.png"))
+### The Role of Geography in COVID-19
+
+```{r figure-county-trends, fig.cap = "Estimates of monthly COVID-19 mortality rates per 100,000 person-years by county organized by Census Division.
+For each division, the median trendline and quantile ranges are shown weighted by county population size.", out.width = "6in", fig.align = 'center', fig.env = "figure*"}
+knitr::include_graphics(here("analysis/02_plot_county_trends/county_trends_by_division.png"))
 ```
+
+Prior literature has demonstrated that geography has played a significant role
+in the spread and impact of of COVID-19. Methods employed to highlight the role
+geography plays have included quantile regression [@sigler_socio-spatial_2021],
+Besag-York-Molli\'e mixed models [@whittle_ecological_2020], spatial cluster analysis
+[@sugg_mapping_2021], geographically weighted regression [@mollalo_gis-based_2020; @park_covid-19_2021], and others.
+
+In the US context, one of the key aspects to the geographic story of COVID-19's
+spread and diffusion was the early surge of cases and epicenter in New York City
+during March 2020 [@thompson_covid-19_2020] followed by subsequent waves of cases in the
+South and Midwest [@shumaker_covid-19_2020; @scott_these_2020; @glenza_covid_2020]. As Park et al. 
+stated in their article *COVID-19 Deaths in the United States: Shifts in Hot Spots over the Three Phases of the Pandemic and the Spatiotemporally Varying Impact of Pandemic Vulnerability* where they summarized
+US trends from March 2020 to May 2021, "hot spots have shifted from densely populated cities and the states with a high percentage of socially vulnerable individuals to the states with relatively relaxed social distancing requirements, and then to the states with low vaccination rates" [@park_covid-19_2021].
+
+When considering the drivers of the COVID-19 pandemic, it's necessary to note that 
+geography and social conditions are inextricably linked. In July 2021, the CDC
+reported that "the COVID-19 cumulative death rate in non-metropolitan areas has
+exceeded that of metropolitan areas since December 2020," noting that of the
+approximately 1/5th of Americans who live in rural areas, many "are considered
+highly vulnerable according to CDC’s Social Vulnerability Index (SVI), which
+includes factors such as housing, transportation, socioeconomic status, race,
+and ethnicity" [@cdc_location_2021]. Moreover, rural communities often have
+lower health insurance rates, higher disability rates, older populations, and
+limited access to health care. One of CDC's Morbidity and Mortality Weekly
+Reports found that vaccination against COVID-19 was lower in rural communities
+than in urban communities between December 2020 and April 2021
+[@murthy_disparities_2021].
+
+Figure \@ref(fig:figure-county-trends) shows the monthly COVID-19 mortality
+rates for the counties grouped within each of the nine U.S. Census Divisions
+[@us_department_of_commerce_economics_and_statistics_administration_census_2000]. 
+The figure summarizes each division's median mortality rates
+weighted by county population size. Quantile ranges are included to illustrate
+the range of mortality rates observed, again weighted by county population size.
+Weighting based on county population size is used as opposed to equal weighting
+for calculating the median and quantile ranges so that in divisions where low
+population counties predominate, mortality rate estimates that are noisy due to
+small population counts are not weighted equally to more reliable mortality rate
+estimates from larger population counties. Notably, the mortality associated
+with the early surge of cases starting in NYC and spreading through New York,
+New Jersey, and Massachusetts is visible in the Middle Atlantic and New England
+division figures. The figure also illustrates how the first peak in the
+mortality time-series for states in the Midwest (West North Central, East North
+Central) occurred later, in late 2020 and going into early 2021.
+
+### The Social Determinants of COVID-19 Mortality
+
+<!-- 
+Want to cite: 
+* Mude et al (racial disparities meta-analysis);
+* Chen & Krieger on their article "Revealing the Unequal Burden" 
+* Hanage et al COVID-19: US federal accountability for entry, spread and inequities
+* Kim et al on COVID-19 magnifying the effect of health disparities (esp. re: testing (re: care providers, transportation equity))
+* Romano et al (CDC MMWR on trends and disparities in hospitalizations by race/ethnicity and region)
+* Okonkwo et al: accelerating health inequities
+* Del Rios: COVID-19 is an inverse equity story, not a racial equity success story
+* Goldstein/NYT: During the Omicron surge, Black New Yorkers were hospitalized at a rate more than twice that of white residents.
+* Mueller/NYT In Rural America, Covid Hits Black and Hispanic People Hardest
+* Hill et al / KFF COVID-19 Cases and Deaths by Race/Ethnicity: Current Data and Changes Over Time
+* Lo et al The polarization of politics and public opinion and their effects on racial inequality in COVID mortality
+* Johnson and Keating: Whites now more likely to die from covid than Blacks: Why the pandemic shifted 
+* Kaashoek The Evolving Roles of US Political Partisanship and Social Vulnerability in the COVID-19 Pandemic from February 2020 - February 2021
+* 
+* Krieger et al on a warning against static measures
+
+Outline:
+
+Disparities became immediately obvious: 
+  Though many have repeated the oft-quoted line "viruses do not discriminate,"
+  it was immediately clear that COVID-19 would lay bare the sharp inequities
+  that exist in tge US; almost immediately some groups were shown to be
+  experiencing higher rates of COVID-19 infection and mortality. Those groups
+  have been shown to include marginalized racial/ethnic groups, those who do not
+  speak English, transportation insecurity, the elderly, those exposed to
+  chronic pollution, those living in crowded housing or areas with financial
+  insecurity, and those with disabilities, housing instability and/or
+  homelessness, pre-existing conditions (including chronic obstructive pulmonary
+  disease, diabetes, heart diseases, etc.).  Smoking.  Obesity.  Acute Kidney
+  Injury.
+
+There are some aspects of this story that we still don't know: 
+  The details on COVID-19 rates are not availability stratified by both disability
+  and race/ethnicity simultaneously making it difficult to understand how much of
+  racial/ethnic disparities in COVID-19 are attributable to increased disability
+  rates among historically (and even now) marginalized racial/ethnic groups.
+
+  Moreover, it is known that many of the risk factors for increased COVID-19 severity 
+  are prevalent at increased rates among sexual minority persons compared to heterosexual
+  persons, suggesting that sexual minority persons may be at increased risk of 
+  severe COVID-19.
+
+Vaccines didn't right the boat: 
+  Again, when vaccines became available, I think many thought "surely this will
+  be the silver bullet that solves the problem";  inequities in who were able to
+  get vaccine appointments reinforced existing patterns in internet access,
+  technological literacy and availability of transportation. Even after the
+  initial vaccine rollout, inequities persist in vaccination clinic locations
+  with 'vaccine deserts' being more likely to be located in less affluent, less 
+  white communities.
+
+Throughout this timeline, it has become increasingly apparent that local-level
+politics represents a critical factor influencing the trajectory of COVID-19.
+
+COVID is not a racial equity success story: 
+  Finally, racial inequities in COVID-19 in the US have been widening and
+  shrinking during the course of the pandemic, largely widening during surges
+  and shrinking when case rates and mortality have been lower.  During Fall 
+  2022, the age-adjusted COVID-19 mortality rate for white non-Hispanic 
+  people surpassed the mortality rate for Black or African/American people. 
+  This has been explained as being due to those who still remain unvaccinated
+  being more likely to be Republican and white. While reducing racial inequities
+  in COVID-19 mortality rates during Fall 2022, may sound like a good thing, the
+  cumulative burden of COVID-19 leaving in its wake long-COVID, lost loved ones,
+  jobs lost, and damage to individuals' health that may never be reversed that
+  disproportionately falls on marginalized racial/ethnic groups does not
+  represent a health equity success story.
+-->
+
+  - Insert section on prior modeling results
+  - Insert discussion motivating the need for spatiotemporal modeling
 
@@
 
-  * Counts of deaths by month with underlying cause of death as
-  COVID-19 (ICD-10 code U07.1) from CDC WONDER, Provisional Multiple Cause of Death
-  Data
-  * Population Size Estimates for 2020 from the U.S. Census (Redistricting 
-  File Public-Law 94-171 Dataset). *Note:*  While the complete data for the 2020
-  U.S. Census have not yet been released, the summary tables including
-  population counts have been released as mandated by Congress for the purposes
-  of redistricting and updating the boundaries of the 435 voting Congressional
-  Districts, which are what is made us of here as population size estimates.
-  * Percent Black, percent Hispanic, percent Asian, percent American Indian or
-  Alaska Native, Percent below the federal poverty line, median household
-  income, median age, percent 65 years or older, and percent 75 years or older
-  from the 2014-2019 5-year American Community Survey. 
-  *Note:* while more recent American Community Survey 5-year estimates are
-  available, survey data collection in 2020 was severely affected by the
-  pandemic with the 2020 data release denoted as "experimental".
+  * Counts of COVID-19 deaths [@the_new_york_times_coronavirus_2021].
+  * Population size estimates for 2020 from the U.S. Census (Redistricting 
+  File Public-Law 94-171 Dataset) [@us_census_bureau_2020_2021].
+  * Median age, median household income, racial/ethnic composition, population
+  density, percent below the federal poverty line, and number of households with
+  high (\$100k+)/low (<\$25k) household income by racial/ethnic group from the
+  2014-2019 5-year American Community Survey [@us_census_bureau_american_2020]
+  through the `tidyverse` R package [@walker_tidycensus_2022]. 
+  * Votes cast in the 2020 presidential election [@mit_county_2022]
 
@@
 
+Traditional generalized linear models fit a regression model 
+using the functional relation $g(\mu_i) = \mathbf X_i \beta$ 
+where 
+Examples of common link functions include the identity function for linear regression,
+$\log$ for Poisson regression, and the $\text{logit}$ function for logistic regression.
+
+Generalized additive models (GAMs) improve upon generalized linear models by
+allowing for the fitting of smooth functions that transform the $x$ variables.
+This is a convenient means to include nonlinear relationships between the
+outcome and predictor variables. Whereas a generalized linear model may have looked like 
+$$ g(\mu_i) = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} ...$$ 
+a generalized additive model could look like 
+$$ g(\mu_i) = \mathbf A_i \mathbf \theta + f_1(x_{i1}) + f_2(x_{i2}) + f_3(x_{i3}, x_{i4}) + \dots$$
+where $\mu_i \equiv \mathbb E(Y_i)$, $\mathbf A_i$ is a row of the model matrix 
+for any strictly non-parametric model components, $\mathbf \theta$ is the corresponding
+parameter vector, and the $f_j$ are smooth functions of the covariates $x_k$ [@wood_generalized_2017].
+
 # Results
+
+## Evaluating Spatial and Temporal Autocorrelation
+
+```{r figure county neighboring graph, fig.cap = "A) The spatial neighbor graph for counties in the US.
+In this graph, county centroids are connected if the two counties are adjacent to one another (i.e., share a boundary).
+This and similar graphs that represent neighbor relationships or geographic proximity are often used in modeling contexts to create spatial weights that account for which regions help to explain a given region's observed data.", fig.show='hold', out.width = '50%', fig.env = "figure*"}
+desired_caption <- 
+  " B) Moran's scatter plot \\citep{anselin_simple_1996} depicts for each region (in our case, counties) on the x-axis the region rate and on the y-axis the average of the region's neighbors' rates.
+This diagnostic plot provides a simple means to assess to what degree neighboring regions share similar rates.
+In a setting where regions have high correlation, Moran's scatter plot will exhibit a high degree of positive correlation."
+knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/county_neighboring_relations.png")) 
+knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/morans_plot.png"))
+```
+
+```{r figure 4, fig.cap = ""}
+knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/temporal_acf.png"))
+```
+
+## Regression Modeling Results
 
@@
 with the coefficient estimates from the GAMs.
+
+```{r figure 5, fig.cap = ""}
+knitr::include_graphics(here("analysis/09_spatiotemporal_models/temporal_panel_figure.png"))
+```
+
+```{r figure 6, fig.cap = ""}
+knitr::include_graphics(here("analysis/09_spatiotemporal_models/panel_figure.png"))
+```
 
@@
 -->
-
-# Supplementary Materials

### 2022-11-14 analysis/09_spatiotemporal_models/spatiotemporal_models.R

@@ -10,6 +10,7 @@ library(sf)
 library(gtable)
 library(grid)
 library(cowplot)
+library(patchwork)
 
 
 # load analytic dataset
@@ -96,7 +97,7 @@ guide_gengrob.my_triangle_colourbar <- function(...) {
 
 
 # function to plot the model predictions for a given year and month
-visualize_gam_map <- function(model, year, month) {
+visualize_gam_map <- function(model, year, month, fulltitle = TRUE) {
 
   # filter our dataframe for relevant observations
   selected_df <- spatiotemporal_df %>% filter(year == {{ year }}, month == {{ month }})
@@ -122,7 +123,7 @@ visualize_gam_map <- function(model, year, month) {
     labs(fill = "Smoothed\nMortality\nRate") +
     theme_bw() +
     ggtitle(
-      "Spatio-temporally Smoothed Mortality Rate per 100,000 Person-Years",
+      if (fulltitle) "Spatio-temporally Smoothed Mortality Rate per 100,000 Person-Years" else "",
       subtitle = paste0(month.name[month], ", ", year)
     )+
     theme(legend.position = 'right')
@@ -155,6 +156,37 @@ for (i in 1:length(year_vec)) {
 # combine frames to create animation
 setwd(here("analysis/09_spatiotemporal_models/animation/"))
 system("convert -delay 40 -loop 0 *.png spatiotemporal_animation.gif \( +clone -set delay 500 \) +swap +delete animation_with_pause.gif")
+
+# create temporal panel figure
+
+visualize_gam_map(model, year_vec[5], month_vec[5], fulltitle = F) +
+visualize_gam_map(model, year_vec[6], month_vec[6], fulltitle = F) +
+visualize_gam_map(model, year_vec[7], month_vec[7], fulltitle = F) +
+visualize_gam_map(model, year_vec[8], month_vec[8], fulltitle = F) +
+
+visualize_gam_map(model, year_vec[9], month_vec[9], fulltitle = F) +
+visualize_gam_map(model, year_vec[10], month_vec[10], fulltitle = F) +
+visualize_gam_map(model, year_vec[11], month_vec[11], fulltitle = F) +
+visualize_gam_map(model, year_vec[12], month_vec[12], fulltitle = F) +
+
+visualize_gam_map(model, year_vec[13], month_vec[13], fulltitle = F) +
[…]

### 2022-11-23 manuscript/manuscript.rmd
Working on the manuscript

@@
 linkcolor: black
+header-includes: 
+  - \usepackage{pdflscape}
+  - \newcommand{\blandscape}{\begin{landscape}}
+  - \newcommand{\elandscape}{\end{landscape}}
 ---
@@
 
-In the US context, one of the key aspects to the geographic story of COVID-19's
-spread and diffusion was the early surge of cases and epicenter in New York City
-during March 2020 [@thompson_covid-19_2020] followed by subsequent waves of cases in the
-South and Midwest [@shumaker_covid-19_2020; @scott_these_2020; @glenza_covid_2020]. As Park et al. 
-stated in their article *COVID-19 Deaths in the United States: Shifts in Hot Spots over the Three Phases of the Pandemic and the Spatiotemporally Varying Impact of Pandemic Vulnerability* where they summarized
-US trends from March 2020 to May 2021, "hot spots have shifted from densely populated cities and the states with a high percentage of socially vulnerable individuals to the states with relatively relaxed social distancing requirements, and then to the states with low vaccination rates" [@park_covid-19_2021].
-
-When considering the drivers of the COVID-19 pandemic, it's necessary to note that 
-geography and social conditions are inextricably linked. In July 2021, the CDC
-reported that "the COVID-19 cumulative death rate in non-metropolitan areas has
-exceeded that of metropolitan areas since December 2020," noting that of the
-approximately 1/5th of Americans who live in rural areas, many "are considered
-highly vulnerable according to CDC’s Social Vulnerability Index (SVI), which
-includes factors such as housing, transportation, socioeconomic status, race,
-and ethnicity" [@cdc_location_2021]. Moreover, rural communities often have
-lower health insurance rates, higher disability rates, older populations, and
-limited access to health care. One of CDC's Morbidity and Mortality Weekly
-Reports found that vaccination against COVID-19 was lower in rural communities
-than in urban communities between December 2020 and April 2021
-[@murthy_disparities_2021].
-
 Figure \@ref(fig:figure-county-trends) shows the monthly COVID-19 mortality
@@
 
+In the US context, one of the key aspects to the geographic story of COVID-19's
+spread and diffusion was the early surge of cases and epicenter in New York City
+during March 2020 [@thompson_covid-19_2020] followed by subsequent waves of cases in the
+South and Midwest [@shumaker_covid-19_2020; @scott_these_2020; @glenza_covid_2020]. As Park et al. 
+stated summarized trends in the US from March 2020 to May 2021, "hot spots have shifted from densely populated cities and the states with a high percentage of socially vulnerable individuals to the states with relatively relaxed social distancing requirements, and then to the states with low vaccination rates" [-@park_covid-19_2021].
+
+When considering the drivers of the COVID-19 pandemic, it's necessary to note that 
+geography and social conditions are inextricably linked. In July 2021, the CDC
+reported that "the COVID-19 cumulative death rate in non-metropolitan areas has
+exceeded that of metropolitan areas since December 2020," noting that of the
+approximately 1/5th of Americans who live in rural areas, many "are considered
+highly vulnerable according to CDC’s Social Vulnerability Index (SVI), which
+includes factors such as housing, transportation, socioeconomic status, race,
+and ethnicity" [@cdc_location_2021]. Moreover, rural communities often have
+lower health insurance rates, higher disability rates, older populations, and
+limited access to health care. One of CDC's Morbidity and Mortality Weekly
+Reports found that vaccination against COVID-19 was lower in rural communities
+than in urban communities between December 2020 and April 2021
+[@murthy_disparities_2021].
+
 ### The Social Determinants of COVID-19 Mortality
 
-<!-- 
-Want to cite: 
-* Mude et al (racial disparities meta-analysis);
-* Chen & Krieger on their article "Revealing the Unequal Burden" 
-* Hanage et al COVID-19: US federal accountability for entry, spread and inequities
-* Kim et al on COVID-19 magnifying the effect of health disparities (esp. re: testing (re: care providers, transportation equity))
-* Romano et al (CDC MMWR on trends and disparities in hospitalizations by race/ethnicity and region)
-* Okonkwo et al: accelerating health inequities
-* Del Rios: COVID-19 is an inverse equity story, not a racial equity success story
-* Goldstein/NYT: During the Omicron surge, Black New Yorkers were hospitalized at a rate more than twice that of white residents.
-* Mueller/NYT In Rural America, Covid Hits Black and Hispanic People Hardest
-* Hill et al / KFF COVID-19 Cases and Deaths by Race/Ethnicity: Current Data and Changes Over Time
-* Lo et al The polarization of politics and public opinion and their effects on racial inequality in COVID mortality
-* Johnson and Keating: Whites now more likely to die from covid than Blacks: Why the pandemic shifted 
-* Kaashoek The Evolving Roles of US Political Partisanship and Social Vulnerability in the COVID-19 Pandemic from February 2020 - February 2021
-* 
-* Krieger et al on a warning against static measures
+Even since the beginning of the COVID-19 outbreak in the US, data reflected
+sharp inequities in mortality rates.  During January 22nd to May 5th 2020,
+county COVID-19 mortality rates were 4.94 (95% CI 4.78, 5.09) times higher in
+counties in the highest quintile of percent People of Color (61%-100%) compared
+to counties in the lowest quintile of percent People of Color (0%-17.2%)
+[@chen_revealing_2021].  This was not wholly unanticipated: as COVID-19 was
+beginning to take off in the US, some were already calling attention to the fact
+that COVID-19 threatened to exacerbate existing disparities
+[@kim_covid-19_2020]. Kim, Marrast, and Conigliaro noted three structural
+barriers in COVID-19 prevention and care: 1) originally requiring New York
+residents to have a doctor's prescription for a COVID-19 test reduced the
+opportunity for healthcare for People of Color as they are less likely to have a
+primary care provider; 2) drive-through testing made testing disadvantaged those
+who rely on public transportation; 3) quarantining at home while waiting the
+7-10 days originally required for COVID-19 test results to come back posed an
+economic and social challenge that many in already financially difficult
+situations may not have been able to take on [-@kim_covid-19_2020]. Others noted
+yet more reasons why COVID-19 threatened to worsen an already inequitable
+healthcare landscape in the US: those who reside in prisons and jails,
+immigrants and undocumented people, people with disabilities, and people
+experiencing homeless all face additional challenges in seeking and getting the
+healthcare they deserve [@okonkwo_covid-19_2021]. It's not just a hypothetical 
+that those with more structurally enfranchised privileges have been more able to
+mitigate their risk of negative health outcomes associated with COVID-19 
+— during February 1 to April 1, 2020, New York City residents who lived in more affluent 
+neighborhoods were more likely to have left the city, while New York City residents
+from more Black and Hispanic neighborhoods were more likely to continue working [@coven_disparities_2020].
 
-Outline:
+> Contrary to the oft used phrase that the ‘virus does not discriminate’, the
+data presented here suggest that this virus, as many other infectious diseases,
+has the greatest implications for the most vulnerable people. The intersections
+between health and human rights are clear—the health of a society and
+vulnerability to a pandemic are directly related to its human rights track
+record for those who are marginalised. [@okonkwo_covid-19_2021]
 
-Disparities became immediately obvious: 
-  Though many have repeated the oft-quoted line "viruses do not discriminate,"
-  it was immediately clear that COVID-19 would lay bare the sharp inequities
-  that exist in tge US; almost immediately some groups were shown to be
-  experiencing higher rates of COVID-19 infection and mortality. Those groups
-  have been shown to include marginalized racial/ethnic groups, those who do not
-  speak English, transportation insecurity, the elderly, those exposed to
-  chronic pollution, those living in crowded housing or areas with financial
-  insecurity, and those with disabilities, housing instability and/or
-  homelessness, pre-existing conditions (including chronic obstructive pulmonary
-  disease, diabetes, heart diseases, etc.).  Smoking.  Obesity.  Acute Kidney
-  Injury.
+When vaccines became available, vaccine appointments were often only available
+to be scheduled through online web-portals contributing to the inequities
+between those who had internet access and technological literacy and those who
+didn't [@press_inequities_2021]. Vaccination sites have not been equally
+distributed and areas determined to be vaccine deserts have been found to have
+disproportionately more Black and Hispanic residents [@rader_spatial_2021]. In
+fact, healthcare facilities in counties with higher Black composition had 32%
+(95% CI 14%-47%) lower odds of serving as vaccine sites
+[@hernandez_disparities_2022].
 
-There are some aspects of this story that we still don't know: 
-  The details on COVID-19 rates are not availability stratified by both disability
-  and race/ethnicity simultaneously making it difficult to understand how much of
-  racial/ethnic disparities in COVID-19 are attributable to increased disability
-  rates among historically (and even now) marginalized racial/ethnic groups.
-
-  Moreover, it is known that many of the risk factors for increased COVID-19 severity 
-  are prevalent at increased rates among sexual minority persons compared to heterosexual
-  persons, suggesting that sexual minority persons may be at increased risk of 
-  severe COVID-19.
-
-Vaccines didn't right the boat: 
-  Again, when vaccines became available, I think many thought "surely this will
-  be the silver bullet that solves the problem";  inequities in who were able to
-  get vaccine appointments reinforced existing patterns in internet access,
-  technological literacy and availability of transportation. Even after the
-  initial vaccine rollout, inequities persist in vaccination clinic locations
-  with 'vaccine deserts' being more likely to be located in less affluent, less 
-  white communities.
-
-Throughout this timeline, it has become increasingly apparent that local-level
-politics represents a critical factor influencing the trajectory of COVID-19.
-
-COVID is not a racial equity success story: 
-  Finally, racial inequities in COVID-19 in the US have been widening and
-  shrinking during the course of the pandemic, largely widening during surges
-  and shrinking when case rates and mortality have been lower.  During Fall 
-  2022, the age-adjusted COVID-19 mortality rate for white non-Hispanic 
-  people surpassed the mortality rate for Black or African/American people. 
-  This has been explained as being due to those who still remain unvaccinated
-  being more likely to be Republican and white. While reducing racial inequities
-  in COVID-19 mortality rates during Fall 2022, may sound like a good thing, the
-  cumulative burden of COVID-19 leaving in its wake long-COVID, lost loved ones,
-  jobs lost, and damage to individuals' health that may never be reversed that
-  disproportionately falls on marginalized racial/ethnic groups does not
-  represent a health equity success story.
--->
-
-  - Insert section on prior modeling results
-  - Insert discussion motivating the need for spatiotemporal modeling
+What vaccination has been administered hasn't suddenly erased the unequal burden
+of COVID-19 either; in August 2022 the New York Times was reporting "Black death
+rates at this winter’s peak were greater than those of white people by 34
+percent in rural areas, 40 percent in small or medium cities and 57 percent in
+big cities and their suburbs" [@goldstein_during_2022]. As COVID-19 case and
+mortality rates have waxed and waned, the inequities have widen and shrunken,
+often with racial/ethnic inequities growing during times when COVID-19 rates
+have surged [@hill_covid-19_2022]. During 2022, the age-standardized
+COVID-19 mortality rates for white people have, at times, been slightly 
+higher than those of Black and Hispanic people, predominantly because the 
+mortality rate among white people has increased [@johnson_whites_2022]. 
+It's important to note that white COVID-19 mortality rates overtaking 
+the Black mortality rates does not imply that equity has been established: neither
+does this undo the cumulative impact of mortality (which has been twice as high for
+Black people compared to white people [@hill_covid-19_2022]), nor does it imply 
+the underlying systemic barriers to equity have been overturned
+[@del_rios_covid-19_2022]. As Del Rios, Chomilo, and Lewis note, instead,
+COVID-19 leaves in its wake more years of life expectancy lost, wages lost, and
+degradation of mental health in Communities of Color.
 
@@
   through the `tidyverse` R package [@walker_tidycensus_2022]. 
-  * Votes cast in the 2020 presidential election [@mit_county_2022]
+  * Votes cast in the 2020 presidential election [@mit_election_data_and_science_lab_county_2022]
 
@@
 
-Traditional generalized linear models fit a regression model 
-using the functional relation $g(\mu_i) = \mathbf X_i \beta$ 
-where 
-Examples of common link functions include the identity function for linear regression,
-$\log$ for Poisson regression, and the $\text{logit}$ function for logistic regression.
+Traditional generalized linear models fit a regression model using the
+functional relation $g(\mu_i) = \mathbf X_i \beta$ where Examples of common link
+functions include the identity function for linear regression, $\log$ for
+Poisson regression, and the $\text{logit}$ function for logistic regression.
 
-Generalized additive models (GAMs) improve upon generalized linear models by
+In contrast, generalized additive models (GAMs) improve upon generalized linear models by
 allowing for the fitting of smooth functions that transform the $x$ variables.
@@
 
+```{r gam-demo-figure, fig.cap = "A demonstration of fitting regression models using B-spline basis functions.
+Part A.
+shows the simulated dataset.
+Part B.
+shows the B-spline basis functions.
+Part C.
+shows the weighting of the B-spline basis functions and their weighted sum with an intercept added to create the regression model fit to the data."}
+
+knitr::include_graphics(here("analysis/10_supplementary_gam_figures/gam_demo_figure.png"))
+```
+
+The smooth functions estimated as part of fitting a GAM are constructed using
+spline basis functions. These spline basis functions allow for the smooth
+interpolation of trends in the data allowing for the incorporation of
+nonlinearity. To avoid overfitting the data, a penalty term is introduced that
+controls the degree of "wiggliness" or smoothness and this penalty term is fit
+using generalized cross validation. B-splines are one kind of spline basis function 
+that is commonly used and are especially popular because of their property that they 
+are non-negative on only a finite interval.
+
+While a variety of varying kinds of spline bases exist (cubic splines, B- and
+P-splines, thin-plate splines, etc.), our particular application setting
+warrants the use of tensor-product smooths because we require anisotropic
+penalization. That is to say, when we seek to create spatiotemporally smoothed
+model estimates, it's inappropriate to assume that the amount of smoothing
+across space should be the same as the amount of smoothing across time because
+they're in fundamentally different units (space being measured in Cartesian
+coordinates and time being measured in years, months, days, etc.). More details
+about tensor-product smooths are available in Simon Wood's *Generalized Additive Models - An Introduction with R* [-@wood_generalized_2017].
+
+## Overdispersion and Event Modeling
+
+Above and beyond using the GAM framework to allow for flexible, nonlinear
+relationships between our observed county-level variables and COVID-19
+mortality, we must have a model specification that agrees with the data
+generating process. In our case, the data generated are counts of deaths per
+population, and the class of models most suited to represent counts of events
+are Poisson, quasi-Poisson, and negative binomial models. Here we've chosen to 
+use the negative binomial model as it accounts for overdispersion and has 
+the intuitive interpretation of a Poisson model with gamma distributed underlying
+rate parameter [@gelman_bayesian_2013].
+
+A negative binomial model in the context of modeling count data can be written
+$$y_i \sim \text{NegativeBinomial}(u_i \theta_i, \phi),$$
+where $y_i$ are the counts observed, $u_i$ is the exposure, $\theta_i$ are
+the rates, and $\phi$ determines the amount of overdispersion. The rates
+are modeled as $\theta_i = e^{X_i \beta}$ where $X_i$ are the observed
+covariates and $\beta$ are the coefficients on the covariates corresponding
+to log rate ratios. The logarithm of the exposure, $\log(u_i)$ is often called 
+(and later herein referenced as) the offset. In epidemiological contexts, the offset 
+is often representative of the amount of person-time during which observations were recorded. Whereas the Poisson model holds that $\text{var}(y) = \mu$ where $\mu$
+is the average rate, the negative binomial model instead assumes that
+$\text{var}(y) = \mu + \mu^2/\phi$. The Poisson model
+is a special case of the negative binomial model when $\phi \to \infty$.
+
+An alternative and equivalent formulation of the negative binomial that is commonly 
+used makes the connection to the Poisson model even more clear: instead of 
+$\theta_i$ and $\phi$, using $\alpha$ and $\beta$, 
+$$y \sim \text{NegativeBinomial}(\alpha, \beta), \text{ and}$$
+$$\text{NegativeBinomial}(y|\alpha, \beta) = \int \text{Poisson}(y|\theta) \text{Gamma}(\theta | \alpha, \beta) d\theta.$$
+
+Introductions to and additional exposition on the negative binomial model,
+especially in the context of modeling counts of events outcome data, are
+available in *Bayesian Data Analysis* and *Regression and Other Stories*
+[@gelman_regression_2020; @gelman_bayesian_2013].
+
+## Variables of Interest
+
+Besides age, which is the greatest risk factor for COVID-19 mortality, we have
+included the following variables as covariates of interest:
+
+  * Population Density per Square Mile
+  * Median Household Income
+  * Proportion in Poverty
+  * the Index of Concentration at the Extremes for Racialized Economic Segregation 
+  * Political Lean in the 2020 Election (1 = 100% Republican votes, -1 = 100% Democratic Votes)
+
+#### Index of Concentration at the Extremes for Racialized Economic Segregation
+
+The Index of Concentration at the Extremes (ICE) is a measure which describes
+how concentrated a given area's population is in terms of the extreme ends of
+privilege and marginalization [@krieger_public_2016]. 
+In general, the ICE measure is formulated as
+
+$$ \text{ICE} = \frac{\text{Number of People in Most Privileged Category} - \text{Number of People in Least Privileged Category}}{\text{Total Population}}$$
+
+Applying the ICE approach to a specific context involves defining the axes of
+privilege of interest.  In this case, data on racialized economic segregation
+are used from the US Census American Community Survey on White households
+earning more than $100,000 a year (the most privileged group) or households of
+People of Color earning less than $25,000 a year (the least privileged group).
+This variable is referred to throughout as the ICE for racialized economic
+segregation, or `ICEraceinc` in the code.
+
+Compared with the Gini
+coefficient which is one of the most popular methods for summarizing area-level
+rates of inequities, the ICE the major advantage that it is more suitable for
+describing inequities at smaller area levels [@krieger_public_2016]. While the
+Gini coefficient measures within-area dissimilarity (as in, for example, how
+unequal wealth is distributed within a county), the ICE measure establishes
+where on a spectrum a given county's population falls allowing for comparison
+across counties. The Gini coefficient suffers from the fact that areas which are
+made up of relatively homogeneous populations will appear as having low
+within-area inequality (and therefore low Gini coefficient). Instead, the ICE
+measures how much of the population is either privileged or not. The Gini
+coefficient remains useful for reporting on the degree of inequity in larger
+areal units (like countries, states, and regions), but at smaller areal units
+(like counties, ZIP codes, census tracts) can be more difficult to interpret and
+compare. Maps of ICE measure can elucidate what spatial social segregation and
+polarization exist, and the ICE for racialized economic segregation has been
+repeatedly and significantly associated with COVID-19 outcomes
+[@hanage_covid-19_2020; @chen_revealing_2021; @krieger_relationship_2022;
+@eichenbaum_health_2022; @brown_ecological_2021; @sonderlund_racialized_2022;
+@saha_neighborhood-level_2020].
+
+#### Political Lean
+
+Political lean has been associated with COVID-19 mortality in numerous studies,
+with plausible mechanisms explaining the association including differences in
+non-pharmaceutical intervention uptake (mask usage, social distancing,
+quarantining), differences in rates of vaccination, differences in political
+leadership's messaging, resource allocation, and the adoption of policy
+interventions [@krieger_relationship_2022; @kaashoek_evolving_2021; @leonhardt_red_2021; @gonzalez_conservatism_2021; @grossman_political_2020].
+
 # Results
+
+## Application 1: Non-Spatial Covariate Effects Over Time
+
+The GAM models are fit with the following structure using 
+the `gam` function from the `mgcv` package [@wood_mgcv_2022]:
+
+```{r gam-demo-code, echo = TRUE, eval=FALSE}
+gam(
+  formula = deaths ~ s(median_age) + te(covariate, date, d=c(1,1)), # regression formula
+  offset = log(popsize/1e5/12), # our offset represents the person-time
+  data = df,     # our data-set of county-level observations
+  family = nb(link='log')  # indicates negative binomial family and a log-link function
+)
+```
+
+The formula used puts a one-dimensional smoothing spline to represent a
+nonlinear age-effect and a two-dimensional tensor-product smooth on the
+interaction between the given covariate and the date. The `d=c(1,1)` argument
+provides the instruction necessary to consider covariate and date as being on
+separate scales and therefore to fit the tensor-product smooth with anisotropy —
+that is, to allow for independent amounts of scaling in the dimensions of the
+covariate and time. The offset used structures the regression to estimate rates
+in units of person-time per 100,000 person-years. Since the death counts are
+aggregated to the monthly level, the person-time in units of 100,000
+person-years are calculated by taking each county's population size, dividing by
+100,000, and dividing by 12 (for the 12 months in a year).
+
+The above structure is used to estimate models for our different `covariate`
+variables of interest: median income, percent in poverty, the ICE for racialized
+economic segregation, political lean.  Whereas the other models treat median age
+as a variable to adjust for, the model presenting the effect of median age over
+time treats median age as the main covariate of interest including it as the 
+`te(covariate, date, d=c(1,1))` and dropping the `s(median_age)` term which is 
+redundant. Results of these models are summarized in \@ref(fig:one-variable-over-time)
+where the additional COVID-19 mortality associated with each covariate is visualized.
+
+(ref:medianage) [@us_census_bureau_nation_2022]
+
+```{r one-variable-over-time, fig.cap = "Additional COVID-19 mortality associated with covariates over time.
+A) age and time, B) log (base 10) population density and time, C) median income and time, D) proportion in poverty and time, E) ICEraceinc and time, F) political lean and time.
+In panels B-F the effect of age is marginalized out using the median age in the US, 38.8 (ref:medianage).", fig.align = 'center', out.height = '75%'}
+knitr::include_graphics(here("analysis/04_one_variable_over_time/panel_figure.png"))
+```
+
+Likelihood ratio tests confirmed that models with covariates interacted with
+time had significantly lower residual deviance ($p \leq 2.2e^{-16}$ for all
+models) compared to models only including a spline term on median age and the
+given covariate not interacted with time. The model interacting median age 
+and time was compared to a model with a spline for median age alone. Akaike Information
+Criteria values were also lower for all models compared to models that did
+not interact the covariates with time.
+
+Another non-spatial application of GAMs to describe the distribution of 
+COVID-19 mortality in the US over time is to consider the associations of 
+mortality with three-way interactions of time and two covariates taken together. 
+In the following example, the formula used is `deaths ~ te(date, ICEraceinc, median_age, d=c(1,1,1))`.  Again, the `d` argument specifying the marginal 
+basis dimensions is used indicate that each of the `date`, `ICEraceinc`, and 
+`median_age` measures are in different units and should not be smoothed assuming
+that a one unit difference in one variable is comparable to a one unit difference
+in another variable.
+
+```{r two-variables-over-time, fig.cap="The changing interacted effect of the ICE for racialized economic segregation and age over time.
+A) May 2020, B) November 2020, C) May 2021, D) October 2021, E) February 2022, F) August 2022.
+An animated version is available online at "}
+knitr::include_graphics(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/panel_figure.png"))
+```
 
@@
 
-  * Missing data 
   * Ecological fallacy 
@@
   * Vaccination and mobility effects
+  * Equal Smoothing Across All Areas
+  * Soap Film Smoothers
 

### 2022-11-23 analysis/04_one_variable_over_time/one_variable_over_time.R

@@ -1,6 +1,8 @@
 library(here)
 library(mgcv)
 library(mgcViz)
+library(patchwork)
+library(cowplot)
 devtools::load_all(".")
 
 # load our analytic dataset
@@ -9,9 +11,25 @@ df <- create_analytic_dataset()
 # create a numeric version of date
 df$date_int <- as.integer(lubridate::year(df$date)) + (as.integer(lubridate::month(df$date))-1)/12
 
+# median income
+model <- gam(
+  formula = deaths ~ s(median_age) + te(date_int, median_income, d = c(1,1)),
+  offset = log(popsize/1e5/12),
+  data = df,
+  family = nb()
+)
+png(filename = here("analysis/04_one_variable_over_time/income_and_time.png"))
+mgcv::vis.gam(model, n.grid = 50, theta = 35, phi = 32, zlab = "", contour.col = 'topo', type = 'response',
+              cond = list(median_age = 38.8),
+              view = c('date_int', 'median_income'),
+              ticktype = "detailed", color = "topo", main = "Interaction by Median Income and Time",
+              xlab = "\n\nDate", ylab = '\n\nMedian Income', nticks = 4)
+dev.off()
+
+
 # poverty
 model <- gam(
-  formula = deaths ~ te(date_int, prop_in_poverty, d = c(1,1)),
+  formula = deaths ~ s(median_age) + te(date_int, prop_in_poverty, d = c(1,1)),
   offset = log(popsize/1e5/12),
   data = df,
   family = nb()
@@ -21,74 +39,41 @@ png(
   filename = here("analysis/04_one_variable_over_time/poverty.png")
 )
 mgcv::vis.gam(model, n.grid = 50, theta = 35, phi = 32, zlab = "", contour.col = 'topo', type = 'response',
+              cond = list(median_age = 38.8),
+              view = c('date_int', 'prop_in_poverty'),
               ticktype = "detailed", color = "topo", main = "Interaction by Poverty and Time",
               xlab = "\n\nDate", ylab = '\n\nProportion in Poverty', nticks = 4)
 dev.off()
 
-
-# median income
-model <- gam(
-  formula = deaths ~ te(date_int, median_income, d = c(1,1)),
-  offset = log(popsize/1e5/12),
-  data = df,
-  family = nb()
-)
-png(filename = here("analysis/04_one_variable_over_time/income_and_time.png"))
[…]

### 2022-11-23 analysis/05_two_variables_at_a_time/two_variables_at_a_time.R

@@ -107,3 +107,19 @@ setwd(here("analysis/05_two_variables_at_a_time/animation_income_age/"))
 # use imagemagick to create a gif
 system("convert -delay 40 -loop 0 *.png income_age_animation.gif \( +clone -set delay 500 \) +swap +delete animation_with_pause.gif")
 
+
+# Panel figure
+plot_grid(
+  ggdraw() + draw_image(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/03.png")),
+  ggdraw() + draw_image(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/09.png")),
+  ggdraw() + draw_image(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/15.png")),
+  ggdraw() + draw_image(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/20.png")),
+  ggdraw() + draw_image(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/24.png")),
+  ggdraw() + draw_image(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/30.png")),
+
+  labels = 'AUTO',
+  nrow = 2
+)
+
+ggsave(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/panel_figure.png"), width = 8, height = 5.5, bg='white')
+

### 2022-11-23 analysis/10_supplementary_gam_figures/supplemental_gam_figure.R

@@ -1,0 +1,73 @@
+# supplementary gam figures
+library(ggplot2)
+library(tibble)
+library(magrittr)
+library(dplyr)
+library(splines)
+library(here)
+
+df <- tibble::tibble(x = seq(0,10,length.out=100) + rnorm(n = 100, sd = .25),
+                 y = (x-6)^3/100 + x^2/100 + rnorm(n=100, sd = .1) + 5)
+
+
+set.seed(2022)
+n <- 100
+df <- tibble::tibble(x = seq(0,10,length.out=n),
+                 y = (x-6)^3/100 + rnorm(n=n, sd = .1) + 2.5)
+
+ggplot(df, aes(x = x, y = y)) +
+  geom_point() +
+  theme_bw() +
+  ggtitle("A. Simulated Sample Observations") ->
+  scatterplot
+
+basis <- bs(seq(0,10, length.out=1000), df = 15)
+
+basis_tidy <- basis %>%
+  as.data.frame() %>%
+  mutate(x = seq(1,10,length.out=1000)) %>%
+  tidyr::pivot_longer(
+    cols = setdiff(colnames(.), "x"),
+    names_to = "basis",
+    values_to = 'y'
+  )
+
+basis_tidy %>% # filter(basis != '15') %>%
+  ggplot(aes(x = x, y = y, color = basis)) +
+  geom_line() +
+  theme_bw()+
+  scale_color_viridis_d() +
+  theme(legend.position = 'none') +
+  ggtitle("B. B-Spline Basis Functions") ->
+  basis_plot
+
+model <- lm(y ~ bs(x, df = 15), data = df)
+
+basis_fit <- t( coef(model)[2:16] * t(basis[,c(1:15)]) ) %>%
+  rowSums(na.rm = T) + coef(model)[1]
+
+basis_expansion <- t( coef(model)[2:16] * t(basis[,c(1:15)]) ) %>%
+  as.data.frame() %>%
+  mutate(x = seq(0,10,length.out=1000)) %>%
+  tidyr::pivot_longer(
+    cols = setdiff(colnames(.), "x"),
+    names_to = "basis",
+    values_to = 'y'
+  )
+
+ggplot(df, aes(x = x, y = y)) +
+  geom_line(data = data.frame(x = seq(0, 10, length.out = 1000), y = basis_fit), size = 1.5, alpha = 0.8, color = 'blue') +
+  geom_point(alpha = 0.8) +
[…]

### 2022-11-27 manuscript/manuscript.rmd
adding readme page for spatiotemporal animation

@@
 ---
-title: Documenting Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models
+title: Understanding Complex Patterns in Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models
 author: 
@@
 abstract: |
-  We present three types of applications of generalized additive models (GAMs) to
+  I present three types of applications of generalized additive models (GAMs) to
   COVID-19 mortality rates in the US for the purpose of advancing methods to
   document inequities with respect to which communities suffered disproportionate
-  COVID-19 mortality rates. First, GAMs can be used to describe the changing
+  COVID-19 mortality rates at specific times during the first three years of the 
+  pandemic. First, GAMs can be used to describe the changing
   relationship between COVID-19 mortality and county-level covariates
@@
   be used to perform spatiotemporal smoothing that pools information over time and
-  space so that county level mortality rates that tend to be noisy due to small
+  space to address statistical instability due to small
   population counts or stochasticity can be summarized by a smooth, dynamic latent
   surface describing the mortality risk associated with geographic locations over
-  time. Third, estimation of the associations between county-level covariates and
-  COVID-19 mortality after controlling for spatiotemporal effects allow for the
-  distinguishing of what patterns in COVID-19 mortality were more plausibly due to
-  geography than variation in county-level covariates. Each of these approaches
-  provides a valuable approach and perspective to documenting inequities in
+  time. Third, estimation of COVID-19 mortality associations with county-level covariates
+  conditional on smooth spatiotemporal variation allows for more rigorous consideration
+  of how socio-environmental contexts and policies may have impacted COVID-19 mortality.
+  Each of these approaches
+  provides a valuable perspective to documenting inequities in
   COVID-19 mortality by addressing the question of which populations have suffered
@@
 [@the_white_house_executive_2021; @the_white_house_advancing_2022], representing
-a revitalized focus on health justice and underlining the importance of adequate
-data reporting systems that measure the prevalent health inequities. To this
-end, we use generalized additive models (GAMs) as a flexible regression
+a revitalized commitment to health equity and underlining the importance of adequate
+data reporting systems that report timely estimates of prevalent health inequities. To this
+end, I use generalized additive models (GAMs) as a flexible regression
 framework to illustrate the evolving roles and relationships sociodemographic,
@@
 the range of mortality rates observed, again weighted by county population size.
-Weighting based on county population size is used as opposed to equal weighting
-for calculating the median and quantile ranges so that in divisions where low
-population counties predominate, mortality rate estimates that are noisy due to
-small population counts are not weighted equally to more reliable mortality rate
-estimates from larger population counties. Notably, the mortality associated
+Notably, the mortality associated
 with the early surge of cases starting in NYC and spreading through New York,
@@
 experiencing homeless all face additional challenges in seeking and getting the
-healthcare they deserve [@okonkwo_covid-19_2021]. It's not just a hypothetical 
-that those with more structurally enfranchised privileges have been more able to
-mitigate their risk of negative health outcomes associated with COVID-19 
-— during February 1 to April 1, 2020, New York City residents who lived in more affluent 
-neighborhoods were more likely to have left the city, while New York City residents
-from more Black and Hispanic neighborhoods were more likely to continue working [@coven_disparities_2020].
+healthcare they deserve [@okonkwo_covid-19_2021]. Despite stay-at-home orders
+that were prevalent in many states [@moreland_timing_2020], workers functioning in 
+capacities essential to the functioning of critical infrastructure operations (later
+termed "essential workers") were exposed to heightened risk of COVID-19 transmission 
+[@national_conference_of_state_legislatures_covid-19_2021; @wei_risk_2022; @national_bureau_of_economic_research_measuring_2021; @the_lancet_plight_2020; @hanage_covid-19_2020].
+It's not just a hypothetical that those with more structurally enfranchised
+privileges have been more able to mitigate their risk of negative health
+outcomes associated with COVID-19 — during February 1 to April 1, 2020, New York
+City residents who lived in more affluent neighborhoods were more likely to have
+left the city, while New York City residents from more Black and Hispanic
+neighborhoods were more likely to continue working [@coven_disparities_2020].
 
@@
 A) May 2020, B) November 2020, C) May 2021, D) October 2021, E) February 2022, F) August 2022.
-An animated version is available online at "}
+An animated version is available online at [https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/readme.md](https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/readme.md)"}
 knitr::include_graphics(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/panel_figure.png"))
@@
 
-## Evaluating Spatial and Temporal Autocorrelation
+## Application 2: Spatiotemporal Smoothing
+
+By fitting GAMs with a tensor-product smoothing term on latitude, longitude, and
+time we can estimate a spatiotemporally smoothed trend in COVID-19 mortality. To
+do this, the GAM is constructed similarly as in Application 1 but with the
+formula specified as `deaths ~ te(latitude, longitude, time, d=c(2,1))` where
+`d=c(2,1)` indicates that `latitude` and `longitude` share the same dimensions
+(i.e., both are spatial and in units of degrees) while the `time` data are in separate
+units.
+
+Using GAMs to present spatiotemporally smoothed estimates of COVID-19 mortality
+allows for the synthesis of trends over time. Whereas the raw rates of COVID-19
+mortality are noisy due to 72% of the land area in the US being classified as
+rural and low population density, the spatiotemporally smoothed GAM estimates
+use population weighting via the offset specified to estimate a latent surface
+that represents localized mortality rate averages in spatial coordinates and in
+time.
+
+```{r crude-vs-smoothed, fig.cap = "Comparison of crude vs.
+smoothed mortality rates.
+A) Crude mortality rates per 100,000 person-years in January 2022.
+B) Smoothed mortality rates from a GAM applied to mortality rates from January 2022."}
+knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/side_by_side_crude_and_smoothed.png"))
+```
+
+```{r spatiotemporal-panel-figure, fig.cap = "Spatiotemporally smoothed COVID-19 mortality estimates from a GAM fit to data from March 2020-August 2022.
+Panels show months selected to highlight the changing spatial patterns of COVID-19 mortality risk over time.
+An animated version is available online at []()"}
+knitr::include_graphics(here("analysis/09_spatiotemporal_models/temporal_panel_figure.png"))
+```
+
+## Application 3: Estimating Covariate Effects Adjusted for Spatiotemporal Autocorrelation
+
+### Evaluating Spatial and Temporal Autocorrelation
 
@@
 with the coefficient estimates from the GAMs.
-
-```{r figure 5, fig.cap = ""}
-knitr::include_graphics(here("analysis/09_spatiotemporal_models/temporal_panel_figure.png"))
-```
 

### 2022-11-27 analysis/08_spatiotemporal_autocorrelation/spatiotemporal_autocorrelation.R

@@ -265,9 +265,10 @@ model <- bam(
   data = df_for_bam
 )
 
-ggplot(df_spatial, aes(fill = sqrt(crude_rate_per_100k_py))) +
+ggplot(df_spatial, aes(fill = crude_rate_per_100k_py)) +
   geom_sf(size = 0) +
-  scale_fill_distiller(palette = 'Reds', direction = 1) +
+  # scale_fill_distiller(palette = 'Reds', direction = 1) +
+  scale_fill_viridis_c(option = 'turbo') +
   theme(legend.position = 'bottom')
 
 ggsave(here("analysis/08_spatiotemporal_autocorrelation/base_map.png"))
@@ -327,19 +328,21 @@ ggplot(df_spatial %>% sf::st_transform(crs = "EPSG:2163"), aes(fill = crude_rate
   scale_fill_viridis_c(
     option = 'turbo',
     direction = 1,
-    breaks = c(0, 100, 250, 500),
-    limits = c(0, 500),
-    oob = scales::squish
-  ) +
-  scale_fill_distiller(
-    palette = 'Reds',
-    breaks = c(0, 100, 250, 500),
-    direction = 1,
-    trans = scales::sqrt_trans(),
+    breaks = c(0, 100, 200, 300, 400, 500),
     limits = c(0, 500),
     oob = scales::squish,
     guide = my_triangle_colourbar()
+
   ) +
+  # scale_fill_distiller(
+  #   palette = 'Reds',
+  #   breaks = c(0, 100, 250, 500),
+  #   direction = 1,
+  #   trans = scales::sqrt_trans(),
+  #   limits = c(0, 500),
+  #   oob = scales::squish,
+  #   guide = my_triangle_colourbar()
+  # ) +
   labs(fill = stringr::str_wrap("Crude Mortality Rate per 100,000",10)) +
   theme_bw() +
   theme(legend.position = 'right')
@@ -348,20 +351,36 @@ ggsave("analysis/08_spatiotemporal_autocorrelation/crude_mortality_map_january20
 
 ggplot(df_spatial %>% sf::st_transform(crs = "EPSG:2163"), aes(fill = df_latlon$pred_rate)) +
   geom_sf(size = 0) +
-  scale_fill_distiller(
-    palette = 'Reds',
-    breaks = c(0, 100, 500, 1000),
+  scale_fill_viridis_c(
+    option = 'turbo',
     direction = 1,
-    trans = scales::sqrt_trans(),
+    breaks = c(0, 100, 200, 300, 400, 500),
[…]

### 2022-11-27 analysis/09_spatiotemporal_models/spatiotemporal_models.R

@@ -141,13 +141,13 @@ for (i in 1:length(year_vec)) {
   ggsave(
     filename = here(
       paste0("analysis/09_spatiotemporal_models/animation/",
-      stringr::str_pad(
-        i,
-        side = 'left',
-        width = 2,
-        pad = '0'
-      ),
-      ".png"
+             stringr::str_pad(
+               i,
+               side = 'left',
+               width = 2,
+               pad = '0'
+             ),
+             ".png"
       )),
     plot = plt,
     width = 10, height = 6)
@@ -160,30 +160,30 @@ system("convert -delay 40 -loop 0 *.png spatiotemporal_animation.gif \( +clone -
 # create temporal panel figure
 
 visualize_gam_map(model, year_vec[5], month_vec[5], fulltitle = F) +
-visualize_gam_map(model, year_vec[6], month_vec[6], fulltitle = F) +
-visualize_gam_map(model, year_vec[7], month_vec[7], fulltitle = F) +
-visualize_gam_map(model, year_vec[8], month_vec[8], fulltitle = F) +
+  visualize_gam_map(model, year_vec[6], month_vec[6], fulltitle = F) +
+  visualize_gam_map(model, year_vec[7], month_vec[7], fulltitle = F) +
+  visualize_gam_map(model, year_vec[8], month_vec[8], fulltitle = F) +
 
-visualize_gam_map(model, year_vec[9], month_vec[9], fulltitle = F) +
-visualize_gam_map(model, year_vec[10], month_vec[10], fulltitle = F) +
-visualize_gam_map(model, year_vec[11], month_vec[11], fulltitle = F) +
-visualize_gam_map(model, year_vec[12], month_vec[12], fulltitle = F) +
+  visualize_gam_map(model, year_vec[9], month_vec[9], fulltitle = F) +
+  visualize_gam_map(model, year_vec[10], month_vec[10], fulltitle = F) +
+  visualize_gam_map(model, year_vec[11], month_vec[11], fulltitle = F) +
+  visualize_gam_map(model, year_vec[12], month_vec[12], fulltitle = F) +
 
-visualize_gam_map(model, year_vec[13], month_vec[13], fulltitle = F) +
-visualize_gam_map(model, year_vec[14], month_vec[14], fulltitle = F) +
-visualize_gam_map(model, year_vec[15], month_vec[15], fulltitle = F) +
[…]

### 2022-11-27 manuscript/manuscript.rmd
updating; first draft done of discussion + conclusions

@@
   \noindent Abbreviations used: United States (US), Coronavirus Disease 2019 (COVID-19), Generalized
-  Additive Model (GAM), Centers for Disease Control and Prevention (CDC)
+  Additive Model (GAM), Centers for Disease Control and Prevention (CDC), Index of Concentration at the Extremes (ICE)
 bibliography: bibliography.bib
@@
 
-Prior literature has demonstrated that geography has played a significant role
-in the spread and impact of of COVID-19. Methods employed to highlight the role
-geography plays have included quantile regression [@sigler_socio-spatial_2021],
-Besag-York-Molli\'e mixed models [@whittle_ecological_2020], spatial cluster analysis
-[@sugg_mapping_2021], geographically weighted regression [@mollalo_gis-based_2020; @park_covid-19_2021], and others.
+Prior literature has demonstrated that the spread and impact of COVID-19 has
+varied geography over time reflecting dynamics in the timing of introduction and
+transmission events. Methods employed to highlight the geographic patterning in
+COVID-19 outcomes have included quantile regression
+[@sigler_socio-spatial_2021], Besag-York-Molli\'e mixed models
+[@whittle_ecological_2020], spatial cluster analysis [@sugg_mapping_2021],
+geographically weighted regression [@mollalo_gis-based_2020;
+@park_covid-19_2021], and others.
 
@@
   * Counts of COVID-19 deaths [@the_new_york_times_coronavirus_2021].
-  * Population size estimates for 2020 from the U.S. Census (Redistricting 
-  File Public-Law 94-171 Dataset) [@us_census_bureau_2020_2021].
+  * Population size estimates for 2020 from the U.S. Census [@us_census_bureau_2020_2021].
   * Median age, median household income, racial/ethnic composition, population
@@
 
-Traditional generalized linear models fit a regression model using the
-functional relation $g(\mu_i) = \mathbf X_i \beta$ where Examples of common link
-functions include the identity function for linear regression, $\log$ for
-Poisson regression, and the $\text{logit}$ function for logistic regression.
-
-In contrast, generalized additive models (GAMs) improve upon generalized linear models by
+Generalized additive models (GAMs) improve upon generalized linear models by
 allowing for the fitting of smooth functions that transform the $x$ variables.
 This is a convenient means to include nonlinear relationships between the
-outcome and predictor variables. Whereas a generalized linear model may have looked like 
+outcome and predictor variables. Whereas a generalized linear model may look like 
 $$ g(\mu_i) = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} ...$$ 
@@
 $$y_i \sim \text{NegativeBinomial}(u_i \theta_i, \phi),$$
-where $y_i$ are the counts observed, $u_i$ is the exposure, $\theta_i$ are
+where $y_i$ are the counts observed, $u_i$ is the "exposure", $\theta_i$ are
 the rates, and $\phi$ determines the amount of overdispersion. The rates
@@
 
-Besides age, which is the greatest risk factor for COVID-19 mortality, we have
+Besides age, which is a strong determinant of COVID-19 mortality risk, we have
 included the following variables as covariates of interest:
@@
 
-The GAM models are fit with the following structure using 
+The GAM models shown in Figure \@ref(fig:one-variable-over-time) are fit with the following structure using 
 the `gam` function from the `mgcv` package [@wood_mgcv_2022]:
@@
 variables of interest: median income, percent in poverty, the ICE for racialized
-economic segregation, political lean.  Whereas the other models treat median age
-as a variable to adjust for, the model presenting the effect of median age over
-time treats median age as the main covariate of interest including it as the 
-`te(covariate, date, d=c(1,1))` and dropping the `s(median_age)` term which is 
-redundant. Results of these models are summarized in \@ref(fig:one-variable-over-time)
-where the additional COVID-19 mortality associated with each covariate is visualized.
+economic segregation, political lean.  The model presenting median age treats
+median age as the main covariate of interest including it as the `te(covariate,
+date, d=c(1,1))`. Results of these models are summarized in Figure
+\@ref(fig:one-variable-over-time) where the additional COVID-19 mortality
+associated with each covariate is visualized.
 
@@
 
-Another non-spatial application of GAMs to describe the distribution of 
+A more complex non-spatial application of GAMs to describe the distribution of 
 COVID-19 mortality in the US over time is to consider the associations of 
@@
 that a one unit difference in one variable is comparable to a one unit difference
-in another variable.
+in another variable. This approach is represented in Figure \@ref(fig:two-variables-over-time).
 
@@
 A) May 2020, B) November 2020, C) May 2021, D) October 2021, E) February 2022, F) August 2022.
-An animated version is available online at [https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/readme.md](https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/readme.md)"}
+An animated version is available online at \\url{https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/readme.md}"}
 knitr::include_graphics(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/panel_figure.png"))
@@
 
+\newpage
 ## Application 2: Spatiotemporal Smoothing
@@
 do this, the GAM is constructed similarly as in Application 1 but with the
-formula specified as `deaths ~ te(latitude, longitude, time, d=c(2,1))` where
+smoothing term specified as `te(latitude, longitude, time, d=c(2,1))` where
 `d=c(2,1)` indicates that `latitude` and `longitude` share the same dimensions
-(i.e., both are spatial and in units of degrees) while the `time` data are in separate
-units.
+(i.e., both are spatial and in units of degrees) while the `time` data are in
+separate units. Note that results presented in Application 2 and 3 are based on
+county-level data from the contiguous US excluding Alaska and Hawaii.
 
@@
 mortality are noisy due to 72% of the land area in the US being classified as
-rural and low population density, the spatiotemporally smoothed GAM estimates
+rural and low population density [@health_resources__services_administration_defining_2022], the spatiotemporally smoothed GAM estimates
 use population weighting via the offset specified to estimate a latent surface
 that represents localized mortality rate averages in spatial coordinates and in
+time.
+
+Figure \@ref(fig:crude-vs-smoothed) shows the difference between crude mortality
+rates and spatially smoothed mortality rates in January 2022 to illustrate the
+level of noise present in raw rates and how spatially smoothing synthesizes
+local geographic patterns into interpretable trends that can be meaningfully
+interpreted as local area mortality risk levels pooling information across
+nearby county rates.
+
+Figure \@ref(fig:spatiotemporal-panel-figure) shows the results from a
+spatiotemporally smoothed model in select months to highlight how spatiotemporal
+smoothing can yield results that intuitively synthesize trends in space and
 time.
@@
 Panels show months selected to highlight the changing spatial patterns of COVID-19 mortality risk over time.
-An animated version is available online at []()"}
+An animated version is available online at \\url{https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/09_spatiotemporal_models/animation/readme.md}"}
 knitr::include_graphics(here("analysis/09_spatiotemporal_models/temporal_panel_figure.png"))
@@
 
+\newpage
 ## Application 3: Estimating Covariate Effects Adjusted for Spatiotemporal Autocorrelation
 
-### Evaluating Spatial and Temporal Autocorrelation
+The final application of GAMs presented here is to estimate the effects of
+covariates after adjusting for a spatiotemporally smoothed term as shown in
+Application 2. This has the interpretation of asking what are the changes to
+COVID-19 mortality rates associated with each covariate after taking into
+account local geographic patterns over time. In particular, this is relevant for
+understanding what county-level measures are associated with elevated COVID-19
+mortality even after adjusting for where COVID-19 rates were locally elevated
+or depressed due to variation in the timing of local introduction, transmission,
+and diffusion.
 
-```{r figure county neighboring graph, fig.cap = "A) The spatial neighbor graph for counties in the US.
-In this graph, county centroids are connected if the two counties are adjacent to one another (i.e., share a boundary).
-This and similar graphs that represent neighbor relationships or geographic proximity are often used in modeling contexts to create spatial weights that account for which regions help to explain a given region's observed data.", fig.show='hold', out.width = '50%', fig.env = "figure*"}
-desired_caption <- 
-  " B) Moran's scatter plot \\citep{anselin_simple_1996} depicts for each region (in our case, counties) on the x-axis the region rate and on the y-axis the average of the region's neighbors' rates.
-This diagnostic plot provides a simple means to assess to what degree neighboring regions share similar rates.
-In a setting where regions have high correlation, Moran's scatter plot will exhibit a high degree of positive correlation."
-knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/county_neighboring_relations.png")) 
-knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/morans_plot.png"))
+```{r neighboring-graph-morans-i, fig.cap = "The spatial autocorrelation of COVID-19 rates in January 2022.
+Panel A shows the graph of which US counties are connected by neighboring relations.
+Panel B shows a scatterplot depicting each US counties' COVID-19 rate (x-axis) compared to the average of its neighbors' COVID-19 mortality rates (y-axis).
+A regression line shows the association between counties' COVID-19 rates and their neighboring counties' COVID-19 rates.
+Dotted lines indicated the average for US county-level crude COVID-19 mortality rates (vertical) and for the average of each counties' neighboring counties' COVID-19 rates (horizontal)."}
+knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/county_network_and_morans_plot.png"))
 ```
 
-```{r figure 4, fig.cap = ""}
+```{r temporal-acf, fig.cap = "The median (black) and quantiles (shaded bands) of temporal autocorrelation of counties' COVID-19 mortality rates over time starting from April 2020.", out.width = '50%', fig.align='center'}
 knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/temporal_acf.png"))
@@
 
-## Regression Modeling Results
+To motivate the need for and application of spatiotemporal smoothing in
+assessing the associated changes in COVID-19 mortality with county-level
+covariates, Figure \@ref(fig:neighboring-graph-morans-i) presents the Moran's I
+diagnostic plot for January 2022 which summarizes the amount of autocorrelation
+between counties' COVID-19 mortality rates and a population-weighted average of
+each counties' neighboring counties' COVID-19 mortality rates. The observed
+autocorrelation indicates that there is an association between counties'
+COVID-19 rates and their neighboring counties' COVID-19 rates, implying that
+without taking this patterning into account, regression models that do not
+explicitly model the effect of spatial autocorrelation may be biased due to
+inappropriately assuming that county-level data are independent of one another. Second, 
+while the Moran's I plot demonstrates the spatial-correlation present in the data
+at a single time-point, the following Figure \@ref(fig:temporal-acf) shows the
+tendency of county COVID-19 observations to be autocorrelated between subsequent
+months and moreover there appears to be an oscillating trend at the 6-month and
+12-month marks suggestive of effects due to the dynamics associated with
+temporary infection induced immunity and seasonality.
 
-We used GAMs to describe the following:
-
-  * the changing associations between individual area-based measures
-  and COVID-19 mortality over time;
-  * the changing associations between bivariate area-based measures
-  and COVID-19 mortality over time; 
-  * associations with area-based measures and COVID-19 mortality after
-  controlling for spatio-temporal autocorrelation.
-
-Additionally, as sensitivity analyses, random forest models were fit to compare
-with the coefficient estimates from the GAMs.
-
-```{r figure 6, fig.cap = ""}
+```{r with-and-without-spatiotemporal-adjustment, fig.cap = "Estimates of increases in COVID-19 mortality associated with county-level covariates without (first) and with (second) adjustment for spatiotemporal autocorrelation.
+Panels A and B show COVID-19 mortality increases associated with varying levels of median age over time; C and D show associations with log population density; E and F median income; G and H for proportion in poverty; I and J the ICE for racialized economic segregation; K and L political lean.", out.width = '85%', fig.align='center'}
 knitr::include_graphics(here("analysis/09_spatiotemporal_models/panel_figure.png"))
@@
 
+Figure \@ref(fig:with-and-without-spatiotemporal-adjustment) shows the effects associated
+with each covariate over time with and without adjustment for spatiotemporal trends. 
+The models for population density and median income show particularly marked changes
+with the spike at high population density and high median incomes disappearing 
+after adjusting for a spatiotemporally smoothed term. This likely reflects the
+early emergence and surge of COVID-19 in the greater New York City area as being
+predominantly a feature of the combined effect of human geography and infectious
+disease dynamics where the earliest introductions of COVID-19 were made to dense
+social networks in the US. Other effects remain qualitatively changed very little,
+suggesting that the trends in COVID-19 mortality associated with these county-level
+measures remains durable when taking into account spatiotemporal autocorrelation.
+
+\newpage
 # Discussion
 
-## Limitations
+This paper presents three applications of GAMs to county-level COVID-19
+mortality data from the US during March 2020 to August 2022. This paper
+addresses an important need, which is to assess what measures of the
+socioeconomic and geopolitical landscape are associated with increases in
+COVID-19 mortality in a way that takes into account how infectious disease
+dynamics are locally correlated in space and time. However, the approaches
+outlined in this paper are not without several limitations which include (at
+least) the following: the potential for cross-level confounding, lack of
+individual level data, an inability to include all possible metrics, further
+improvements to the methods used may be warranted, and we are unable to make
+claims about the causal relationships between the variables investigated.
 
-  * Ecological fallacy 
-  * Age effects
-  * Fixed population denominators -> real-time denominator estimates
-  * Vaccination and mobility effects
-  * Equal Smoothing Across All Areas
-  * Soap Film Smoothers
+In analyses that use area-level aggregated data, the ecologic fallacy refers to
+the possibility that associations between area-level measures and outcomes may
+not reflect the associations that would be observed if analyzing individual
+level data on the same measures and outcomes as a result of cross-level
+confounding [@greenland_ecologic_2001]. It's important to note that
+some of the variables are included to contextualize the places in which people
+live, reside, and work and area-based social metrics should be included in
+individual-level analyses so as to accurately capture the effects of both
+individual-level risk factors and risk factors that stem from the contexts in
+which people exist [@testa_public_2022].
+
+One particular improvement that could be made in future work is to address age
+adjustment by using direct or indirect age standardization [@anderson_age_1998] or 
+to create age-group stratified models. In the applications presented here,
+county-level median age is adjusted for in order to control for age effects.
+
+Additional variables like vaccination rates, mobility data, further variables
+that relate to risk factors for COVID-19, and other COVID-19 related outcomes
+like case trends and hospitalizations may be worthwhile to investigate under a
+similar framework. The emphasis on trends in COVID-19 mortality in the
+applications in this paper was done to prioritize presenting data that are less
+subject to the reporting inconsistencies in COVID-19 cases and
+hospitalizations [@galaitsi_challenges_2021].
+
+Further improvements to the methods employed here could include use of soap film
+smoothing splines which are more appropriate in settings where non-convex
+geographic boundaries are common (as in with peninsulas and bays)
+[@wood_soap_2008]. Taking a critical eye to the assumptions employed reveals yet
+another area of potential improvement, which is the nature of how the complexity of the 
+smoothing splines employed are penalized in the GAM framework. It is conceivable that
+geographic changes in COVID-19 mortality rates are not equally smooth across 
+all regions of the US, potentially warranting rapid changes in some areas 
+and smooth changes in other areas. The Bayesian wombling literature provides
+the means to identify areas of rapid change in latent surface models and may pose
+a fruitful direction of exploration for future efforts to understand trends
+in COVID-19 mortality rates in the US [@gelfand_bayesian_2015].
 
@@
 
+This work introduces generalized additive models (GAMs) applied to COVID-19
+mortality data to document the disparities and inequities in which US counties
+at what times suffered the worst mortality outcomes. The effort to document
+inequities and explain the driving mechanisms behind them is crucial in building
+up the evidence necessary to enact policy changes that can mitigate unjust and
+preventable harm in the future. To this end, it is critical that efforts to
+model infectious disease dynamics take into account how disease transmission and
+outcomes are spatiotemporally patterned, violating any misplaced assumptions
+that area-level disease rates can be treated as independent observations. By
+presenting novel applications of GAMs to trends in COVID-19 mortality, this
+paper illustrates new ways of understanding and visualizing the disproportionate
+burden some communities have suffered incorporating complex patterns of
+interaction across socioeconomic and geopolitical variables and spatiotemporal
+trends.
+
+\newpage
 # References
-
-<!-- 1 million covid deaths: 
-https://coronavirus.jhu.edu/from-our-experts/u-s-officially-surpasses-1-million-covid-19-deaths -->
-
-<!-- nearly 225,000 Americans have died since the start of 2022
-https://abcnews.go.com/Health/225000-americans-died-covid-19-start-2022-viral/story?id=90339579 -->
-
-<!-- 
-vaccine timeline
-https://www.mayoclinic.org/coronavirus-covid-19/history-disease-outbreaks-vaccine-timeline/covid-19
--->
-
-<!--
-less than 4% of americans had gotten the booster 1 month into availability
-https://www.nbcnews.com/health/health-news/updated-covid-booster-shots-doses-administered-cdc-rcna48960
--->
-
-<!-- 2020 ACS data are experimental 
-https://www.census.gov/newsroom/press-releases/2021/experimental-2020-acs-1-year-data.html -->
-
-<!--
-Spatial-temporal generalized additive model for modeling COVID-19 mortality risk in Toronto, Canada
-https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8257405/ -->
-
-<!-- 
-Using GAM functions and Markov-Switching models in an evaluation framework to assess countries’ performance in controlling the COVID-19 pandemic
-https://bmcpublichealth.biomedcentral.com/articles/10.1186/s12889-021-11891-6 -->
-
-<!--
-Statistical modelling of COVID-19 data: Putting generalized additive models to work
-https://journals.sagepub.com/doi/full/10.1177/1471082X221124628
--->
-
-<!--
-Inference following multiple imputation for generalized additive models: an investigation of the median p-value rule with applications to the Pulmonary Hypertension Association Registry and Colorado COVID-19 hospitalization data
-https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-022-01613-w
--->
-
-<!--
-Generalized Additive Models to Capture the Death Rates in Canada COVID-19
-https://link.springer.com/chapter/10.1007/978-3-030-85053-1_7 -->
-
-<!--
-Contribution of Two-Dose Vaccination Toward the Reduction of COVID-19 Cases, ICU Hospitalizations and Deaths in Chile Assessed Through Explanatory Generalized Additive Models for Location, Scale, and Shape
-https://www.frontiersin.org/articles/10.3389/fpubh.2022.815036/full -->
-
-<!-- 
-Combining graph neural networks and spatio-temporal disease models to improve the prediction of weekly COVID-19 cases in Germany
-https://www.nature.com/articles/s41598-022-07757-5 -->
-
-<!-- 
-Association between mobility patterns and COVID-19 transmission in the USA: a mathematical modelling study
-https://www.sciencedirect.com/science/article/pii/S1473309920305533 -->
-
-<!-- 
-Examining spatiotemporal evolution of racial/ethnic disparities in human mobility and COVID-19 health outcomes: Evidence from the contiguous United States
-https://www.sciencedirect.com/science/article/pii/S2210670721007721 -->
-
-<!--
-Social, economic, and environmental factors influencing the basic reproduction number of COVID-19 across countries
-https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0252373 -->
-
-<!-- 
-https://asbates.rbind.io/2019/05/03/gams-for-time-series/
--->
-
-<!--
-https://environmentalcomputing.net/statistics/gams/
--->
-
-<!--
-https://petolau.github.io/Analyzing-double-seasonal-time-series-with-GAM-in-R/
--->

### 2022-11-27 R/create_analytic_dataset.R

@@ -1,11 +1,6 @@
 
 #' Create Analytic Dataset
 #'
-#' @examples
-#' library(ggdist)
-#' df <- create_analytic_dataset()
-#'
-#'
 create_analytic_dataset <- function() {
 
   # load deaths and 2020 population size data

### 2022-11-27 R/load_covid19_deaths.R

@@ -1,25 +1,4 @@
 
-#' Load State COVID-19 Deaths
-load_state_covid19_deaths <- function(clean = TRUE) {
-
-  # load data
-  df <- readr::read_delim(file = system.file(
-      "cdc_wonder/States - Provisional Mortality Statistics, 2018 through Last Month.txt",
-      package = 'covid.gradient.estimation'
-    )
-  )
-
-  # optional cleaning steps
-  if (clean) {
-    df %<>% janitor::clean_names()
-    df %<>% filter(is.na(notes))
-    df %<>% select(-notes, -population, -crude_rate)
-  }
-
-  return(df)
-}
-
-
 #' Load County COVID-19 Deaths
 load_county_covid19_deaths <- function(clean = TRUE) {
 

### 2022-11-27 analysis/02_plot_county_trends/plot_county_trends.R

@@ -55,12 +55,13 @@ plt_orig <- df %>%
   stat_lineribbon(
     mapping = aes(
       group = division,
-      fill = division,
+      # fill = division,
       color = division,
       fill_ramp = stat(level)
     ),
     alpha = .75,
-    color = 'grey20'
+    color = 'grey20',
+    fill = 'red'
   ) +
   geom_line(
     mapping = aes(size = ' ')
@@ -107,7 +108,8 @@ wtd_quantile_df <-
     ci95_low = Hmisc::wtd.quantile(x = crude_rate_per_100k_py, weights = popsize, probs = .025, na.rm=T)
   )
 
-palette <- scales::hue_pal()(9) # RColorBrewer::brewer.pal(n = 9, name = 'Set1')
+# palette <- scales::hue_pal()(9) # RColorBrewer::brewer.pal(n = 9, name = 'Set1')
+palette <- rep('red', 9)
 palette_named <- setNames(palette, unique(wtd_quantile_df$division))
 
 plt <-
@@ -120,20 +122,20 @@ plt <-
   geom_ribbon(
     mapping = aes(ymax = ci95_high, ymin = ci95_low, group = NULL, y = NULL),
     size = 0,
-    alpha = 0.7,
-    fill = colorspace::lighten(palette_named[wtd_quantile_df$division], .7)
+    alpha = 0.6,
+    fill = "#ffe3d8" # colorspace::lighten(palette_named[wtd_quantile_df$division], .8)
   ) +
   geom_ribbon(
     mapping = aes(ymax = ci80_high, ymin = ci80_low, group = NULL, y = NULL),
     size = 0,
-    alpha = 0.7,
-    fill = colorspace::lighten(palette_named[wtd_quantile_df$division], .3)
+    alpha = 0.8,
+    fill = "#ffa78f" # colorspace::lighten(palette_named[wtd_quantile_df$division], .4)
   ) +
   geom_ribbon(
     mapping = aes(ymax = ci50_high, ymin = ci50_low, group = NULL, y = NULL),
     size = 0,
-    alpha = 0.8,
-    fill = palette_named[wtd_quantile_df$division]
+    alpha = 1,
+    fill = "red" # palette_named[wtd_quantile_df$division]
   ) +
   geom_line(
     mapping = aes(y = median, group = NULL),

### 2022-11-27 analysis/08_spatiotemporal_autocorrelation/spatiotemporal_autocorrelation.R

@@ -149,8 +149,8 @@ ggsave("analysis/08_spatiotemporal_autocorrelation/seasonality_3.png", width=8, 
 county_map <- urbnmapr::get_urbn_map('counties', sf=TRUE)
 state_map <- urbnmapr::get_urbn_map('states', sf=TRUE)
 
-# filter for just values from january 2021
-df_jan2021 <- county_map %>% left_join(df %>% filter(date == "2021-01-01"), by = c('county_fips' = 'geoid'))
+# filter for just values from january 2022
+df_jan2022 <- county_map %>% left_join(df %>% filter(date == "2022-01-01"), by = c('county_fips' = 'geoid'))
 
 # plot the county map
 ggplot(county_map) +
@@ -177,12 +177,12 @@ ggplot(county_map) +
 ggsave(here("analysis/08_spatiotemporal_autocorrelation/county_neighboring_relations.png"), width = 10, height = 6)
 
 # use january 2021 covid mortality rates with no missing data
-df_jan2021_no_NAs <- df_jan2021 %>% filter(! is.na(crude_rate_per_100k_py))
-nb_q <- spdep::poly2nb(df_jan2021_no_NAs)
+df_jan2022_no_NAs <- df_jan2022 %>% filter(! is.na(crude_rate_per_100k_py))
+nb_q <- spdep::poly2nb(df_jan2022_no_NAs)
 
 # use moran's I approach to assess spatial autocorrelation
-mp <- moran.plot(x = df_jan2021_no_NAs$crude_rate_per_100k_py, listw = nb2listw(nb_q, zero.policy = TRUE), zero.policy = TRUE,
-                 labels=as.character(paste0(df_jan2021_no_NAs$county, ", ", df_jan2021_no_NAs$state_abbv)), pch=19)
+mp <- moran.plot(x = df_jan2022_no_NAs$crude_rate_per_100k_py, listw = nb2listw(nb_q, zero.policy = TRUE), zero.policy = TRUE,
+                 labels=as.character(paste0(df_jan2022_no_NAs$county, ", ", df_jan2022_no_NAs$state_abbv)), pch=19)
 
 mp %<>% mutate(
   quadrant = case_when(
@@ -203,7 +203,7 @@ ggplot(mp, aes(x=x, y=wx, color = quadrant)) + geom_point(alpha = .3) +
   geom_hline(yintercept=mean(mp$wx), lty=2) +
[…]

### 2022-11-27 analysis/09_spatiotemporal_models/spatiotemporal_models.R

@@ -45,7 +45,7 @@ spatiotemporal_df %<>% filter(state %in% setdiff(c(state.name, 'District of Colu
 
 # fit bam (e.g., GAM optimized for very large datasets)
 model <- bam(
-  formula = deaths ~ te(latitude, longitude, date_int, d = c(2, 1)),
+  formula = deaths ~ s(median_age) + te(latitude, longitude, date_int, d = c(2, 1)),
   offset = log(popsize/1e5/12),
   family = nb(),
   data = spatiotemporal_df
@@ -192,7 +192,7 @@ ggsave(here("analysis/09_spatiotemporal_models/temporal_panel_figure.png"), widt
 # ICEraceinc without spatiotemporal adjustment
 
 model <- bam(
-  formula = deaths ~ # te(latitude, longitude, date_int, d = c(2, 1)) +
+  formula = deaths ~ s(median_age) +
     te(ICEraceinc, date_int, d = c(1,1)),
   offset = log(popsize/1e5/12),
   family = nb(),
@@ -203,6 +203,7 @@ png(filename = here("analysis/09_spatiotemporal_models/ICEraceinc_and_time_no_sp
 vis.gam(
   model,
   view = c('date_int', 'ICEraceinc'),
+  cond = list(age = 38.8),
   n.grid = 50,
   theta = 35,
   phi = 32,
@@ -222,7 +223,7 @@ dev.off()
 # now consider how ICEraceinc matters after taking into account spatiotemporal autocorrelation
 
 model <- bam(
-  formula = deaths ~ te(latitude, longitude, date_int, d = c(2, 1)) +
+  formula = deaths ~ s(median_age) + te(latitude, longitude, date_int, d = c(2, 1)) +
     te(ICEraceinc, date_int, d = c(1,1)),
   offset = log(popsize/1e5/12),
   family = nb(),
@@ -233,6 +234,7 @@ png(filename = here("analysis/09_spatiotemporal_models/ICEraceinc_and_time.png")
 vis.gam(
   model,
   view = c('date_int', 'ICEraceinc'),
+  cond = list(age = 38.8),
   n.grid = 50,
   theta = 35,
   phi = 32,
@@ -251,7 +253,7 @@ dev.off()
 # median age without adjusting for spatiotemporal autocorrelation
 
 model <- bam(
-  formula = deaths ~ # te(latitude, longitude, date_int, d = c(2, 1)) +
+  formula = deaths ~ s(age) +
[…]

### 2022-12-01 manuscript/manuscript.rmd
A is for the strictly parametric terms
The definition of  $\mathbf{A_i}$ given in the manuscript is incorrect. **A** is the model matrix for the strictly parametric terms (i.e terms one might include in a LM or GLM). In general non-parametric is a bit of a misnomer, as all terms in the GAM as you (and Simon) define it are parametric once we've formed the basis expansion(s) for any smooth terms in the model. The distinction between what is and is not parametric gets complicated as models of this form:

```r
m <- gam(y ~ x1 * f1 + bs(x2) + s(x3) + te(x4, x5), data = df, method = "REML")
```

are

@@
 where $\mu_i \equiv \mathbb E(Y_i)$, $\mathbf A_i$ is a row of the model matrix 
-for any strictly non-parametric model components, $\mathbf \theta$ is the corresponding
+for any strictly parametric model components, $\mathbf \theta$ is the corresponding
 parameter vector, and the $f_j$ are smooth functions of the covariates $x_k$ [@wood_generalized_2017].

### 2022-12-01 manuscript/manuscript.rmd
prepare for arxiv submission

@@
     number_sections: false
+    keep_tex: true
 abstract: |
@@
   space to address statistical instability due to small
-  population counts or stochasticity can be summarized by a smooth, dynamic latent
-  surface describing the mortality risk associated with geographic locations over
+  population counts or stochasticity resulting in a smooth, dynamic latent
+  risk surface summarizing the mortality risk associated with geographic locations over
   time. Third, estimation of COVID-19 mortality associations with county-level covariates
-  conditional on smooth spatiotemporal variation allows for more rigorous consideration
+  conditional on a smooth spatiotemporal risk surface allows for more rigorous consideration
   of how socio-environmental contexts and policies may have impacted COVID-19 mortality.
@@
   \noindent Abbreviations used: United States (US), Coronavirus Disease 2019 (COVID-19), Generalized
-  Additive Model (GAM), Centers for Disease Control and Prevention (CDC), Index of Concentration at the Extremes (ICE)
+  Additive Model (GAM), Centers for Disease Control and Prevention (CDC), Index of Concentration at the Extremes (ICE), Confidence Interval (CI)
 bibliography: bibliography.bib
@@
 ---
-
-<!--
-TODO:
-
-Throughout/Overarching: 
-  - Do I need to update all the models to separately control for age? Probably.
-  - Do I want to have a supplementary materials section?
-  - Talk about ecosocial theory (?) (Who benefits from harm done to whom?)
-
-Abstract:
-  - Add link to GitHub repository
-
-Intro: 
-  - add geographic background, like where the first cases were reported, 
-  how COVID exploded in NYC, then the surges in the Midwest and South
-  - add a primer on the relevance of spatiotemporal modeling
-
-Methods: 
-  - add intro to ICEraceinc (methods)
-  - describe negative binomial models
-
-Results:
-  - add figure for temporal two-way interaction effects
-  - present rate ratios (results)
-  - present model comparisons (results, F-tests, AIC)
-
-Discussion:
-  - talk about how COVID has not been a win for health equity (discussion)
-  - talk about limitations to being causal
-  - talk about future directions
-
-Conclusion:
-  - summarize
--->
 
@@
 
-Having passed over 1 million COVID-19 deaths in the United States earlier this
-year [@donovan_us_2022], and facing uncertain prospects for the third COVID-19
-winter looming even as new iterations on the COVID-19 vaccines become available,
-it remains critical that inequities in COVID-19 outcomes are documented and
+Having passed over 1 million COVID-19 deaths in the United States in 2022
+[@donovan_us_2022], and facing uncertain prospects for the third COVID-19 winter
+looming even as new iterations on the COVID-19 vaccines become available, it
+remains critical that inequities in COVID-19 outcomes are documented and
 analyzed to reckon with the unjust and unfair burden of preventable illness.
@@
 [@noauthor_united_2022].  The new bivalent vaccines released at the end of
-August 2022 contain mRNA sequences from both the original strain as well as
-the recently emergent BA.4 and BA.5 lineages in an effort to make the nation's immunity more
-up-to-date and robust against the myriad of phylogenetic directions the COVID-19
-virus is evolving to explore [@office_of_the_commissioner_coronavirus_2022].
-Despite the updated bivalent boosters representing a significant step forward in
-prevention strategy, less than 4% of eligible Americans had taken the booster in
-the first month after it became available [@bendix_less_2022; @lambert_most_2022]. 
-As such, and with an enduring history of inequities in health care access 
-in the US [@ortega_ending_2021; @blendon_inequities_2002; @bailey_how_2021;
+August 2022 contain mRNA sequences from both the original strain as well as the
+recently emergent BA.4 and BA.5 lineages in an effort to make the nation's
+immunity more up-to-date and robust against the myriad of phylogenetic
+directions the COVID-19 virus is evolving to explore
+[@office_of_the_commissioner_coronavirus_2022]. Despite the updated bivalent
+boosters representing a significant step forward in prevention strategy, less
+than 4% of eligible Americans had taken the booster in the first month after it
+became available [@bendix_less_2022; @lambert_most_2022]. As such, and with an
+enduring history of inequities in health care access in the US
+[@ortega_ending_2021; @blendon_inequities_2002; @bailey_how_2021;
 @okonkwo_covid-19_2021; @rapp_statelevel_2022;
@@
 For each division, the median trendline and quantile ranges are shown weighted by county population size.", out.width = "6in", fig.align = 'center', fig.env = "figure*"}
-knitr::include_graphics(here("analysis/02_plot_county_trends/county_trends_by_division.png"))
+# knitr::include_graphics(here("analysis/02_plot_county_trends/county_trends_by_division.png"))
+knitr::include_graphics(here("manuscript/figures/county_trends_by_division.png"))
 ```
@@
 The figure summarizes each division's median mortality rates
-weighted by county population size. Quantile ranges are included to illustrate
-the range of mortality rates observed, again weighted by county population size.
+weighted by county population size. 
 Notably, the mortality associated
-with the early surge of cases starting in NYC and spreading through New York,
+with the early surge of cases starting in New York City and spreading through New York,
 New Jersey, and Massachusetts is visible in the Middle Atlantic and New England
@@
 during March 2020 [@thompson_covid-19_2020] followed by subsequent waves of cases in the
-South and Midwest [@shumaker_covid-19_2020; @scott_these_2020; @glenza_covid_2020]. As Park et al. 
-stated summarized trends in the US from March 2020 to May 2021, "hot spots have shifted from densely populated cities and the states with a high percentage of socially vulnerable individuals to the states with relatively relaxed social distancing requirements, and then to the states with low vaccination rates" [-@park_covid-19_2021].
+South and Midwest [@shumaker_covid-19_2020; @scott_these_2020; @glenza_covid_2020]. As Park et al.\ 
+summarized the trends in the US from March 2020 to May 2021, "hot spots have shifted from densely populated cities and the states with a high percentage of socially vulnerable individuals to the states with relatively relaxed social distancing requirements, and then to the states with low vaccination rates" [-@park_covid-19_2021].
 
@@
 that COVID-19 threatened to exacerbate existing disparities
-[@kim_covid-19_2020]. Kim, Marrast, and Conigliaro noted three structural
-barriers in COVID-19 prevention and care: 1) originally requiring New York
+[@kim_covid-19_2020]. Kim, Marrast, and Conigliaro noted at least three structural
+barriers in COVID-19 prevention and care: 1) originally requiring
 residents to have a doctor's prescription for a COVID-19 test reduced the
@@
 primary care provider; 2) drive-through testing made testing disadvantaged those
-who rely on public transportation; 3) quarantining at home while waiting the
+who rely on public transportation; and 3) quarantining at home while waiting the
 7-10 days originally required for COVID-19 test results to come back posed an
@@
 yet more reasons why COVID-19 threatened to worsen an already inequitable
-healthcare landscape in the US: those who reside in prisons and jails,
+healthcare landscape in the US: in particular, those who reside in prisons and jails,
 immigrants and undocumented people, people with disabilities, and people
 experiencing homeless all face additional challenges in seeking and getting the
-healthcare they deserve [@okonkwo_covid-19_2021]. Despite stay-at-home orders
-that were prevalent in many states [@moreland_timing_2020], workers functioning in 
-capacities essential to the functioning of critical infrastructure operations (later
-termed "essential workers") were exposed to heightened risk of COVID-19 transmission 
-[@national_conference_of_state_legislatures_covid-19_2021; @wei_risk_2022; @national_bureau_of_economic_research_measuring_2021; @the_lancet_plight_2020; @hanage_covid-19_2020].
-It's not just a hypothetical that those with more structurally enfranchised
-privileges have been more able to mitigate their risk of negative health
-outcomes associated with COVID-19 — during February 1 to April 1, 2020, New York
-City residents who lived in more affluent neighborhoods were more likely to have
-left the city, while New York City residents from more Black and Hispanic
+healthcare they deserve [@okonkwo_covid-19_2021]. Even though stay-at-home orders
+designed to mitigate spread that were prevalent in many states
+[@moreland_timing_2020], workers functioning in capacities essential to the
+functioning of critical infrastructure operations (later termed "essential
+workers") were exposed to heightened risk of COVID-19 transmission
+[@national_conference_of_state_legislatures_covid-19_2021; @wei_risk_2022;
+@national_bureau_of_economic_research_measuring_2021; @the_lancet_plight_2020;
+@hanage_covid-19_2020]. It's clear that those with more
+structurally enfranchised privileges have been more able to mitigate their risk
+of negative health outcomes associated with COVID-19 — during February 1 to
+April 1, 2020, New York City residents who lived in more affluent neighborhoods
+were more likely to have left the city, while New York City residents from more
+Black and Hispanic
 neighborhoods were more likely to continue working [@coven_disparities_2020].
@@
 Generalized additive models (GAMs) improve upon generalized linear models by
-allowing for the fitting of smooth functions that transform the $x$ variables.
-This is a convenient means to include nonlinear relationships between the
+allowing for the fitting of smooth functions that transform right-hand-side variables.
+This is a convenient means to account for nonlinear relationships between the
 outcome and predictor variables. Whereas a generalized linear model may look like 
@@
 $$ g(\mu_i) = \mathbf A_i \mathbf \theta + f_1(x_{i1}) + f_2(x_{i2}) + f_3(x_{i3}, x_{i4}) + \dots$$
-where $\mu_i \equiv \mathbb E(Y_i)$, $\mathbf A_i$ is a row of the model matrix 
-for any strictly parametric model components, $\mathbf \theta$ is the corresponding
+where the expected value of the outcome is given as $\mu_i \equiv \mathbb E(Y_i)$, 
+$\mathbf A_i$ is a row of the model matrix 
+for any strictly non-parametric model components, $\mathbf \theta$ is the corresponding
 parameter vector, and the $f_j$ are smooth functions of the covariates $x_k$ [@wood_generalized_2017].
@@
 
-knitr::include_graphics(here("analysis/10_supplementary_gam_figures/gam_demo_figure.png"))
+# knitr::include_graphics(here("analysis/10_supplementary_gam_figures/gam_demo_figure.png"))
+knitr::include_graphics(here("manuscript/figures/gam_demo_figure.png"))
 ```
@@
 
-While a variety of varying kinds of spline bases exist (cubic splines, B- and
+While a variety of spline-based approaches exist (cubic splines, B- and
 P-splines, thin-plate splines, etc.), our particular application setting
@@
 
-Besides age, which is a strong determinant of COVID-19 mortality risk, we have
-included the following variables as covariates of interest:
+The following variables have been included as covariates of interest:
 
+  * Median Age
   * Population Density per Square Mile
@@
 privilege of interest.  In this case, data on racialized economic segregation
-are used from the US Census American Community Survey on White households
+are used from the US Census American Community Survey on white households
 earning more than $100,000 a year (the most privileged group) or households of
@@
 coefficient which is one of the most popular methods for summarizing area-level
-rates of inequities, the ICE the major advantage that it is more suitable for
+rates of inequities, the ICE has the advantage that it is suitable for
 describing inequities at smaller area levels [@krieger_public_2016]. While the
@@
 
-The formula used puts a one-dimensional smoothing spline to represent a
+The formula used puts a one-dimensional smoothing spline on median age to represent a
 nonlinear age-effect and a two-dimensional tensor-product smooth on the
@@
 median age as the main covariate of interest including it as the `te(covariate,
-date, d=c(1,1))`. Results of these models are summarized in Figure
+date, d=c(1,1))` and dropping the `s(median_age)` term which otherwise becomes redundant. 
+Results of these models are summarized in Figure
 \@ref(fig:one-variable-over-time) where the additional COVID-19 mortality
@@
 In panels B-F the effect of age is marginalized out using the median age in the US, 38.8 (ref:medianage).", fig.align = 'center', out.height = '75%'}
-knitr::include_graphics(here("analysis/04_one_variable_over_time/panel_figure.png"))
+# knitr::include_graphics(here("analysis/04_one_variable_over_time/panel_figure.png"))
+knitr::include_graphics(here("manuscript/figures/nonspatial_panel_figure.png"))
 ```
@@
 An animated version is available online at \\url{https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/readme.md}"}
-knitr::include_graphics(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/panel_figure.png"))
+# knitr::include_graphics(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/panel_figure.png"))
+knitr::include_graphics(here("manuscript/figures/ICEraceinc_age_panel_figure.png"))
 ```
@@
 level of noise present in raw rates and how spatially smoothing synthesizes
-local geographic patterns into interpretable trends that can be meaningfully
-interpreted as local area mortality risk levels pooling information across
+local geographic patterns into trends that can be meaningfully
+interpreted as local area mortality risk levels with information information pooled across
 nearby county rates.
@@
 spatiotemporally smoothed model in select months to highlight how spatiotemporal
-smoothing can yield results that intuitively synthesize trends in space and
+smoothing can yield results that synthesize trends in space and
 time.
@@
 B) Smoothed mortality rates from a GAM applied to mortality rates from January 2022."}
-knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/side_by_side_crude_and_smoothed.png"))
+# knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/side_by_side_crude_and_smoothed.png"))
+knitr::include_graphics(here("manuscript/figures/side_by_side_crude_and_smoothed.png"))
 ```
@@
 An animated version is available online at \\url{https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/09_spatiotemporal_models/animation/readme.md}"}
-knitr::include_graphics(here("analysis/09_spatiotemporal_models/temporal_panel_figure.png"))
+# knitr::include_graphics(here("analysis/09_spatiotemporal_models/temporal_panel_figure.png"))
+knitr::include_graphics(here("manuscript/figures/temporal_panel_figure.png"))
 ```
@@
 The final application of GAMs presented here is to estimate the effects of
-covariates after adjusting for a spatiotemporally smoothed term as shown in
-Application 2. This has the interpretation of asking what are the changes to
+covariates after adjusting for a spatiotemporally smoothed latent risk surface. 
+This has the interpretation of asking what are the changes to
 COVID-19 mortality rates associated with each covariate after taking into
-account local geographic patterns over time. In particular, this is relevant for
-understanding what county-level measures are associated with elevated COVID-19
-mortality even after adjusting for where COVID-19 rates were locally elevated
-or depressed due to variation in the timing of local introduction, transmission,
-and diffusion.
+account regional geographic patterns in COVID-19 mortality over time. In
+particular, this is relevant for understanding what county-level measures are
+associated with elevated COVID-19 mortality even after adjusting for where
+COVID-19 rates were locally elevated or depressed due to variation in the timing
+of local introduction, transmission, and diffusion.
 
 ```{r neighboring-graph-morans-i, fig.cap = "The spatial autocorrelation of COVID-19 rates in January 2022.
-Panel A shows the graph of which US counties are connected by neighboring relations.
-Panel B shows a scatterplot depicting each US counties' COVID-19 rate (x-axis) compared to the average of its neighbors' COVID-19 mortality rates (y-axis).
-A regression line shows the association between counties' COVID-19 rates and their neighboring counties' COVID-19 rates.
-Dotted lines indicated the average for US county-level crude COVID-19 mortality rates (vertical) and for the average of each counties' neighboring counties' COVID-19 rates (horizontal)."}
-knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/county_network_and_morans_plot.png"))
+Panel A shows the graph of US counties connected according to which are neighbors.
+Panel B shows a scatterplot depicting each US counties' COVID-19 mortality rate (x-axis) compared to the average of its neighbors' COVID-19 mortality rates (y-axis).
+A regression line shows the association between counties' COVID-19 rates and their neighboring counties' COVID-19 mortality rates.
+Dotted lines indicated the average for US county-level crude COVID-19 mortality rates (vertical) and for the average of each counties' neighboring counties' COVID-19 mortality rates (horizontal)."}
+# knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/county_network_and_morans_plot.png"))
+knitr::include_graphics(here("manuscript/figures/county_network_and_morans_plot.png"))
 ```
 
-```{r temporal-acf, fig.cap = "The median (black) and quantiles (shaded bands) of temporal autocorrelation of counties' COVID-19 mortality rates over time starting from April 2020.", out.width = '50%', fig.align='center'}
-knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/temporal_acf.png"))
+```{r temporal-acf, fig.cap = "The median (black) and quantiles (shaded bands) of temporal autocorrelation of counties' COVID-19 mortality rates over time during April 2020 — August 2022.", out.width = '50%', fig.align='center'}
+# knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/temporal_acf.png"))
+knitr::include_graphics(here("manuscript/figures/temporal_acf.png"))
 ```
@@
 diagnostic plot for January 2022 which summarizes the amount of autocorrelation
-between counties' COVID-19 mortality rates and a population-weighted average of
+between counties' COVID-19 mortality rates and the average of
 each counties' neighboring counties' COVID-19 mortality rates. The observed
 autocorrelation indicates that there is an association between counties'
-COVID-19 rates and their neighboring counties' COVID-19 rates, implying that
+COVID-19 rates and their neighboring counties' COVID-19 mortality rates, implying that
 without taking this patterning into account, regression models that do not
@@
 tendency of county COVID-19 observations to be autocorrelated between subsequent
-months and moreover there appears to be an oscillating trend at the 6-month and
-12-month marks suggestive of effects due to the dynamics associated with
-temporary infection induced immunity and seasonality.
+months.
 
@@
 Panels A and B show COVID-19 mortality increases associated with varying levels of median age over time; C and D show associations with log population density; E and F median income; G and H for proportion in poverty; I and J the ICE for racialized economic segregation; K and L political lean.", out.width = '85%', fig.align='center'}
-knitr::include_graphics(here("analysis/09_spatiotemporal_models/panel_figure.png"))
+# knitr::include_graphics(here("analysis/09_spatiotemporal_models/panel_figure.png"))
+knitr::include_graphics(here("manuscript/figures/spatiotemporal_panel_figure.png"))
 ```
 
-Figure \@ref(fig:with-and-without-spatiotemporal-adjustment) shows the effects associated
-with each covariate over time with and without adjustment for spatiotemporal trends. 
-The models for population density and median income show particularly marked changes
-with the spike at high population density and high median incomes disappearing 
-after adjusting for a spatiotemporally smoothed term. This likely reflects the
-early emergence and surge of COVID-19 in the greater New York City area as being
-predominantly a feature of the combined effect of human geography and infectious
-disease dynamics where the earliest introductions of COVID-19 were made to dense
-social networks in the US. Other effects remain qualitatively changed very little,
-suggesting that the trends in COVID-19 mortality associated with these county-level
-measures remains durable when taking into account spatiotemporal autocorrelation.
+Figure \@ref(fig:with-and-without-spatiotemporal-adjustment) shows the
+additional COVID-19 mortality associated with each covariate over time without and
+with adjustment for spatiotemporal trends. The models for population density
+and median income show particularly marked changes with the spike at high
+population density and high median incomes disappearing after adjusting for a
+spatiotemporally smoothed term. This likely reflects the early emergence and
+surge of COVID-19 in the greater New York City area as being predominantly a
+feature of the combined effect of human geography and infectious disease
+dynamics where the earliest introductions of COVID-19 were made to dense social
+networks in the US. Other effects remain qualitatively changed little,
+suggesting that the trends in COVID-19 mortality associated with these
+county-level measures remains durable when taking into account spatiotemporal
+autocorrelation.
 
@@
 individual level data, an inability to include all possible metrics, further
-improvements to the methods used may be warranted, and we are unable to make
+improvements to the methods used may be warranted, and an inability to make
 claims about the causal relationships between the variables investigated.
@@
 confounding [@greenland_ecologic_2001]. It's important to note that
-some of the variables are included to contextualize the places in which people
-live, reside, and work and area-based social metrics should be included in
-individual-level analyses so as to accurately capture the effects of both
-individual-level risk factors and risk factors that stem from the contexts in
-which people exist [@testa_public_2022].
+some of the variables in this analysis are included to contextualize the places
+in which people live, reside, and work and area-based social metrics should be
+included even in individual-level analyses so as to accurately capture the
+effects of both individual-level risk factors and risk factors that stem from
+the contexts in which people exist [@testa_public_2022].
 
@@
 to create age-group stratified models. In the applications presented here,
-county-level median age is adjusted for in order to control for age effects.
+county-level median age is adjusted for in order to control for age effects as
+age-specific data were not publicly available without substantial levels of
+suppression.
 
@@
 [@wood_soap_2008]. Taking a critical eye to the assumptions employed reveals yet
-another area of potential improvement, which is the nature of how the complexity of the 
-smoothing splines employed are penalized in the GAM framework. It is conceivable that
+another area of potential improvement, which is the nature of how 
+smoothing splines are penalized in the GAM framework. It is conceivable that
 geographic changes in COVID-19 mortality rates are not equally smooth across 
 all regions of the US, potentially warranting rapid changes in some areas 
-and smooth changes in other areas. The Bayesian wombling literature provides
-the means to identify areas of rapid change in latent surface models and may pose
-a fruitful direction of exploration for future efforts to understand trends
-in COVID-19 mortality rates in the US [@gelfand_bayesian_2015].
+and more smooth changes in other areas in a risk surface model. The Bayesian
+wombling literature provides the means to identify areas of rapid change in
+latent surface models and may pose a fruitful direction of exploration for
+future efforts to understand trends in COVID-19 mortality rates in the US
+[@gelfand_bayesian_2015].
 
@@
 mortality data to document the disparities and inequities in which US counties
-at what times suffered the worst mortality outcomes. The effort to document
+suffered the worst mortality outcomes and at what times. The effort to document
 inequities and explain the driving mechanisms behind them is crucial in building
@@
 paper illustrates new ways of understanding and visualizing the disproportionate
-burden some communities have suffered incorporating complex patterns of
+burden some communities have suffered by incorporating complex patterns of
 interaction across socioeconomic and geopolitical variables and spatiotemporal

### 2022-12-01 manuscript/manuscript.tex
prepare for arxiv submission

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Understanding Complex Patterns in Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models},
  pdfauthor={Christian Testa1,},
  colorlinks=true,
  linkcolor={black},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

\title{Understanding Complex Patterns in Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models}
\author{Christian Testa\textsuperscript{1,*}}
\date{November 30, 2022}

\begin{document}
\maketitle
\begin{abstract}
I present three types of applications of generalized additive models (GAMs) to
COVID-19 mortality rates in the US for the purpose of advancing methods to
document inequities with respect to which communities suffered disproportionate
COVID-19 mortality rates at specific times during the first three years of the
pandemic. First, GAMs can be used to describe the changing
relationship between COVID-19 mortality and county-level covariates
(sociodemographic, economic, and political metrics) over time. Second, GAMs can
be used to perform spatiotemporal smoothing that pools information over time and
space to address statistical instability due to small
population counts or stochasticity resulting in a smooth, dynamic latent
risk surface summarizing the mortality risk associated with geographic locations over
time. Third, estimation of COVID-19 mortality associations with county-level covariates
conditional on a smooth spatiotemporal risk surface allows for more rigorous consideration
of how socio-environmental contexts and policies may have impacted COVID-19 mortality.
Each of these approaches
provides a valuable perspective to documenting inequities in
COVID-19 mortality by addressing the question of which populations have suffered
the worst burden of COVID-19 mortality taking into account the nonlinear
spatial, temporal, and social patterning of disease.
\newline

\noindent Abbreviations used: United States (US), Coronavirus Disease 2019 (COVID-19), Generalized
Additive Model (GAM), Centers for Disease Control and Prevention (CDC), Index of Concentration at the Extremes (ICE), Confidence Interval (CI)
\end{abstract}

\textsuperscript{1} Department of Social and Behavioral Sciences, Harvard T.H. Chan School of Public Health

\textsuperscript{*} Correspondence: \href{mailto:ctesta@hsph.harvard.edu}{Christian Testa \textless{}\href{mailto:ctesta@hsph.harvard.edu}{\nolinkurl{ctesta@hsph.harvard.edu}}\textgreater{}}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

As we enter the third winter with the novel coronavirus disease
COVID-19 in the United States, evidence documenting the intense disparities in
COVID-19 mortality rates comparing socially advantaged and disadvantaged
populations continues to mount. Eliminating inequities in health outcomes has
been stated as a major policy goal of the Biden administration
(\protect\hyperlink{ref-the_white_house_executive_2021}{The White House, 2021}, \protect\hyperlink{ref-the_white_house_advancing_2022}{2022}), representing
a revitalized commitment to health equity and underlining the importance of adequate
data reporting systems that report timely estimates of prevalent health inequities. To this
end, I use generalized additive models (GAMs) as a flexible regression
framework to illustrate the evolving roles and relationships sociodemographic,
geographic, and economic conditions have with respect to trends in COVID-19
mortality. The code, data, and documentation necessary to reproduce the analyses
contained in this paper are online and free to access at \url{https://github.com/ctesta01/covid.gradient.estimation}.

\hypertarget{background}{%
\subsubsection{Background}\label{background}}

Having passed over 1 million COVID-19 deaths in the United States in 2022
(\protect\hyperlink{ref-donovan_us_2022}{Donovan, 2022}), and facing uncertain prospects for the third COVID-19 winter
looming even as new iterations on the COVID-19 vaccines become available, it
remains critical that inequities in COVID-19 outcomes are documented and
analyzed to reckon with the unjust and unfair burden of preventable illness.
Even though the first vaccines were granted emergency use authorization by the
U.S. Food and Drug Administration in 2020 (\protect\hyperlink{ref-mayo_clinic_history_2022}{Mayo Clinic, 2022}), with the
first shots going in arms in December 2020, COVID-19 is still continuing to
cause hundreds of deaths a day in the US in the fall of 2022
(\protect\hyperlink{ref-noauthor_united_2022}{{``United {States} {COVID} - {Coronavirus} {Statistics} - {Worldometer},''} 2022}). The new bivalent vaccines released at the end of
August 2022 contain mRNA sequences from both the original strain as well as the
recently emergent BA.4 and BA.5 lineages in an effort to make the nation's
immunity more up-to-date and robust against the myriad of phylogenetic
directions the COVID-19 virus is evolving to explore
(\protect\hyperlink{ref-office_of_the_commissioner_coronavirus_2022}{Office of the Commissioner, 2022}). Despite the updated bivalent
boosters representing a significant step forward in prevention strategy, less
than 4\% of eligible Americans had taken the booster in the first month after it
became available (\protect\hyperlink{ref-bendix_less_2022}{Bendix, 2022}; \protect\hyperlink{ref-lambert_most_2022}{Lambert, 2022}). As such, and with an
enduring history of inequities in health care access in the US
(\protect\hyperlink{ref-bailey_how_2021}{Bailey et al., 2021}; \protect\hyperlink{ref-blendon_inequities_2002}{Blendon et al., 2002}; \protect\hyperlink{ref-carpenter_health_2021}{Carpenter, 2021}; \protect\hyperlink{ref-chrisler_ageism_2016}{Chrisler et al., 2016}; \protect\hyperlink{ref-feldman_health_2021}{Feldman et al., 2021}; \protect\hyperlink{ref-okonkwo_covid-19_2021}{Okonkwo et al., 2021}; \protect\hyperlink{ref-ortega_ending_2021}{Ortega \& Roby, 2021}; \protect\hyperlink{ref-rapp_statelevel_2022}{Rapp et al., 2022}; \protect\hyperlink{ref-whitehead_outness_2016}{Whitehead et al., 2016}), it is clear that without further intervention not all
communities will be equally able to benefit from the updated vaccines and
inequities in COVID-19 illness and mortality may persist despite the
technological innovations in vaccine technology.

\hypertarget{the-role-of-geography-in-covid-19}{%
\subsubsection{The Role of Geography in COVID-19}\label{the-role-of-geography-in-covid-19}}

\begin{figure*}

{\centering \includegraphics[width=6in]{figures/county_trends_by_division}

}

\caption{Estimates of monthly COVID-19 mortality rates per 100,000 person-years by county organized by Census Division.  For each division, the median trendline and quantile ranges are shown weighted by county population size.}\label{fig:figure-county-trends}
\end{figure*}

Prior literature has demonstrated that the spread and impact of COVID-19 has
varied geography over time reflecting dynamics in the timing of introduction and
transmission events. Methods employed to highlight the geographic patterning in
COVID-19 outcomes have included quantile regression
(\protect\hyperlink{ref-sigler_socio-spatial_2021}{Sigler et al., 2021}), Besag-York-Molli\textquotesingle e mixed models
(\protect\hyperlink{ref-whittle_ecological_2020}{Whittle \& Diaz-Artiles, 2020}), spatial cluster analysis (\protect\hyperlink{ref-sugg_mapping_2021}{Sugg et al., 2021}),
geographically weighted regression (\protect\hyperlink{ref-mollalo_gis-based_2020}{Mollalo et al., 2020}; \protect\hyperlink{ref-park_covid-19_2021}{Park et al., 2021}), and others.

Figure \ref{fig:figure-county-trends} shows the monthly COVID-19 mortality
rates for the counties grouped within each of the nine U.S. Census Divisions
(\protect\hyperlink{ref-us_department_of_commerce_economics_and_statistics_administration_census_2000}{U.S. Department of Commerce Economics and Statistics Administration \& U.S. Census Bureau, 2000}).
The figure summarizes each division's median mortality rates
weighted by county population size.
Notably, the mortality associated
with the early surge of cases starting in New York City and spreading through New York,
New Jersey, and Massachusetts is visible in the Middle Atlantic and New England
division figures. The figure also illustrates how the first peak in the
mortality time-series for states in the Midwest (West North Central, East North
Central) occurred later, in late 2020 and going into early 2021.

In the US context, one of the key aspects to the geographic story of COVID-19's
spread and diffusion was the early surge of cases and epicenter in New York City
during March 2020 (\protect\hyperlink{ref-thompson_covid-19_2020}{Thompson, 2020}) followed by subsequent waves of cases in the
South and Midwest (\protect\hyperlink{ref-glenza_covid_2020}{Glenza, 2020}; \protect\hyperlink{ref-scott_these_2020}{Scott, 2020}; \protect\hyperlink{ref-shumaker_covid-19_2020}{Shumaker \& Wu, 2020}). As Park et al.~
summarized the trends in the US from March 2020 to May 2021, ``hot spots have shifted from densely populated cities and the states with a high percentage of socially vulnerable individuals to the states with relatively relaxed social distancing requirements, and then to the states with low vaccination rates'' (\protect\hyperlink{ref-park_covid-19_2021}{2021}).

When considering the drivers of the COVID-19 pandemic, it's necessary to note that
geography and social conditions are inextricably linked. In July 2021, the CDC
reported that ``the COVID-19 cumulative death rate in non-metropolitan areas has
exceeded that of metropolitan areas since December 2020,'' noting that of the
approximately 1/5th of Americans who live in rural areas, many ``are considered
highly vulnerable according to CDC's Social Vulnerability Index (SVI), which
includes factors such as housing, transportation, socioeconomic status, race,
and ethnicity'' (\protect\hyperlink{ref-cdc_location_2021}{CDC, 2021}). Moreover, rural communities often have
lower health insurance rates, higher disability rates, older populations, and
limited access to health care. One of CDC's Morbidity and Mortality Weekly
Reports found that vaccination against COVID-19 was lower in rural communities
than in urban communities between December 2020 and April 2021
(\protect\hyperlink{ref-murthy_disparities_2021}{Murthy, 2021}).

\hypertarget{the-social-determinants-of-covid-19-mortality}{%
\subsubsection{The Social Determinants of COVID-19 Mortality}\label{the-social-determinants-of-covid-19-mortality}}

Even since the beginning of the COVID-19 outbreak in the US, data reflected
sharp inequities in mortality rates. During January 22nd to May 5th 2020,
county COVID-19 mortality rates were 4.94 (95\% CI 4.78, 5.09) times higher in
counties in the highest quintile of percent People of Color (61\%-100\%) compared
to counties in the lowest quintile of percent People of Color (0\%-17.2\%)
(\protect\hyperlink{ref-chen_revealing_2021}{Chen \& Krieger, 2021}). This was not wholly unanticipated: as COVID-19 was
beginning to take off in the US, some were already calling attention to the fact
that COVID-19 threatened to exacerbate existing disparities
(\protect\hyperlink{ref-kim_covid-19_2020}{Kim et al., 2020}). Kim, Marrast, and Conigliaro noted at least three structural
barriers in COVID-19 prevention and care: 1) originally requiring
residents to have a doctor's prescription for a COVID-19 test reduced the
opportunity for healthcare for People of Color as they are less likely to have a
primary care provider; 2) drive-through testing made testing disadvantaged those
who rely on public transportation; and 3) quarantining at home while waiting the
7-10 days originally required for COVID-19 test results to come back posed an
economic and social challenge that many in already financially difficult
situations may not have been able to take on (\protect\hyperlink{ref-kim_covid-19_2020}{2020}). Others noted
yet more reasons why COVID-19 threatened to worsen an already inequitable
healthcare landscape in the US: in particular, those who reside in prisons and jails,
immigrants and undocumented people, people with disabilities, and people
experiencing homeless all face additional challenges in seeking and getting the
healthcare they deserve (\protect\hyperlink{ref-okonkwo_covid-19_2021}{Okonkwo et al., 2021}). Even though stay-at-home orders
designed to mitigate spread that were prevalent in many states
(\protect\hyperlink{ref-moreland_timing_2020}{Moreland et al., 2020}), workers functioning in capacities essential to the
functioning of critical infrastructure operations (later termed ``essential
workers'') were exposed to heightened risk of COVID-19 transmission
(\protect\hyperlink{ref-hanage_covid-19_2020}{Hanage et al., 2020}; \protect\hyperlink{ref-national_bureau_of_economic_research_measuring_2021}{National Bureau of Economic Research, 2021}; \protect\hyperlink{ref-national_conference_of_state_legislatures_covid-19_2021}{National Conference of State Legislatures, 2021}; \protect\hyperlink{ref-the_lancet_plight_2020}{The Lancet, 2020}; \protect\hyperlink{ref-wei_risk_2022}{Wei et al., 2022}). It's clear that those with more
structurally enfranchised privileges have been more able to mitigate their risk
of negative health outcomes associated with COVID-19 --- during February 1 to
April 1, 2020, New York City residents who lived in more affluent neighborhoods
were more likely to have left the city, while New York City residents from more
Black and Hispanic
neighborhoods were more likely to continue working (\protect\hyperlink{ref-coven_disparities_2020}{Coven \& Gupta, 2020}).

\begin{quote}
Contrary to the oft used phrase that the `virus does not discriminate', the
data presented here suggest that this virus, as many other infectious diseases,
has the greatest implications for the most vulnerable people. The intersections
between health and human rights are clear---the health of a society and
vulnerability to a pandemic are directly related to its human rights track
record for those who are marginalised. (\protect\hyperlink{ref-okonkwo_covid-19_2021}{Okonkwo et al., 2021})
\end{quote}

When vaccines became available, vaccine appointments were often only available
to be scheduled through online web-portals contributing to the inequities
between those who had internet access and technological literacy and those who
didn't (\protect\hyperlink{ref-press_inequities_2021}{Press et al., 2021}). Vaccination sites have not been equally
distributed and areas determined to be vaccine deserts have been found to have
disproportionately more Black and Hispanic residents (\protect\hyperlink{ref-rader_spatial_2021}{Rader et al., 2021}). In
fact, healthcare facilities in counties with higher Black composition had 32\%
(95\% CI 14\%-47\%) lower odds of serving as vaccine sites
(\protect\hyperlink{ref-hernandez_disparities_2022}{Hernandez et al., 2022}).

What vaccination has been administered hasn't suddenly erased the unequal burden
of COVID-19 either; in August 2022 the New York Times was reporting ``Black death
rates at this winter's peak were greater than those of white people by 34
percent in rural areas, 40 percent in small or medium cities and 57 percent in
big cities and their suburbs'' (\protect\hyperlink{ref-goldstein_during_2022}{Goldstein, 2022}). As COVID-19 case and
mortality rates have waxed and waned, the inequities have widen and shrunken,
often with racial/ethnic inequities growing during times when COVID-19 rates
have surged (\protect\hyperlink{ref-hill_covid-19_2022}{Hill \& Artiga, 2022}). During 2022, the age-standardized
COVID-19 mortality rates for white people have, at times, been slightly
higher than those of Black and Hispanic people, predominantly because the
mortality rate among white people has increased (\protect\hyperlink{ref-johnson_whites_2022}{Johnson \& Keating, 2022}).
It's important to note that white COVID-19 mortality rates overtaking
the Black mortality rates does not imply that equity has been established: neither
does this undo the cumulative impact of mortality (which has been twice as high for
Black people compared to white people (\protect\hyperlink{ref-hill_covid-19_2022}{Hill \& Artiga, 2022})), nor does it imply
the underlying systemic barriers to equity have been overturned
(\protect\hyperlink{ref-del_rios_covid-19_2022}{Del Rios et al., 2022}). As Del Rios, Chomilo, and Lewis note, instead,
COVID-19 leaves in its wake more years of life expectancy lost, wages lost, and
degradation of mental health in Communities of Color.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{data-sources}{%
\subsection{Data Sources}\label{data-sources}}

The following variables were retrieved at the county level:

\begin{itemize}
\tightlist
\item
  Counts of COVID-19 deaths (\protect\hyperlink{ref-the_new_york_times_coronavirus_2021}{The New York Times, 2021}).
\item
  Population size estimates for 2020 from the U.S. Census (\protect\hyperlink{ref-us_census_bureau_2020_2021}{US Census Bureau, 2021}).
\item
  Median age, median household income, racial/ethnic composition, population
  density, percent below the federal poverty line, and number of households with
  high (\$100k+)/low (\textless\$25k) household income by racial/ethnic group from the
  2014-2019 5-year American Community Survey (\protect\hyperlink{ref-us_census_bureau_american_2020}{US Census Bureau, 2020})
  through the \texttt{tidyverse} R package (\protect\hyperlink{ref-walker_tidycensus_2022}{Walker \& Herman, 2022}).
\item
  Votes cast in the 2020 presidential election (\protect\hyperlink{ref-mit_election_data_and_science_lab_county_2022}{MIT Election Data and Science Lab, 2022})
\end{itemize}

\hypertarget{generalized-additive-models}{%
\subsection{Generalized Additive Models}\label{generalized-additive-models}}

Generalized additive models (GAMs) improve upon generalized linear models by
allowing for the fitting of smooth functions that transform right-hand-side variables.
This is a convenient means to account for nonlinear relationships between the
outcome and predictor variables. Whereas a generalized linear model may look like
\[ g(\mu_i) = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} ...\]
a generalized additive model could look like
\[ g(\mu_i) = \mathbf A_i \mathbf \theta + f_1(x_{i1}) + f_2(x_{i2}) + f_3(x_{i3}, x_{i4}) + \dots\]
where the expected value of the outcome is given as \(\mu_i \equiv \mathbb E(Y_i)\),
\(\mathbf A_i\) is a row of the model matrix
for any strictly non-parametric model components, \(\mathbf \theta\) is the corresponding
parameter vector, and the \(f_j\) are smooth functions of the covariates \(x_k\) (\protect\hyperlink{ref-wood_generalized_2017}{S. N. Wood, 2017}).

\begin{figure}
\includegraphics[width=1\linewidth]{figures/gam_demo_figure} \caption{A demonstration of fitting regression models using B-spline basis functions. Part A. shows the simulated dataset. Part B. shows the B-spline basis functions. Part C. shows the weighting of the B-spline basis functions and their weighted sum with an intercept added to create the regression model fit to the data.}\label{fig:gam-demo-figure}
\end{figure}

The smooth functions estimated as part of fitting a GAM are constructed using
spline basis functions. These spline basis functions allow for the smooth
interpolation of trends in the data allowing for the incorporation of
nonlinearity. To avoid overfitting the data, a penalty term is introduced that
controls the degree of ``wiggliness'' or smoothness and this penalty term is fit
using generalized cross validation. B-splines are one kind of spline basis function
that is commonly used and are especially popular because of their property that they
are non-negative on only a finite interval.

While a variety of spline-based approaches exist (cubic splines, B- and
P-splines, thin-plate splines, etc.), our particular application setting
warrants the use of tensor-product smooths because we require anisotropic
penalization. That is to say, when we seek to create spatiotemporally smoothed
model estimates, it's inappropriate to assume that the amount of smoothing
across space should be the same as the amount of smoothing across time because
they're in fundamentally different units (space being measured in Cartesian
coordinates and time being measured in years, months, days, etc.). More details
about tensor-product smooths are available in Simon Wood's \emph{Generalized Additive Models - An Introduction with R} (\protect\hyperlink{ref-wood_generalized_2017}{2017}).

\hypertarget{overdispersion-and-event-modeling}{%
\subsection{Overdispersion and Event Modeling}\label{overdispersion-and-event-modeling}}

Above and beyond using the GAM framework to allow for flexible, nonlinear
relationships between our observed county-level variables and COVID-19
mortality, we must have a model specification that agrees with the data
generating process. In our case, the data generated are counts of deaths per
population, and the class of models most suited to represent counts of events
are Poisson, quasi-Poisson, and negative binomial models. Here we've chosen to
use the negative binomial model as it accounts for overdispersion and has
the intuitive interpretation of a Poisson model with gamma distributed underlying
rate parameter (\protect\hyperlink{ref-gelman_bayesian_2013}{Gelman et al., 2013}).

A negative binomial model in the context of modeling count data can be written
\[y_i \sim \text{NegativeBinomial}(u_i \theta_i, \phi),\]
where \(y_i\) are the counts observed, \(u_i\) is the ``exposure'', \(\theta_i\) are
the rates, and \(\phi\) determines the amount of overdispersion. The rates
are modeled as \(\theta_i = e^{X_i \beta}\) where \(X_i\) are the observed
covariates and \(\beta\) are the coefficients on the covariates corresponding
to log rate ratios. The logarithm of the exposure, \(\log(u_i)\) is often called
(and later herein referenced as) the offset. In epidemiological contexts, the offset
is often representative of the amount of person-time during which observations were recorded. Whereas the Poisson model holds that \(\text{var}(y) = \mu\) where \(\mu\)
is the average rate, the negative binomial model instead assumes that
\(\text{var}(y) = \mu + \mu^2/\phi\). The Poisson model
is a special case of the negative binomial model when \(\phi \to \infty\).

An alternative and equivalent formulation of the negative binomial that is commonly
used makes the connection to the Poisson model even more clear: instead of
\(\theta_i\) and \(\phi\), using \(\alpha\) and \(\beta\),
\[y \sim \text{NegativeBinomial}(\alpha, \beta), \text{ and}\]
\[\text{NegativeBinomial}(y|\alpha, \beta) = \int \text{Poisson}(y|\theta) \text{Gamma}(\theta | \alpha, \beta) d\theta.\]

Introductions to and additional exposition on the negative binomial model,
especially in the context of modeling counts of events outcome data, are
available in \emph{Bayesian Data Analysis} and \emph{Regression and Other Stories}
(\protect\hyperlink{ref-gelman_regression_2020}{Gelman et al., 2020}, \protect\hyperlink{ref-gelman_bayesian_2013}{2013}).

\hypertarget{variables-of-interest}{%
\subsection{Variables of Interest}\label{variables-of-interest}}

The following variables have been included as covariates of interest:

\begin{itemize}
\tightlist
\item
  Median Age
\item
  Population Density per Square Mile
\item
  Median Household Income
\item
  Proportion in Poverty
\item
  the Index of Concentration at the Extremes for Racialized Economic Segregation
\item
  Political Lean in the 2020 Election (1 = 100\% Republican votes, -1 = 100\% Democratic Votes)
\end{itemize}

\hypertarget{index-of-concentration-at-the-extremes-for-racialized-economic-segregation}{%
\paragraph{Index of Concentration at the Extremes for Racialized Economic Segregation}\label{index-of-concentration-at-the-extremes-for-racialized-economic-segregation}}

The Index of Concentration at the Extremes (ICE) is a measure which describes
how concentrated a given area's population is in terms of the extreme ends of
privilege and marginalization (\protect\hyperlink{ref-krieger_public_2016}{Krieger et al., 2016}).
In general, the ICE measure is formulated as

\[ \text{ICE} = \frac{\text{Number of People in Most Privileged Category} - \text{Number of People in Least Privileged Category}}{\text{Total Population}}\]

Applying the ICE approach to a specific context involves defining the axes of
privilege of interest. In this case, data on racialized economic segregation
are used from the US Census American Community Survey on white households
earning more than \$100,000 a year (the most privileged group) or households of
People of Color earning less than \$25,000 a year (the least privileged group).
This variable is referred to throughout as the ICE for racialized economic
segregation, or \texttt{ICEraceinc} in the code.

Compared with the Gini
coefficient which is one of the most popular methods for summarizing area-level
rates of inequities, the ICE has the advantage that it is suitable for
describing inequities at smaller area levels (\protect\hyperlink{ref-krieger_public_2016}{Krieger et al., 2016}). While the
Gini coefficient measures within-area dissimilarity (as in, for example, how
unequal wealth is distributed within a county), the ICE measure establishes
where on a spectrum a given county's population falls allowing for comparison
across counties. The Gini coefficient suffers from the fact that areas which are
made up of relatively homogeneous populations will appear as having low
within-area inequality (and therefore low Gini coefficient). Instead, the ICE
measures how much of the population is either privileged or not. The Gini
coefficient remains useful for reporting on the degree of inequity in larger
areal units (like countries, states, and regions), but at smaller areal units
(like counties, ZIP codes, census tracts) can be more difficult to interpret and
compare. Maps of ICE measure can elucidate what spatial social segregation and
polarization exist, and the ICE for racialized economic segregation has been
repeatedly and significantly associated with COVID-19 outcomes
(\protect\hyperlink{ref-brown_ecological_2021}{Brown et al., 2021}; \protect\hyperlink{ref-chen_revealing_2021}{Chen \& Krieger, 2021}; \protect\hyperlink{ref-eichenbaum_health_2022}{Eichenbaum \& Tate, 2022}; \protect\hyperlink{ref-hanage_covid-19_2020}{Hanage et al., 2020}; \protect\hyperlink{ref-krieger_relationship_2022}{Krieger et al., 2022}; \protect\hyperlink{ref-saha_neighborhood-level_2020}{Saha \& Feldman, 2020}; \protect\hyperlink{ref-sonderlund_racialized_2022}{Sonderlund et al., 2022}).

\hypertarget{political-lean}{%
\paragraph{Political Lean}\label{political-lean}}

Political lean has been associated with COVID-19 mortality in numerous studies,
with plausible mechanisms explaining the association including differences in
non-pharmaceutical intervention uptake (mask usage, social distancing,
quarantining), differences in rates of vaccination, differences in political
leadership's messaging, resource allocation, and the adoption of policy
interventions (\protect\hyperlink{ref-gonzalez_conservatism_2021}{Gonzalez et al., 2021}; \protect\hyperlink{ref-grossman_political_2020}{Grossman et al., 2020}; \protect\hyperlink{ref-kaashoek_evolving_2021}{Kaashoek et al., 2021}; \protect\hyperlink{ref-krieger_relationship_2022}{Krieger et al., 2022}; \protect\hyperlink{ref-leonhardt_red_2021}{Leonhardt, 2021}).

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{application-1-non-spatial-covariate-effects-over-time}{%
\subsection{Application 1: Non-Spatial Covariate Effects Over Time}\label{application-1-non-spatial-covariate-effects-over-time}}

The GAM models shown in Figure \ref{fig:one-variable-over-time} are fit with the following structure using
the \texttt{gam} function from the \texttt{mgcv} package (\protect\hyperlink{ref-wood_mgcv_2022}{S. Wood, 2022}):

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{gam}\NormalTok{(}
  \AttributeTok{formula =}\NormalTok{ deaths }\SpecialCharTok{\textasciitilde{}} \FunctionTok{s}\NormalTok{(median\_age) }\SpecialCharTok{+} \FunctionTok{te}\NormalTok{(covariate, date, }\AttributeTok{d=}\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{)), }\CommentTok{\# regression formula}
  \AttributeTok{offset =} \FunctionTok{log}\NormalTok{(popsize}\SpecialCharTok{/}\FloatTok{1e5}\SpecialCharTok{/}\DecValTok{12}\NormalTok{), }\CommentTok{\# our offset represents the person{-}time}
  \AttributeTok{data =}\NormalTok{ df,     }\CommentTok{\# our data{-}set of county{-}level observations}
  \AttributeTok{family =} \FunctionTok{nb}\NormalTok{(}\AttributeTok{link=}\StringTok{\textquotesingle{}log\textquotesingle{}}\NormalTok{)  }\CommentTok{\# indicates negative binomial family and a log{-}link function}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

The formula used puts a one-dimensional smoothing spline on median age to represent a
nonlinear age-effect and a two-dimensional tensor-product smooth on the
interaction between the given covariate and the date. The \texttt{d=c(1,1)} argument
provides the instruction necessary to consider covariate and date as being on
separate scales and therefore to fit the tensor-product smooth with anisotropy ---
that is, to allow for independent amounts of scaling in the dimensions of the
covariate and time. The offset used structures the regression to estimate rates
in units of person-time per 100,000 person-years. Since the death counts are
aggregated to the monthly level, the person-time in units of 100,000
person-years are calculated by taking each county's population size, dividing by
100,000, and dividing by 12 (for the 12 months in a year).

The above structure is used to estimate models for our different \texttt{covariate}
variables of interest: median income, percent in poverty, the ICE for racialized
economic segregation, political lean. The model presenting median age treats
median age as the main covariate of interest including it as the \texttt{te(covariate,\ date,\ d=c(1,1))} and dropping the \texttt{s(median\_age)} term which otherwise becomes redundant.
Results of these models are summarized in Figure
\ref{fig:one-variable-over-time} where the additional COVID-19 mortality
associated with each covariate is visualized.

\begin{figure}

{\centering \includegraphics[width=1\linewidth,height=0.75\textheight]{figures/nonspatial_panel_figure}

}

\caption{Additional COVID-19 mortality associated with covariates over time. A) age and time, B) log (base 10) population density and time, C) median income and time, D) proportion in poverty and time, E) ICEraceinc and time, F) political lean and time. In panels B-F the effect of age is marginalized out using the median age in the US, 38.8 (\protect\hyperlink{ref-us_census_bureau_nation_2022}{US Census Bureau, 2022}).}\label{fig:one-variable-over-time}
\end{figure}

Likelihood ratio tests confirmed that models with covariates interacted with
time had significantly lower residual deviance (\(p \leq 2.2e^{-16}\) for all
models) compared to models only including a spline term on median age and the
given covariate not interacted with time. The model interacting median age
and time was compared to a model with a spline for median age alone. Akaike Information
Criteria values were also lower for all models compared to models that did
not interact the covariates with time.

A more complex non-spatial application of GAMs to describe the distribution of
COVID-19 mortality in the US over time is to consider the associations of
mortality with three-way interactions of time and two covariates taken together.
In the following example, the formula used is \texttt{deaths\ \textasciitilde{}\ te(date,\ ICEraceinc,\ median\_age,\ d=c(1,1,1))}. Again, the \texttt{d} argument specifying the marginal
basis dimensions is used indicate that each of the \texttt{date}, \texttt{ICEraceinc}, and
\texttt{median\_age} measures are in different units and should not be smoothed assuming
that a one unit difference in one variable is comparable to a one unit difference
in another variable. This approach is represented in Figure \ref{fig:two-variables-over-time}.

\begin{figure}
\includegraphics[width=1\linewidth]{figures/ICEraceinc_age_panel_figure} \caption{The changing interacted effect of the ICE for racialized economic segregation and age over time. A) May 2020, B) November 2020, C) May 2021, D) October 2021, E) February 2022, F) August 2022. An animated version is available online at \url{https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/readme.md}}\label{fig:two-variables-over-time}
\end{figure}

\newpage

\hypertarget{application-2-spatiotemporal-smoothing}{%
\subsection{Application 2: Spatiotemporal Smoothing}\label{application-2-spatiotemporal-smoothing}}

By fitting GAMs with a tensor-product smoothing term on latitude, longitude, and
time we can estimate a spatiotemporally smoothed trend in COVID-19 mortality. To
do this, the GAM is constructed similarly as in Application 1 but with the
smoothing term specified as \texttt{te(latitude,\ longitude,\ time,\ d=c(2,1))} where
\texttt{d=c(2,1)} indicates that \texttt{latitude} and \texttt{longitude} share the same dimensions
(i.e., both are spatial and in units of degrees) while the \texttt{time} data are in
separate units. Note that results presented in Application 2 and 3 are based on
county-level data from the contiguous US excluding Alaska and Hawaii.

Using GAMs to present spatiotemporally smoothed estimates of COVID-19 mortality
allows for the synthesis of trends over time. Whereas the raw rates of COVID-19
mortality are noisy due to 72\% of the land area in the US being classified as
rural and low population density (\protect\hyperlink{ref-health_resources__services_administration_defining_2022}{Health Resources \& Services Administration, 2022}), the spatiotemporally smoothed GAM estimates
use population weighting via the offset specified to estimate a latent surface
that represents localized mortality rate averages in spatial coordinates and in
time.

Figure \ref{fig:crude-vs-smoothed} shows the difference between crude mortality
rates and spatially smoothed mortality rates in January 2022 to illustrate the
level of noise present in raw rates and how spatially smoothing synthesizes
local geographic patterns into trends that can be meaningfully
interpreted as local area mortality risk levels with information information pooled across
nearby county rates.

Figure \ref{fig:spatiotemporal-panel-figure} shows the results from a
spatiotemporally smoothed model in select months to highlight how spatiotemporal
smoothing can yield results that synthesize trends in space and
time.

\begin{figure}
\includegraphics[width=1\linewidth]{figures/side_by_side_crude_and_smoothed} \caption{Comparison of crude vs. smoothed mortality rates. A) Crude mortality rates per 100,000 person-years in January 2022. B) Smoothed mortality rates from a GAM applied to mortality rates from January 2022.}\label{fig:crude-vs-smoothed}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{figures/temporal_panel_figure} \caption{Spatiotemporally smoothed COVID-19 mortality estimates from a GAM fit to data from March 2020-August 2022. Panels show months selected to highlight the changing spatial patterns of COVID-19 mortality risk over time. An animated version is available online at \url{https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/09_spatiotemporal_models/animation/readme.md}}\label{fig:spatiotemporal-panel-figure}
\end{figure}

\newpage

\hypertarget{application-3-estimating-covariate-effects-adjusted-for-spatiotemporal-autocorrelation}{%
\subsection{Application 3: Estimating Covariate Effects Adjusted for Spatiotemporal Autocorrelation}\label{application-3-estimating-covariate-effects-adjusted-for-spatiotemporal-autocorrelation}}

The final application of GAMs presented here is to estimate the effects of
covariates after adjusting for a spatiotemporally smoothed latent risk surface.
This has the interpretation of asking what are the changes to
COVID-19 mortality rates associated with each covariate after taking into
account regional geographic patterns in COVID-19 mortality over time. In
particular, this is relevant for understanding what county-level measures are
associated with elevated COVID-19 mortality even after adjusting for where
COVID-19 rates were locally elevated or depressed due to variation in the timing
of local introduction, transmission, and diffusion.

\begin{figure}
\includegraphics[width=1\linewidth]{figures/county_network_and_morans_plot} \caption{The spatial autocorrelation of COVID-19 rates in January 2022. Panel A shows the graph of US counties connected according to which are neighbors. Panel B shows a scatterplot depicting each US counties' COVID-19 mortality rate (x-axis) compared to the average of its neighbors' COVID-19 mortality rates (y-axis). A regression line shows the association between counties' COVID-19 rates and their neighboring counties' COVID-19 mortality rates.  Dotted lines indicated the average for US county-level crude COVID-19 mortality rates (vertical) and for the average of each counties' neighboring counties' COVID-19 mortality rates (horizontal).}\label{fig:neighboring-graph-morans-i}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=0.5\linewidth]{figures/temporal_acf}

}

\caption{The median (black) and quantiles (shaded bands) of temporal autocorrelation of counties' COVID-19 mortality rates over time during April 2020 — August 2022.}\label{fig:temporal-acf}
\end{figure}

To motivate the need for and application of spatiotemporal smoothing in
assessing the associated changes in COVID-19 mortality with county-level
covariates, Figure \ref{fig:neighboring-graph-morans-i} presents the Moran's I
diagnostic plot for January 2022 which summarizes the amount of autocorrelation
between counties' COVID-19 mortality rates and the average of
each counties' neighboring counties' COVID-19 mortality rates. The observed
autocorrelation indicates that there is an association between counties'
COVID-19 rates and their neighboring counties' COVID-19 mortality rates, implying that
without taking this patterning into account, regression models that do not
explicitly model the effect of spatial autocorrelation may be biased due to
inappropriately assuming that county-level data are independent of one another. Second,
while the Moran's I plot demonstrates the spatial-correlation present in the data
at a single time-point, the following Figure \ref{fig:temporal-acf} shows the
tendency of county COVID-19 observations to be autocorrelated between subsequent
months.

\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{figures/spatiotemporal_panel_figure}

}

\caption{Estimates of increases in COVID-19 mortality associated with county-level covariates without (first) and with (second) adjustment for spatiotemporal autocorrelation. Panels A and B show COVID-19 mortality increases associated with varying levels of median age over time; C and D show associations with log population density; E and F median income; G and H for proportion in poverty; I and J the ICE for racialized economic segregation; K and L political lean.}\label{fig:with-and-without-spatiotemporal-adjustment}
\end{figure}

Figure \ref{fig:with-and-without-spatiotemporal-adjustment} shows the
additional COVID-19 mortality associated with each covariate over time without and
with adjustment for spatiotemporal trends. The models for population density
and median income show particularly marked changes with the spike at high
population density and high median incomes disappearing after adjusting for a
spatiotemporally smoothed term. This likely reflects the early emergence and
surge of COVID-19 in the greater New York City area as being predominantly a
feature of the combined effect of human geography and infectious disease
dynamics where the earliest introductions of COVID-19 were made to dense social
networks in the US. Other effects remain qualitatively changed little,
suggesting that the trends in COVID-19 mortality associated with these
county-level measures remains durable when taking into account spatiotemporal
autocorrelation.

\newpage

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

This paper presents three applications of GAMs to county-level COVID-19
mortality data from the US during March 2020 to August 2022. This paper
addresses an important need, which is to assess what measures of the
socioeconomic and geopolitical landscape are associated with increases in
COVID-19 mortality in a way that takes into account how infectious disease
dynamics are locally correlated in space and time. However, the approaches
outlined in this paper are not without several limitations which include (at
least) the following: the potential for cross-level confounding, lack of
individual level data, an inability to include all possible metrics, further
improvements to the methods used may be warranted, and an inability to make
claims about the causal relationships between the variables investigated.

In analyses that use area-level aggregated data, the ecologic fallacy refers to
the possibility that associations between area-level measures and outcomes may
not reflect the associations that would be observed if analyzing individual
level data on the same measures and outcomes as a result of cross-level
confounding (\protect\hyperlink{ref-greenland_ecologic_2001}{Greenland, 2001}). It's important to note that
some of the variables in this analysis are included to contextualize the places
in which people live, reside, and work and area-based social metrics should be
included even in individual-level analyses so as to accurately capture the
effects of both individual-level risk factors and risk factors that stem from
the contexts in which people exist (\protect\hyperlink{ref-testa_public_2022}{Testa et al., 2022}).

One particular improvement that could be made in future work is to address age
adjustment by using direct or indirect age standardization (\protect\hyperlink{ref-anderson_age_1998}{Anderson et al., 1998}) or
to create age-group stratified models. In the applications presented here,
county-level median age is adjusted for in order to control for age effects as
age-specific data were not publicly available without substantial levels of
suppression.

Additional variables like vaccination rates, mobility data, further variables
that relate to risk factors for COVID-19, and other COVID-19 related outcomes
like case trends and hospitalizations may be worthwhile to investigate under a
similar framework. The emphasis on trends in COVID-19 mortality in the
applications in this paper was done to prioritize presenting data that are less
subject to the reporting inconsistencies in COVID-19 cases and
hospitalizations (\protect\hyperlink{ref-galaitsi_challenges_2021}{Galaitsi et al., 2021}).

Further improvements to the methods employed here could include use of soap film
smoothing splines which are more appropriate in settings where non-convex
geographic boundaries are common (as in with peninsulas and bays)
(\protect\hyperlink{ref-wood_soap_2008}{S. N. Wood et al., 2008}). Taking a critical eye to the assumptions employed reveals yet
another area of potential improvement, which is the nature of how
smoothing splines are penalized in the GAM framework. It is conceivable that
geographic changes in COVID-19 mortality rates are not equally smooth across
all regions of the US, potentially warranting rapid changes in some areas
and more smooth changes in other areas in a risk surface model. The Bayesian
wombling literature provides the means to identify areas of rapid change in
latent surface models and may pose a fruitful direction of exploration for
future efforts to understand trends in COVID-19 mortality rates in the US
(\protect\hyperlink{ref-gelfand_bayesian_2015}{Gelfand \& Banerjee, 2015}).

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

This work introduces generalized additive models (GAMs) applied to COVID-19
mortality data to document the disparities and inequities in which US counties
suffered the worst mortality outcomes and at what times. The effort to document
inequities and explain the driving mechanisms behind them is crucial in building
up the evidence necessary to enact policy changes that can mitigate unjust and
preventable harm in the future. To this end, it is critical that efforts to
model infectious disease dynamics take into account how disease transmission and
outcomes are spatiotemporally patterned, violating any misplaced assumptions
that area-level disease rates can be treated as independent observations. By
presenting novel applications of GAMs to trends in COVID-19 mortality, this
paper illustrates new ways of understanding and visualizing the disproportionate
burden some communities have suffered by incorporating complex patterns of
interaction across socioeconomic and geopolitical variables and spatiotemporal
trends.

\newpage

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-anderson_age_1998}{}}%
Anderson, R. N., Rosenberg, H. M., \& National Center for Health Statistics. (1998). Age standardization of death rates; implementation of the year 2000 standard. \emph{National Vital Statistics Report}. \url{https://stacks.cdc.gov/view/cdc/13357}

\leavevmode\vadjust pre{\hypertarget{ref-bailey_how_2021}{}}%
Bailey, Z. D., Feldman, J. M., \& Bassett, M. T. (2021). How {Structural} {Racism} {Works} --- {Racist} {Policies} as a {Root} {Cause} of {U}.{S}. {Racial} {Health} {Inequities}. \emph{New England Journal of Medicine}, \emph{384}(8), 768--773. \url{https://doi.org/10.1056/NEJMms2025396}

\leavevmode\vadjust pre{\hypertarget{ref-bendix_less_2022}{}}%
Bendix, A. (2022). Less than 4\% of eligible people have gotten updated {Covid} booster shots, one month into the rollout. In \emph{NBC News}. \url{https://www.nbcnews.com/health/health-news/updated-covid-booster-shots-doses-administered-cdc-rcna48960}

\leavevmode\vadjust pre{\hypertarget{ref-blendon_inequities_2002}{}}%
Blendon, R. J., Schoen, C., DesRoches, C. M., Osborn, R., Scoles, K. L., \& Zapert, K. (2002). Inequities {In} {Health} {Care}: {A} {Five}-{Country} {Survey}. \emph{Health Affairs}, \emph{21}(3), 182--191. \url{https://doi.org/10.1377/hlthaff.21.3.182}

\leavevmode\vadjust pre{\hypertarget{ref-brown_ecological_2021}{}}%
Brown, K. M., Lewis, J. Y., \& Davis, S. K. (2021). An ecological study of the association between neighborhood racial and economic residential segregation with {COVID}-19 vulnerability in the {United} {States}' capital city. \emph{Annals of Epidemiology}, \emph{59}, 33--36. \url{https://doi.org/10.1016/j.annepidem.2021.04.003}

\leavevmode\vadjust pre{\hypertarget{ref-carpenter_health_2021}{}}%
Carpenter, E. (2021). {``{The} {Health} {System} {Just} {Wasn}'t {Built} for {Us}''}: {Queer} {Cisgender} {Women} and {Gender} {Expansive} {Individuals}' {Strategies} for {Navigating} {Reproductive} {Health} {Care}. \emph{Women's Health Issues}, \emph{31}(5), 478--484. \url{https://doi.org/10.1016/j.whi.2021.06.004}

\leavevmode\vadjust pre{\hypertarget{ref-cdc_location_2021}{}}%
CDC. (2021). Location, location, location --~{COVID} {Data} {Tracker} {Weekly} {Review}. In \emph{Centers for Disease Control and Prevention}. \url{https://www.cdc.gov/coronavirus/2019-ncov/covid-data/covidview/past-reports/07162021.html}

\leavevmode\vadjust pre{\hypertarget{ref-chen_revealing_2021}{}}%
Chen, J. T., \& Krieger, N. (2021). Revealing the {Unequal} {Burden} of {COVID}-19 by {Income}, {Race}/{Ethnicity}, and {Household} {Crowding}: {US} {County} {Versus} {Zip} {Code} {Analyses}. \emph{Journal of Public Health Management and Practice}, \emph{27}(1), S43--S56. \url{https://doi.org/10.1097/PHH.0000000000001263}

\leavevmode\vadjust pre{\hypertarget{ref-chrisler_ageism_2016}{}}%
Chrisler, J. C., Barney, A., \& Palatino, B. (2016). Ageism can be {Hazardous} to {Women}'s {Health}: {Ageism}, {Sexism}, and {Stereotypes} of {Older} {Women} in the {Healthcare} {System}. \emph{Journal of Social Issues}, \emph{72}(1), 86--104. \url{https://doi.org/10.1111/josi.12157}

\leavevmode\vadjust pre{\hypertarget{ref-coven_disparities_2020}{}}%
Coven, J., \& Gupta, A. (2020). \emph{Disparities in {Mobility} {Responses} to {COVID}-19}. \url{https://arpitgupta.info/s/DemographicCovid.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-del_rios_covid-19_2022}{}}%
Del Rios, M., Chomilo, N. T., \& Lewis, N. A. (2022). Covid-19 is an inverse equity story, not a racial equity success story. In \emph{STAT}. \url{https://www.statnews.com/2022/10/25/covid-19-inverse-equity-story-not-racial-equity-success-story/}

\leavevmode\vadjust pre{\hypertarget{ref-donovan_us_2022}{}}%
Donovan, D. (2022). U.{S}. {Officially} {Surpasses} 1 {Million} {COVID}-19 {Deaths}. In \emph{Johns Hopkins Coronavirus Resource Center}. \url{https://coronavirus.jhu.edu/from-our-experts/u-s-officially-surpasses-1-million-covid-19-deaths}

\leavevmode\vadjust pre{\hypertarget{ref-eichenbaum_health_2022}{}}%
Eichenbaum, A., \& Tate, A. D. (2022). Health {Inequity} in {Georgia} {During} the {COVID}-19 {Pandemic}: {An} {Ecological} {Analysis} {Assessing} the {Relationship} {Between} {County}-{Level} {Racial}/{Ethnic} and {Economic} {Polarization} {Using} the {ICE} and {SARS}-{CoV}-2 {Cases}, {Hospitalizations}, and {Deaths} in {Georgia} as of {October} 2020. \emph{Health Equity}, \emph{6}(1), 230--239. \url{https://doi.org/10.1089/heq.2021.0118}

\leavevmode\vadjust pre{\hypertarget{ref-feldman_health_2021}{}}%
Feldman, J. L., Luhur, W. E., Herman, J. L., Poteat, T., \& Meyer, I. H. (2021). Health and health care access in the {US} transgender population health ({TransPop}) survey. \emph{Andrology}, \emph{9}(6), 1707--1718. \url{https://doi.org/10.1111/andr.13052}

\leavevmode\vadjust pre{\hypertarget{ref-galaitsi_challenges_2021}{}}%
Galaitsi, S. E., Cegan, J. C., Volk, K., Joyner, M., Trump, B. D., \& Linkov, I. (2021). The challenges of data usage for the {United} {States}' {COVID}-19 response. \emph{International Journal of Information Management}, \emph{59}, 102352. \url{https://doi.org/10.1016/j.ijinfomgt.2021.102352}

\leavevmode\vadjust pre{\hypertarget{ref-gelfand_bayesian_2015}{}}%
Gelfand, A. E., \& Banerjee, S. (2015). Bayesian wombling: Finding rapid change in spatial maps. \emph{WIREs Computational Statistics}, \emph{7}(5), 307--315. \url{https://doi.org/10.1002/wics.1360}

\leavevmode\vadjust pre{\hypertarget{ref-gelman_bayesian_2013}{}}%
Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2013). Bayesian {Data} {Analysis}. In \emph{Routledge \& CRC Press}. \url{https://www.routledge.com/Bayesian-Data-Analysis/Gelman-Carlin-Stern-Dunson-Vehtari-Rubin/p/book/9781439840955}

\leavevmode\vadjust pre{\hypertarget{ref-gelman_regression_2020}{}}%
Gelman, A., Hill, J., \& Vehtari, A. (2020). \emph{Regression and {Other} {Stories}} (1st ed.). Cambridge University Press. \url{https://doi.org/10.1017/9781139161879}

\leavevmode\vadjust pre{\hypertarget{ref-glenza_covid_2020}{}}%
Glenza, J. (2020). Covid cases increase across {US} as upper midwest sees rapid rise. \emph{The Guardian}. \url{https://www.theguardian.com/world/2020/oct/22/covid-cases-coronavirus-us-midwest}

\leavevmode\vadjust pre{\hypertarget{ref-goldstein_during_2022}{}}%
Goldstein, J. (2022). During the {Omicron} surge, {Black} {New} {Yorkers} were hospitalized at a rate more than twice that of white residents. \emph{The New York Times}. \url{https://www.nytimes.com/2022/03/03/nyregion/omicron-hospitalizations-new-york-city.html}

\leavevmode\vadjust pre{\hypertarget{ref-gonzalez_conservatism_2021}{}}%
Gonzalez, K. E., James, R., Bjorklund, E. T., \& Hill, T. D. (2021). Conservatism and infrequent mask usage: {A} study of {US} counties during the novel coronavirus ({COVID}-19) pandemic. \emph{Social Science Quarterly}, \emph{102}(5), 2368--2382. \url{https://doi.org/10.1111/ssqu.13025}

\leavevmode\vadjust pre{\hypertarget{ref-greenland_ecologic_2001}{}}%
Greenland, S. (2001). Ecologic versus individual-level sources of bias in ecologic estimates of contextual health effects. \emph{International Journal of Epidemiology}, \emph{30}(6), 1343--1350. \url{https://doi.org/10.1093/ije/30.6.1343}

\leavevmode\vadjust pre{\hypertarget{ref-grossman_political_2020}{}}%
Grossman, G., Kim, S., Rexer, J. M., \& Thirumurthy, H. (2020). Political partisanship influences behavioral responses to governors' recommendations for {COVID}-19 prevention in the {United} {States}. \emph{Proceedings of the National Academy of Sciences}, \emph{117}(39), 24144--24153. \url{https://doi.org/10.1073/pnas.2007835117}

\leavevmode\vadjust pre{\hypertarget{ref-hanage_covid-19_2020}{}}%
Hanage, W. P., Testa, C., Chen, J. T., Davis, L., Pechter, E., Seminario, P., Santillana, M., \& Krieger, N. (2020). {COVID}-19: {US} federal accountability for entry, spread, and inequities---lessons for the future. \emph{European Journal of Epidemiology}, \emph{35}(11), 995--1006. \url{https://doi.org/10.1007/s10654-020-00689-2}

\leavevmode\vadjust pre{\hypertarget{ref-health_resources__services_administration_defining_2022}{}}%
Health Resources \& Services Administration. (2022). \emph{Defining {Rural} {Population}}. \url{https://www.hrsa.gov/rural-health/about-us/what-is-rural}

\leavevmode\vadjust pre{\hypertarget{ref-hernandez_disparities_2022}{}}%
Hernandez, I., Dickson, S., Tang, S., Gabriel, N., Berenbrok, L. A., \& Guo, J. (2022). Disparities in distribution of {COVID}-19 vaccines across {US} counties: {A} geographic information system--based cross-sectional study. \emph{PLOS Medicine}, \emph{19}(7), e1004069. \url{https://doi.org/10.1371/journal.pmed.1004069}

\leavevmode\vadjust pre{\hypertarget{ref-hill_covid-19_2022}{}}%
Hill, L., \& Artiga, S. (2022). {COVID}-19 {Cases} and {Deaths} by {Race}/{Ethnicity}: {Current} {Data} and {Changes} {Over} {Time}. In \emph{KFF}. \url{https://www.kff.org/coronavirus-covid-19/issue-brief/covid-19-cases-and-deaths-by-race-ethnicity-current-data-and-changes-over-time/}

\leavevmode\vadjust pre{\hypertarget{ref-johnson_whites_2022}{}}%
Johnson, A., \& Keating, D. (2022). Whites now more likely to die from covid than {Blacks}: {Why} the pandemic shifted. In \emph{Washington Post}. \url{https://www.washingtonpost.com/health/2022/10/19/covid-deaths-us-race/}

\leavevmode\vadjust pre{\hypertarget{ref-kaashoek_evolving_2021}{}}%
Kaashoek, J., Testa, C., Chen, J., Stolerman, L., Krieger, N., Hanage, W. P., \& Santillana, M. (2021). \emph{The {Evolving} {Roles} of {US} {Political} {Partisanship} and {Social} {Vulnerability} in the {COVID}-19 {Pandemic} from {February} 2020 - {February} 2021} {[}\{SSRN\} \{Scholarly\} \{Paper\}{]}. \url{https://doi.org/10.2139/ssrn.3933453}

\leavevmode\vadjust pre{\hypertarget{ref-kim_covid-19_2020}{}}%
Kim, E. J., Marrast, L., \& Conigliaro, J. (2020). {COVID}-19: {Magnifying} the {Effect} of {Health} {Disparities}. \emph{Journal of General Internal Medicine}, \emph{35}(8), 2441--2442. \url{https://doi.org/10.1007/s11606-020-05881-4}

\leavevmode\vadjust pre{\hypertarget{ref-krieger_relationship_2022}{}}%
Krieger, N., Testa, C., Chen, J. T., Hanage, W. P., \& McGregor, A. J. (2022). Relationship of political ideology of {US} federal and state elected officials and key {COVID} pandemic outcomes following vaccine rollout to adults: {April} 2021--{March} 2022. \emph{The Lancet Regional Health -- Americas}, \emph{16}. \url{https://doi.org/10.1016/j.lana.2022.100384}

\leavevmode\vadjust pre{\hypertarget{ref-krieger_public_2016}{}}%
Krieger, N., Waterman, P. D., Spasojevic, J., Li, W., Maduro, G., \& Van Wye, G. (2016). Public {Health} {Monitoring} of {Privilege} and {Deprivation} {With} the {Index} of {Concentration} at the {Extremes}. \emph{American Journal of Public Health}, \emph{106}(2), 256--263. \url{https://doi.org/10.2105/AJPH.2015.302955}

\leavevmode\vadjust pre{\hypertarget{ref-lambert_most_2022}{}}%
Lambert, J. (2022). Most {Americans} {Don}'t {Know} {About} the {Omicron} {Covid} {Boosters}. In \emph{Grid News}. \url{https://www.grid.news/story/science/2022/10/04/most-americans-dont-know-about-the-omicron-covid-boosters-that-spells-trouble-for-the-coming-winter/}

\leavevmode\vadjust pre{\hypertarget{ref-leonhardt_red_2021}{}}%
Leonhardt, D. (2021). Red {Covid}. \emph{The New York Times}. \url{https://www.nytimes.com/2021/09/27/briefing/covid-red-states-vaccinations.html}

\leavevmode\vadjust pre{\hypertarget{ref-mayo_clinic_history_2022}{}}%
Mayo Clinic. (2022). History of {COVID}-19: {Outbreaks} and {Vaccine} {Timeline}. In \emph{MayoClinic.org}. \url{https://www.mayoclinic.org/coronavirus-covid-19/history-disease-outbreaks-vaccine-timeline/covid-19}

\leavevmode\vadjust pre{\hypertarget{ref-mit_election_data_and_science_lab_county_2022}{}}%
MIT Election Data and Science Lab. (2022). \emph{County {Presidential} {Election} {Returns} 2000-2020}. Harvard Dataverse. \url{https://doi.org/10.7910/DVN/VOQCHQ}

\leavevmode\vadjust pre{\hypertarget{ref-mollalo_gis-based_2020}{}}%
Mollalo, A., Vahedi, B., \& Rivera, K. M. (2020). {GIS}-based spatial modeling of {COVID}-19 incidence rate in the continental {United} {States}. \emph{Science of The Total Environment}, \emph{728}, 138884. \url{https://doi.org/10.1016/j.scitotenv.2020.138884}

\leavevmode\vadjust pre{\hypertarget{ref-moreland_timing_2020}{}}%
Moreland, A., Herlihy, C., Tynan, M. A., Sunshine, G., McCord, R. F., Hilton, C., Poovey, J., Werner, A. K., Jones, C. D., Fulmer, E. B., Gundlapalli, A. V., Strosnider, H., Potvien, A., García, M. C., Honeycutt, S., Baldwin, G., CDC Public Health Law Program, CDC COVID-19 Response Team, \& Mitigation Policy Analysis Unit. (2020). Timing of {State} and {Territorial} {COVID}-19 {Stay}-at-{Home} {Orders} and {Changes} in {Population} {Movement} --- {United} {States}, {March} 1--{May} 31, 2020. \emph{MMWR. Morbidity and Mortality Weekly Report}, \emph{69}. \url{https://doi.org/10.15585/mmwr.mm6935a2}

\leavevmode\vadjust pre{\hypertarget{ref-murthy_disparities_2021}{}}%
Murthy, B. P. (2021). Disparities in {COVID}-19 {Vaccination} {Coverage} {Between} {Urban} and {Rural} {Counties} --- {United} {States}, {December} 14, 2020--{April} 10, 2021. \emph{MMWR. Morbidity and Mortality Weekly Report}, \emph{70}. \url{https://doi.org/10.15585/mmwr.mm7020e3}

\leavevmode\vadjust pre{\hypertarget{ref-national_bureau_of_economic_research_measuring_2021}{}}%
National Bureau of Economic Research. (2021). Measuring the {Virus} {Risk} of {Essential} {Workers} and {Dependents}. In \emph{NBER}. \url{https://www.nber.org/digest/202103/measuring-virus-risk-essential-workers-and-dependents}

\leavevmode\vadjust pre{\hypertarget{ref-national_conference_of_state_legislatures_covid-19_2021}{}}%
National Conference of State Legislatures. (2021). {COVID}-19: {Essential} {Workers} in the {States}. In \emph{COVID-19: Essential Workers in the States}. \url{https://www.ncsl.org/research/labor-and-employment/covid-19-essential-workers-in-the-states.aspx}

\leavevmode\vadjust pre{\hypertarget{ref-office_of_the_commissioner_coronavirus_2022}{}}%
Office of the Commissioner. (2022). Coronavirus ({COVID}-19) {Update}: {FDA} {Authorizes} {Moderna}, {Pfizer}-{BioNTech} {Bivalent} {COVID}-19 {Vaccines} for {Use} as a {Booster} {Dose}. In \emph{FDA}. \url{https://www.fda.gov/news-events/press-announcements/coronavirus-covid-19-update-fda-authorizes-moderna-pfizer-biontech-bivalent-covid-19-vaccines-use}

\leavevmode\vadjust pre{\hypertarget{ref-okonkwo_covid-19_2021}{}}%
Okonkwo, N. E., Aguwa, U. T., Jang, M., Barré, I. A., Page, K. R., Sullivan, P. S., Beyrer, C., \& Baral, S. (2021). {COVID}-19 and the {US} response: Accelerating health inequities. \emph{BMJ Evidence-Based Medicine}, \emph{26}(4), 176--179. \url{https://doi.org/10.1136/bmjebm-2020-111426}

\leavevmode\vadjust pre{\hypertarget{ref-ortega_ending_2021}{}}%
Ortega, A. N., \& Roby, D. H. (2021). Ending {Structural} {Racism} in the {US} {Health} {Care} {System} to {Eliminate} {Health} {Care} {Inequities}. \emph{JAMA}, \emph{326}(7), 613--615. \url{https://doi.org/10.1001/jama.2021.11160}

\leavevmode\vadjust pre{\hypertarget{ref-park_covid-19_2021}{}}%
Park, Y. M., Kearney, G. D., Wall, B., Jones, K., Howard, R. J., \& Hylock, R. H. (2021). {COVID}-19 {Deaths} in the {United} {States}: {Shifts} in {Hot} {Spots} over the {Three} {Phases} of the {Pandemic} and the {Spatiotemporally} {Varying} {Impact} of {Pandemic} {Vulnerability}. \emph{International Journal of Environmental Research and Public Health}, \emph{18}(17), 8987. \url{https://doi.org/10.3390/ijerph18178987}

\leavevmode\vadjust pre{\hypertarget{ref-press_inequities_2021}{}}%
Press, V. G., Huisingh-Scheetz, M., \& Arora, V. M. (2021). Inequities in {Technology} {Contribute} to {Disparities} in {COVID}-19 {Vaccine} {Distribution}. \emph{JAMA Health Forum}, \emph{2}(3), e210264. \url{https://doi.org/10.1001/jamahealthforum.2021.0264}

\leavevmode\vadjust pre{\hypertarget{ref-rader_spatial_2021}{}}%
Rader, B., Astley, C. M., Sewalk, K., Delamater, P. L., Cordiano, K., Wronski, L., Rivera, J. M., Hallberg, K., Pera, M. F., Cantor, J., Whaley, C. M., Bravata, D. M., \& Brownstein, J. S. (2021). \emph{Spatial {Accessibility} {Modeling} of {Vaccine} {Deserts} as {Barriers} to {Controlling} {SARS}-{CoV}-2}. medRxiv. \url{https://doi.org/10.1101/2021.06.09.21252858}

\leavevmode\vadjust pre{\hypertarget{ref-rapp_statelevel_2022}{}}%
Rapp, K. S., Volpe, V. V., Hale, T. L., \& Quartararo, D. F. (2022). State--{Level} {Sexism} and {Gender} {Disparities} in {Health} {Care} {Access} and {Quality} in the {United} {States}. \emph{Journal of Health and Social Behavior}, \emph{63}(1), 2--18. \url{https://doi.org/10.1177/00221465211058153}

\leavevmode\vadjust pre{\hypertarget{ref-saha_neighborhood-level_2020}{}}%
Saha, S., \& Feldman, J. M. (2020). \emph{Neighborhood-level {Racial}/{Ethnic} and {Economic} {Inequities} in {COVID}-19 {Burden} {Within} {Urban} {Areas} in the {US} and {Canada}}. medRxiv. \url{https://doi.org/10.1101/2020.12.07.20241018}

\leavevmode\vadjust pre{\hypertarget{ref-scott_these_2020}{}}%
Scott, D. (2020). These 4 {Midwestern} states are seeing worrying {Covid}-19 spikes. In \emph{Vox}. \url{https://www.vox.com/2020/9/2/21418812/covid-19-coronavirus-us-cases-midwest-surge}

\leavevmode\vadjust pre{\hypertarget{ref-shumaker_covid-19_2020}{}}%
Shumaker, L., \& Wu, T. (2020). {COVID}-19 cases surge in {U}.{S}. {Midwest} and {Northeast}. \emph{Reuters}. \url{https://www.reuters.com/article/us-health-coronavirus-usa-trends-idUSKBN26Q30B}

\leavevmode\vadjust pre{\hypertarget{ref-sigler_socio-spatial_2021}{}}%
Sigler, T., Mahmuda, S., Kimpton, A., Loginova, J., Wohland, P., Charles-Edwards, E., \& Corcoran, J. (2021). The socio-spatial determinants of {COVID}-19 diffusion: The impact of globalisation, settlement characteristics and population. \emph{Globalization and Health}, \emph{17}(1), 56. \url{https://doi.org/10.1186/s12992-021-00707-2}

\leavevmode\vadjust pre{\hypertarget{ref-sonderlund_racialized_2022}{}}%
Sonderlund, A. L., Charifson, M., Schoenthaler, A., Carson, T., \& Williams, N. J. (2022). Racialized economic segregation and health outcomes: {A} systematic review of studies that use the {Index} of {Concentration} at the {Extremes} for race, income, and their interaction. \emph{PLOS ONE}, \emph{17}(1), e0262962. \url{https://doi.org/10.1371/journal.pone.0262962}

\leavevmode\vadjust pre{\hypertarget{ref-sugg_mapping_2021}{}}%
Sugg, M. M., Spaulding, T. J., Lane, S. J., Runkle, J. D., Harden, S. R., Hege, A., \& Iyer, L. S. (2021). Mapping community-level determinants of {COVID}-19 transmission in nursing homes: {A} multi-scale approach. \emph{Science of The Total Environment}, \emph{752}, 141946. \url{https://doi.org/10.1016/j.scitotenv.2020.141946}

\leavevmode\vadjust pre{\hypertarget{ref-testa_public_2022}{}}%
Testa, C., Chen, J., Hall, E., Javadi, D., Morgan, J., Rushovich, T., Saha, S., Waterman, P., \& Krieger, N. (2022). \emph{Public {Health} {Disparities} {Geocoding} {Project} 2.0 {Training} {Manual}}. \url{https://phdgp.github.io/PHDGP2.0/index.html}

\leavevmode\vadjust pre{\hypertarget{ref-the_lancet_plight_2020}{}}%
The Lancet. (2020). The plight of essential workers during the {COVID}-19 pandemic. \emph{The Lancet}, \emph{395}(10237), 1587. \url{https://doi.org/10.1016/S0140-6736(20)31200-9}

\leavevmode\vadjust pre{\hypertarget{ref-the_new_york_times_coronavirus_2021}{}}%
The New York Times. (2021). \emph{Coronavirus ({Covid}-19) {Data} in the {United} {States}}. The New York Times. \url{https://github.com/nytimes/covid-19-data}

\leavevmode\vadjust pre{\hypertarget{ref-the_white_house_executive_2021}{}}%
The White House. (2021). Executive {Order} {On} {Advancing} {Racial} {Equity} and {Support} for {Underserved} {Communities} {Through} the {Federal} {Government}. In \emph{The White House}. \url{https://www.whitehouse.gov/briefing-room/presidential-actions/2021/01/20/executive-order-advancing-racial-equity-and-support-for-underserved-communities-through-the-federal-government/}

\leavevmode\vadjust pre{\hypertarget{ref-the_white_house_advancing_2022}{}}%
The White House. (2022). Advancing {Equity} and {Racial} {Justice} {Through} the {Federal} {Government}. In \emph{The White House}. \url{https://www.whitehouse.gov/equity/}

\leavevmode\vadjust pre{\hypertarget{ref-thompson_covid-19_2020}{}}%
Thompson, C. N. (2020). {COVID}-19 {Outbreak} --- {New} {York} {City}, {February} 29--{June} 1, 2020. \emph{MMWR. Morbidity and Mortality Weekly Report}, \emph{69}. \url{https://doi.org/10.15585/mmwr.mm6946a2}

\leavevmode\vadjust pre{\hypertarget{ref-noauthor_united_2022}{}}%
United {States} {COVID} - {Coronavirus} {Statistics} - {Worldometer}. (2022). In \emph{Worldometers}. \url{https://www.worldometers.info/coronavirus/country/us/}

\leavevmode\vadjust pre{\hypertarget{ref-us_department_of_commerce_economics_and_statistics_administration_census_2000}{}}%
U.S. Department of Commerce Economics and Statistics Administration, \& U.S. Census Bureau. (2000). \emph{Census {Regions} and {Divisions} of the {United} {States}}. \url{https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-us_census_bureau_nation_2022}{}}%
US Census Bureau. (2022). Nation {Continues} to {Age} as {It} {Becomes} {More} {Diverse}. In \emph{Census.gov}. \url{https://www.census.gov/newsroom/press-releases/2022/population-estimates-characteristics.html}

\leavevmode\vadjust pre{\hypertarget{ref-us_census_bureau_2020_2021}{}}%
US Census Bureau. (2021). 2020 {Census}: {Redistricting} {File} ({Public} {Law} 94-171) {Dataset}. In \emph{Census.gov}. \url{https://www.census.gov/data/datasets/2020/dec/2020-census-redistricting-summary-file-dataset.html}

\leavevmode\vadjust pre{\hypertarget{ref-us_census_bureau_american_2020}{}}%
US Census Bureau. (2020). American {Community} {Survey} 2015-2019 5-{Year} {Data} {Release}. In \emph{Census.gov}. \url{https://www.census.gov/newsroom/press-kits/2020/acs-5-year.html}

\leavevmode\vadjust pre{\hypertarget{ref-walker_tidycensus_2022}{}}%
Walker, K., \& Herman, M. (2022). \emph{Tidycensus: {Load} {US} {Census} {Boundary} and {Attribute} {Data} as 'tidyverse' and 'sf'-{Ready} {Data} {Frames}}.

\leavevmode\vadjust pre{\hypertarget{ref-wei_risk_2022}{}}%
Wei, C.-F., Lan, F.-Y., Hsu, Y.-T., Lowery, N., Dibona, L., Akkeh, R., Kales, S. N., \& Yang, J. (2022). Risk of {SARS}-{CoV}-2 {Infection} {Among} {Essential} {Workers} in a {Community}-{Based} {Cohort} in the {United} {States}. \emph{Frontiers in Public Health}, \emph{10}. \url{https://www.frontiersin.org/articles/10.3389/fpubh.2022.878208}

\leavevmode\vadjust pre{\hypertarget{ref-whitehead_outness_2016}{}}%
Whitehead, J., Shaver, J., \& Stephenson, R. (2016). Outness, {Stigma}, and {Primary} {Health} {Care} {Utilization} among {Rural} {LGBT} {Populations}. \emph{PLOS ONE}, \emph{11}(1), e0146139. \url{https://doi.org/10.1371/journal.pone.0146139}

\leavevmode\vadjust pre{\hypertarget{ref-whittle_ecological_2020}{}}%
Whittle, R. S., \& Diaz-Artiles, A. (2020). An ecological study of socioeconomic predictors in detection of {COVID}-19 cases across neighborhoods in {New} {York} {City}. \emph{BMC Medicine}, \emph{18}(1), 271. \url{https://doi.org/10.1186/s12916-020-01731-6}

\leavevmode\vadjust pre{\hypertarget{ref-wood_mgcv_2022}{}}%
Wood, S. (2022). \emph{Mgcv: {Mixed} {GAM} {Computation} {Vehicle} with {Automatic} {Smoothness} {Estimation}}. \url{https://CRAN.R-project.org/package=mgcv}

\leavevmode\vadjust pre{\hypertarget{ref-wood_generalized_2017}{}}%
Wood, S. N. (2017). \emph{Generalized {Additive} {Models}: {An} {Introduction} with {R}} (2nd ed.). Chapman \& Hall/CRC Texts in Statistical Science. \url{https://www.routledge.com/Generalized-Additive-Models-An-Introduction-with-R-Second-Edition/Wood/p/book/9781498728331}

\leavevmode\vadjust pre{\hypertarget{ref-wood_soap_2008}{}}%
Wood, S. N., Bravington, M. V., \& Hedley, S. L. (2008). Soap film smoothing. \emph{Journal of the Royal Statistical Society: Series B (Statistical Methodology)}, \emph{70}(5), 931--955. \url{https://doi.org/10.1111/j.1467-9868.2008.00665.x}

\end{CSLReferences}

\end{document}

### 2022-12-01 manuscript/manuscript.rmd

@@
 $\mathbf A_i$ is a row of the model matrix 
-for any strictly non-parametric model components, $\mathbf \theta$ is the corresponding
+for any strictly parametric model components, $\mathbf \theta$ is the corresponding
 parameter vector, and the $f_j$ are smooth functions of the covariates $x_k$ [@wood_generalized_2017].
